1.  **Mammal life spans.** 
Data were collected on life spans (in years) and gestation lengths (in days) for 62 mammals. 
A scatterplot of life span versus length of gestation is shown below.^[The data used in this exercise can be found in the **openintro** package: [`mammals`](http://openintrostat.github.io/openintro/reference/mammals.html).]

    ```{r}
    library(ggplot2)
    library(openintro)
    
    ggplot(mammals, aes(x = gestation, y = life_span)) +
      geom_point(col = COL["blue","full"], alpha = 0.8) +
      theme_minimal() +
      labs(x = "Gestation (days)", y = "Life Span (years)")
    ```

    a.  What type of an association is apparent between life span and length of gestation?

    b.  What type of an association would you expect to see if the axes of the plot were reversed, i.e. if we plotted length of gestation versus life span?

    c.  Are life span and length of gestation independent? Explain your reasoning.

1.  **Associations.**
Indicate which of the plots show (a) a positive association, (b) a negative association, or (c) no association. Also determine if the positive and negative associations are linear or nonlinear. Each part may refer to more than one plot.

    ```{r fig.asp=0.5}
    library(tidyverse)

    set.seed(2306)
    
    # create x
    x <- seq(0, 10, 0.1)
    
    # create y_poslin: positive linear with x
    y_poslin <- x * runif(1, min = 0, max = 10) + 
      rnorm(length(x), mean = 0, sd = runif(1, min = 3, max = 10)) - 
      runif(1, min = 0, max = 3)
    
    # create y_neglin: negative linear with x
    y_neglin <- x * -runif(1, min = 0, max = 5) + 
      rnorm(length(x), mean = 0, sd = runif(1, min = 3, max = 10)) - 
      runif(1, min = 0, max = 5)
    
    # create y_poscur: curved positive with x
    y_poscur <- x^2 + rnorm(length(x), mean = -3, sd = runif(1, min = 3, max = 4))
    
    # create y_none: no association with x
    y_none <- x + rnorm(length(x), mean = 0, sd = runif(1, min = 30, max = 40))
    
    # data frame
    tibble(
      x = x,
      poslin = y_poslin,
      neglin = y_neglin,
      poscur = y_poscur,
      none   = y_none
    ) %>%
      pivot_longer(cols = -x, names_to = "type", values_to = "y") %>%
      arrange(type) %>%
      mutate(
        type_num = case_when(
          type == "poslin" ~ "(1)",
          type == "neglin" ~ "(4)",
          type == "poscur" ~ "(3)",
          type == "none"   ~ "(2)"
        ),
        type_num = as.factor(type_num)
        ) %>%
      ggplot(aes(x = x, y = y)) +
      geom_point(color = COL["blue","full"], alpha = 0.5) +
      facet_grid(~type_num) +
      theme_minimal() +
      theme(
        axis.text = element_blank()
      ) +
      labs(x = NULL, y = NULL)
    ````

1.  **Reproducing bacteria.** 
Suppose that there is only sufficient space and nutrients to support one million bacterial cells in a petri dish. 
You place a few bacterial cells in this petri dish, allow them to reproduce freely, and record the number of bacterial cells in the dish over time. 
Sketch a plot representing the relationship between number of bacterial cells and time.

1.  **Office productivity.** 
Office productivity is relatively low when the employees feel no stress about their work or job security. 
However, high levels of stress can also lead to reduced employee productivity. 
Sketch a plot to represent the relationship between stress and productivity.

1.  **Parameters and statistics.** 
Identify which value represents the sample mean and which value represents the claimed population mean.

    a.  American households spent an average of about \$52 in 2007 on Halloween merchandise such as costumes, decorations and candy. To see if this number had changed, researchers conducted a new survey in 2008 before industry numbers were reported. The survey included 1,500 households and found that average Halloween spending was \$58 per household.

    b.  The average GPA of students in 2001 at a private university was 3.37. A survey on a sample of 203 students from this university yielded an average GPA of 3.59 a decade later.

1.  **Sleeping in college.** 
A recent article in a college newspaper stated that college students get an average of 5.5 hrs of sleep each night. 
A student who was skeptical about this value decided to conduct a survey by randomly sampling 25 students. 
On average, the sampled students slept 6.25 hours per night. 
Identify which value represents the sample mean and which value represents the claimed population mean.

1.  **Days off at a mining plant.** 
Workers at a particular mining site receive an average of 35 days paid vacation, which is lower than the national average. 
The manager of this plant is under pressure from a local union to increase the amount of paid time off. 
However, he does not want to give more days off to the workers because that would be costly. 
Instead he decides he should fire 10 employees in such a way as to raise the average number of days off that are reported by his employees. 
In order to achieve this goal, should he fire employees who have the most number of days off, least number of days off, or those who have about the average number of days off?

1.  **Medians and IQRs.**
For each part, compare distributions A and B based on their medians and IQRs. You do not need to calculate these statistics; simply state how the medians and IQRs compare. 
Make sure to explain your reasoning.

    a.  A: 3, 5, 6, 7, 9; B: 3, 5, 6, 7, 20
    
    b.  A: 3, 5, 6, 7, 9; B: 3, 5, 7, 8, 9

    c.  A: 1, 2, 3, 4, 5; B: 6, 7, 8, 9, 10

    d.  A: 0, 10, 50, 60, 100; B: 0, 100, 500, 600, 1000

1.  **Means and SDs.** 
For each part, compare distributions A and B based on their means and standard deviations. You do not need to calculate these statistics; simply state how the means and the standard deviations compare. Make sure to explain your reasoning.
*Hint:* It may be useful to sketch dot plots of the distributions.

    a. A: 3, 5, 5, 5, 8, 11, 11, 11, 13; B: 3, 5, 5, 5, 8, 11, 11, 11, 20
    
    b. A: -20, 0, 0, 0, 15, 25, 30, 30; B: -40, 0, 0, 0, 15, 25, 30, 30
    
    c. A: 0, 2, 4, 6, 8, 10; B: 20, 22, 24, 26, 28, 30
    
    d. A: 100, 200, 300, 400, 500; B: 0, 50, 300, 550, 600
    

1.  **Histograms and box plots.** 
Describe the distribution in the histograms below and match them to the box plots.

    ```{r out.width="100%", fig.asp=0.5}
    library(ggplot2)
    library(openintro)
    library(patchwork)
    
    set.seed(7365)
    
    df <- tibble(
      sym = rnorm(1000, mean = 60, sd = 3),
      uni = runif(1000, min = 0, max = 100),
      rs  = rgamma(1000, shape = 3, rate = 2)
    ) %>%
      pivot_longer(cols = everything(), names_to = "type", values_to = "x") %>%
      arrange(type) %>%
      mutate(type_label = case_when(
        type == "sym" ~ "(a)",
        type == "uni" ~ "(b)",
        type == "rs"  ~ "(c)",
      ))
    
    p_hist <- ggplot(df, aes(x = x)) +
      geom_histogram(bins = 10, fill = COL["blue", "full"], color = "#4287aa") +
      facet_wrap(~type_label, scales = "free") +
      theme_minimal() +
      theme(axis.text.y = element_blank()) +
      labs(x = "", y = "")
    
    p_box <- df %>%
      mutate(type_label = case_when(
        type == "sym" ~ "(2)",
        type == "uni" ~ "(3)",
        type == "rs"  ~ "(1)",
      )) %>%
      ggplot(aes(y = x)) +
      geom_boxplot(outlier.color = COL["blue", "full"], outlier.alpha = 0.8) +
      facet_wrap(~type_label, scales = "free") +
      theme_minimal() +
      theme(
        axis.text.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()
        ) +
      labs(x = "", y = "")
    
    p_hist + p_box
    ```

1.  **Air quality.**
Daily air quality is measured by the air quality index (AQI) reported by the Environmental Protection Agency. 
This index reports the pollution level and what associated health effects might be a concern. 
The index is calculated for five major air pollutants regulated by the Clean Air Act and takes values from 0 to 300, where a higher value indicates lower air quality. 
AQI was reported for a sample of 91 days in 2011 in Durham, NC. 
The relative frequency histogram below shows the distribution of the AQI values on these days.^[The data used in this exercise can be found in the **openintro** package: [`pm25_2011_durham`](http://openintrostat.github.io/openintro/reference/pm25_2011_durham.html).]
    
    ```{r}
    library(ggplot2)
    library(scales)
    library(openintro)
    
    ggplot(pm25_2011_durham, aes(x = daily_aqi_value)) +
      geom_histogram(aes(y = stat(count) / sum(count)), 
                     binwidth = 5, 
                     fill = COL["blue", "full"], color = "#4287aa") +
      theme_minimal() +
      labs(x = "Daily AQI", y = "Relative frequency")
    ```
    
    a.  Estimate the median AQI value of this sample.

    b.  Would you expect the mean AQI value of this sample to be higher or lower than the median? Explain your reasoning.

    c.  Estimate Q1, Q3, and IQR for the distribution.

    d.  Would any of the days in this sample be considered to have an unusually low or high AQI? Explain your reasoning.

1.  **Median vs. mean.** 
Estimate the median for the 400 observations shown in the histogram, and note whether you expect the mean to be higher or lower than the median.

    ```{r}
    library(tidyverse)
    library(openintro)
    
    set.seed(9823)
    df <- tibble(x = (100 * rbeta(400, 12, 3)))
    
    ggplot(df, aes(x = x)) +
      geom_histogram(binwidth = 5, fill = COL["blue","full"], color = "#4287aa") +
      theme_minimal() +
      labs(x = "", y = "") +
      scale_y_continuous(breaks = seq(0, 80, 20))
    ```

1.  **Histograms vs. box plots.** 
Compare the two plots below. 
What characteristics of the distribution are apparent in the histogram and not in the box plot? 
What characteristics are apparent in the box plot but not in the histogram?
    
    ```{r fig.asp=0.5}
    library(tidyverse)
    library(openintro)
    library(patchwork)
    
    set.seed(12345)
    
    df <- tibble(bimod = c(rnorm(300, mean = 5, sd = 1), 
                           rnorm(300, mean = 12, sd = 1), 
                           runif(25, min = 13, max = 28)))
    
    p_hist <- ggplot(df, aes(x = bimod)) +
      geom_histogram(binwidth = 2.5, fill = COL["blue", "full"], color = "#4287aa") +
      labs(x = NULL, y = NULL) +
      scale_x_continuous(breaks = seq(0, 30, 5), limits = c(0, 30)) +
      theme_minimal()
    
    p_box <- ggplot(df, aes(y = bimod)) +
      geom_boxplot(outlier.color = COL["blue", "full"], outlier.alpha = 0.5) +
      theme_minimal() +
      theme(
        axis.text.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()
        ) +
      scale_y_continuous(breaks = seq(0, 30, 5), limits = c(0, 30)) +
      labs(x = NULL, y = NULL)
    
    p_hist + p_box
    ```

1.  **Facebook friends.** 
Facebook data indicate that 50% of Facebook users have 100 or more friends, and that the average friend count of users is 190. 
What do these findings suggest about the shape of the distribution of number of friends of Facebook users? [@Backstrom:2011]

1.  **Distributions and appropriate statistics.** 
For each of the following, state whether you expect the distribution to be symmetric, right skewed, or left skewed. 
Also specify whether the mean or median would best represent a typical observation in the data, and whether the variability of observations would be best represented using the standard deviation or IQR. 
Explain your reasoning.

    a.  Number of pets per household.

    b.  Distance to work, i.e. number of miles between work and home.

    c.  Heights of adult males.

1.  **Distributions and appropriate statistics.**
For each of the following, state whether you expect the distribution to be symmetric, right skewed, or left skewed. 
Also specify whether the mean or median would best represent a typical observation in the data, and whether the variability of observations would be best represented using the standard deviation or IQR. 
Explain your reasoning.

    a.  Housing prices in a country where 25% of the houses cost below \$350,000, 50% of the houses cost below \$450,000, 75% of the houses cost below \$1,000,000 and there are a meaningful number of houses that cost more than \$6,000,000.

    b.  Housing prices in a country where 25% of the houses cost below \$300,000, 50% of the houses cost below \$600,000, 75% of the houses cost below \$900,000 and very few houses that cost more than \$1,200,000.

    c.  Number of alcoholic drinks consumed by college students in a given week. Assume that most of these students don't drink since they are under 21 years old, and only a few drink excessively.

    d.  Annual salaries of the employees at a Fortune 500 company where only a few high level executives earn much higher salaries than all the other employees.

1.  **Income at the coffee shop.** 
The first histogram below shows the distribution of the yearly incomes of 40 patrons at a college coffee shop. 
Suppose two new people walk into the coffee shop: one making \$225,000 and the other \$250,000. 
The second histogram shows the new income distribution. 
Summary statistics are also provided, rounded to the nearest whole number.

    ```{r}
    library(openintro)
    library(tidyverse)
    library(patchwork)
    library(scales)
    library(knitr)
    library(kableExtra)
    
    set.seed(956)
    
    sal_symmetric <- c(rnorm(40, mean = 65000, sd = 2000))
    sal_skewed    <- c(sal_symmetric, 225000, 250000)
      
    df <- tibble(sal_symmetric = c(sal_symmetric, NA, NA), sal_skewed = sal_skewed) %>%
      pivot_longer(cols = everything(), names_to = "type", values_to = "x") %>%
      mutate(type_label = if_else(type == "sal_symmetric", "Before", "After"))
    
    # plots
    
    p_sym <- df %>% 
      filter(type == "sal_symmetric") %>%
      ggplot(aes(x = x)) +
      geom_histogram(binwidth = 1000, fill = COL["blue","full"], color = "#4287aa") +
      theme_minimal() +
      scale_x_continuous(limits = c(60000, 70000), breaks = seq(60000, 70000, 2000),     labels = label_dollar(scale = 0.001, suffix = "K", accuracy = 1)) +
      scale_y_continuous(limits = c(0, 10), breaks = seq(0, 10, 2)) +
      labs(x = NULL, y = NULL, title = "Before")
    
    p_skew <- df %>% 
      filter(type == "sal_skewed") %>%
      ggplot(aes(x = x)) +
      geom_histogram(binwidth = 1000, fill = COL["blue","full"], color = "#4287aa") +
      theme_minimal() +
      scale_x_continuous(limits = c(60000, 260000), breaks = seq(60000, 260000, 50000), labels = label_dollar(scale = 0.001, suffix = "K", accuracy = 1)) +
      scale_y_continuous(limits = c(0, 10), breaks = seq(0, 10, 2)) +
      labs(x = NULL, y = NULL, title = "After")
    
    p_sym / p_skew

    # table
    
    df %>%
      mutate(type_label = fct_relevel(type_label, "Before", "After")) %>%
      group_by(type_label) %>%
      summarise(
        n      = sum(!is.na(x)),
        Min    = min(x, na.rm = TRUE),
        Q1     = quantile(x, probs = 0.025, na.rm = TRUE),
        Median = median(x, na.rm = TRUE),
        Mean   = mean(x, na.rm = TRUE),
        Max    = max(x, na.rm = TRUE),
        SD     = sd(x, na.rm = TRUE)
      ) %>%
      mutate(across(Min:SD, label_dollar(accuracy = 1))) %>%
      kable(align = "rrrrrrr", col.names = c("", "n", "Min", "Q1", "Median", "Mean", "Max", "SD"), digits = 0) %>%
      kable_styling(full_width = FALSE)
    ```

    a.  Would the mean or the median best represent what we might think of as a typical income for the 42 patrons at this coffee shop? What does this say about the robustness of the two measures?

    b.  Would the standard deviation or the IQR best represent the amount of variability in the incomes of the 42 patrons at this coffee shop? What does this say about the robustness of the two measures?

1.  **Midrange.** 
The *midrange* of a distribution is defined as the average of the maximum and the minimum of that distribution. 
Is this statistic robust to outliers and extreme skew? Explain your reasoning.

1.  **Commute times.** 
The US census collects data on the time it takes Americans to commute to work, among many other variables. 
The histogram below shows the distribution of average commute times in 3,142 US counties in 2017. 
Also shown below is a spatial intensity map of the same data.^[The data used in this exercise can be found in the **usdata** package: [`county_complete`](http://openintrostat.github.io/openintro/reference/county_complete.html).]

    ```{r out.width="90%", fig.asp=0.4}
    library(tidyverse)
    library(maps)
    library(patchwork)
    library(openintro)
    library(scales)
    
    dfips <- maps::county.fips %>%
      as_tibble() %>% 
      extract(polyname, c("region", "subregion"), "^([^,]+),([^,]+)$")
    
    map_county <- map_data("county") %>%
      as_tibble() %>%
      left_join(dfips) %>%
      mutate(fips = case_when(
        subregion == "okaloosa"  & region == "florida"        ~ 12091L,
        subregion == "st martin" & region == "louisiana"      ~ 22099L,
        subregion == "currituck" & region == "north carolina" ~ 37053L,
        # Oglala Lakota Count, see     https://en.wikipedia.org/wiki/Oglala_Lakota_County,_South_Dakota
        subregion == "shannon"   & region == "south dakota"   ~ 46113L, 
        subregion == "galveston" & region == "texas"          ~ 48167L,
        subregion == "accomack"  & region == "virginia"       ~ 51001L,
        subregion == "pierce"    & region == "washington"     ~ 53053L,
        subregion == "san juan"  & region == "washington"     ~ 53055L,
        TRUE ~ fips
      ))
    
    county_for_map <- county_complete %>%
      select(fips, name, state, mean_work_travel_2017)
    
    map_county <- map_county %>%
      left_join(county_for_map, by = "fips")
    
    p_hist <- ggplot(county_complete, aes(x = mean_work_travel_2017)) +
      geom_histogram(binwidth = 2, fill = COL["blue","full"], color = "#4287aa") +
      theme_minimal() +
      labs(x = "Mean work travel (in min)", y = NULL)
    
    p_map <- ggplot(map_county, aes(x = long, y = lat, group = group)) +
      geom_polygon(aes(fill = mean_work_travel_2017)) +
      scale_fill_viridis_c(option = "D", labels = label_percent(scale = 1)) +
      labs(x = NULL, y = NULL, fill = "Mean\ntravel time\n(in min)") +
      theme_minimal() +
      theme(
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
        )
    
    p_hist + p_map + plot_layout(widths = c(1, 2))
    ````

    a.  Describe the numerical distribution and comment on whether or not a log transformation may be advisable for these data.

    b.  Describe the spatial distribution of commuting times using the map.

1.  **Hispanic population.** 
The US census collects data on race and ethnicity of Americans, among many other variables. The histogram below shows the distribution of the percentage of the population that is Hispanic in 3,142 counties in the US in 2010. Also shown is a histogram of logs of these values.^[The data used in this exercise can be found in the **usdata** package: [`county_complete`](http://openintrostat.github.io/openintro/reference/county_complete.html).]

    ```{r out.width="80%", fig.asp=0.5}
    library(tidyverse)
    library(maps)
    library(patchwork)
    library(openintro)
    library(scales)
    
    dfips <- maps::county.fips %>%
      as_tibble() %>% 
      extract(polyname, c("region", "subregion"), "^([^,]+),([^,]+)$")
    
    map_county <- map_data("county") %>%
      as_tibble() %>%
      left_join(dfips) %>%
      mutate(fips = case_when(
        subregion == "okaloosa"  & region == "florida"        ~ 12091L,
        subregion == "st martin" & region == "louisiana"      ~ 22099L,
        subregion == "currituck" & region == "north carolina" ~ 37053L,
        # Oglala Lakota Count, see     https://en.wikipedia.org/wiki/Oglala_Lakota_County,_South_Dakota
        subregion == "shannon"   & region == "south dakota"   ~ 46113L, 
        subregion == "galveston" & region == "texas"          ~ 48167L,
        subregion == "accomack"  & region == "virginia"       ~ 51001L,
        subregion == "pierce"    & region == "washington"     ~ 53053L,
        subregion == "san juan"  & region == "washington"     ~ 53055L,
        TRUE ~ fips
      ))
    
    county_for_map <- county_complete %>%
      select(fips, name, state, hispanic_2017)
    
    map_county <- map_county %>%
      left_join(county_for_map, by = "fips")
    
    p_hist <- ggplot(county_complete, aes(x = hispanic_2017)) +
      geom_histogram(binwidth = 5, fill = COL["blue","full"], color = "#4287aa") +
      theme_minimal() +
      scale_x_continuous(labels = label_percent(scale = 1)) +
      labs(x = "% Hispanic", y = NULL)
    
    p_hist_log <- ggplot(county_complete, aes(x = log(hispanic_2017))) +
      geom_histogram(binwidth = 0.2, fill = COL["blue","full"], color = "#4287aa") +
      theme_minimal() +
      labs(x = "log(% Hispanic)", y = NULL)
    
    p_map <- ggplot(map_county, aes(x = long, y = lat, group = group)) +
      geom_polygon(aes(fill = hispanic_2017)) +
      scale_fill_viridis_c(option = "E", labels = label_percent(scale = 1)) +
      labs(x = NULL, y = NULL, fill = "% Hispanic") +
      theme_minimal() +
      theme(
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
        )
    
    p_hist + p_hist_log
    
    p_map
    ````

    a.  Describe the numerical distribution and comment on why we might want to use log-transformed values in analyzing or modeling these data.

    b.  What features of the distribution of the Hispanic population in US counties are apparent in the map but not in the histogram? What features are apparent in the histogram but not the map?

    c.  Is one visualization more appropriate or helpful than the other? Explain your reasoning.
