# Multivariable and logistic models {#multi-logistic-models}

::: {.uptohere}
This chapter is currently under construction.
:::

::: {.chapterintro}
The principles of simple linear regression lay the foundation for more sophisticated regression models used in a wide range of challenging settings. 
In this chapter, we explore multiple regression, which introduces the possibility of more than one predictor in a linear model, and logistic regression, a technique for predicting categorical outcomes with two levels.
:::

## Regression with multiple predictors

### Indicator and categorical variables as predictors

### Including and assessing many variables in a model

### Adjusted R-squared as a better tool for multiple regression

### Exercises

## Model selection

### Identifying variables in the model that may not be helpful

### Model selection strategies

### Exercises

## Checking model conditions

### Diagnostic plots

### Options for improving the model fit

### Exercises

## Case study: Mario Kart

### Data set and the full model

### Model selection

### Checking model conditions

### Logistic regression

### Resume data

### Modeling the probability of an event

### Building the logistic model with many variables

### Model diagnostics

### Exploring discrimination between groups of different sizes

### Exercises

## Chapter review {#chp4-review}

### Terms

We introduced the following terms in the chapter. 
If you're not sure what some of these terms mean, we recommend you go back in the text and review their definitions.
We are purposefully presenting them in alphabetical order, instead of in order of appearance, so they will be a little more challenging to locate. 
However you should be able to easily spot them as **bolded text**.

```{r eval=FALSE}
make_terms_table(terms_chp_4)
```

### Chapter exercises

::: {.sectionexercise}

```{r intro, child="04-exercises/04-04-chapter-review.Rmd", eval=FALSE}
```

:::

### Interactive R tutorials

Navigate the concepts you've learned in this chapter in R using the following self-paced tutorials. 
All you need is your browser to get started!

::: {.alltutorials}
[Tutorial 4: Multiple and logistic regression](https://openintrostat.github.io/ims-tutorials/04-multivariable-and-logistic-models/)
:::

::: {.singletutorial}
[Tutorial 4 - Lesson 1: Parallel slopes](https://openintro.shinyapps.io/ims-04-multivariable-and-logistic-models-01/)
:::

::: {.singletutorial}
[Tutorial 4 - Lesson 2: Evaluating and extending parallel slopes model](https://openintro.shinyapps.io/ims-04-multivariable-and-logistic-models-02/)
:::

::: {.singletutorial}
[Tutorial 4 - Lesson 3: Multiple regression](https://openintro.shinyapps.io/ims-04-multivariable-and-logistic-models-03/)
:::

::: {.singletutorial}
[Tutorial 4 - Lesson 4: Logistic regression](https://openintro.shinyapps.io/ims-04-multivariable-and-logistic-models-04/)
:::

::: {.singletutorial}
[Tutorial 4 - Lesson 5: Case study - Italian restaurants in NYC](https://openintro.shinyapps.io/ims-04-multivariable-and-logistic-models-05/)
:::

You can also access the full list of tutorials supporting this book [here](https://openintrostat.github.io/ims-tutorials/).

### R labs

Further apply the concepts you've learned in this chapter in R with computational labs that walk you through a data analysis case study.

::: {.singlelab}
[Multiple linear regression - Grading the professor](http://openintrostat.github.io/oilabs-tidy/09_multiple_regression/multiple_regression.html)
:::

::: {.alllabs}
[Full list of labs supporting OpenIntro::Introduction to Modern Statistics](http://openintrostat.github.io/oilabs-tidy/)
:::
