# Multivariable and logistic models {#multi-logistic-models}

::: {.underconstruction}
This chapter is currently under construction.
:::

::: {.chapterintro}
The principles of simple linear regression lay the foundation for more sophisticated regression models used in a wide range of challenging settings. 
In this chapter, we explore multiple regression, which introduces the possibility of more than one predictor in a linear model, and logistic regression, a technique for predicting categorical outcomes with two levels.
:::

## Regression with multiple predictors {#regression-multiple-predictors}

Multiple regression extends simple two-variable regression to the case that still has one response but many predictors (denoted $x_1$, $x_2$, $x_3$, \...). The method is motivated by scenarios where many variables may be simultaneously connected to an output.

We will consider data about loans from the peer-to-peer lender, Lending Club, which is a data set we first encountered in Chapters \@ref(getting-started-with-data). 
The loan data includes terms of the loan as well as information about the borrower. 
The outcome variable we would like to better understand is the interest rate assigned to the loan. 
For instance, all other characteristics held constant, does it matter how much debt someone already has? Does it matter if their income has been verified? 
Multiple regression will help us answer these and other questions.

The data set includes results from 10,000 loans, and we'll be looking at a subset of the available variables, some of which will be new from those we saw in earlier chapters. 
The first six observations in the data set are shown in Table \@ref(tab:loansDataMatrix), and descriptions for each variable are shown in Table \@ref(tab:loansVariables). 
Notice that the past bankruptcy variable (`bankruptcy`) is an indicator variable, where it takes the value 1 if the borrower had a past bankruptcy in their record and 0 if not. 
Using an indicator variable in place of a category name allows for these variables to be
directly used in regression. 
Two of the other variables are categorical (`verified_income` and `issue_month`), each of which can take one of a few different non-numerical values; we'll discuss how these are handled in the model in Section \@ref(ind-and-cat-predictors). 

::: {.data}
The data can be found in the [openintro](http://openintrostat.github.io/openintro) package: [`loans_full_schema`](http://openintrostat.github.io/openintro/reference/loans_full_schema.html). Based on the data in this dataset we have created to new variables: `credit_util` which is calculated as the total credit utilized divided by the total credit limit and `bankruptcy` which turns the number of bankruptcies to an indicator variable (0 for no bankrupties and 1 for at least 1 bankruptcies). We will refer to this modified dataset as `loans`.
:::

```{r loansDataMatrix}
loans <- loans_full_schema %>%
  mutate(
    credit_util = total_credit_utilized / total_credit_limit,
    bankruptcy  = if_else(public_record_bankrupt == 0, 0, 1),
    verified_income = droplevels(verified_income)
    ) %>%
  rename(credit_checks = inquiries_last_12m) %>%
  select(interest_rate, verified_income, debt_to_income, credit_util, bankruptcy, term, issue_month, credit_checks) 

loans %>%
  slice_head(n = 6) %>%
  kable(caption = "First six rows from the `loans_full_schema` data set.") %>%
  kable_styling(full_width = FALSE, bootstrap_options = "striped", latex_options = "striped")
```

```{r loansVariables}
loans_var_def <- tribble(
  ~variable,         ~description,
  "interest_rate",   "Interest rate on the loan, in an annual percentage.",
  "verified_income", "Categorical variable describing whether the borrower's income source and amount have been verified, with levels `Verified`, `Source Verified`, and `Not Verified`.",
  "debt_to_income",  "Debt-to-income ratio, which is the percentage of total debt of the borrower divided by their total income.",
  "credit_util",     "Of all the credit available to the borrower, what fraction are they utilizing. For example, the credit utilization on a credit card would be the card's balance divided by the card's credit limit.",
  "bankruptcy",      "An indicator variable for whether the borrower has a past bankruptcy in their record. This variable takes a value of `1` if the answer is *yes* and `0` if the answer is *no*.",
  "term",            "The length of the loan, in months.",
  "issue_month",     "The month and year the loan was issued, which for these loans is always during the first quarter of 2018.", 
  "credit_checks",   "Number of credit checks in the last 12 months. For example, when filing an application for a credit card, it is common for the company receiving the application to run a credit check.",
)

loans_var_def %>%
  mutate(variable = cell_spec(variable, monospace = TRUE)) %>%
  kable(escape = FALSE, caption = "Variables and their descriptions for the `loans` data set.") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), 
                latex_options = "striped")
```


### Indicator and categorical predictors {#ind-and-cat-predictors}

Let's start by fitting a linear regression model for interest rate with a single predictor indicating whether or not a person has a bankruptcy in their record: 

$$\widehat{rate} = 12.33 + 0.74 \times bankruptcy$$

Results of this model are shown in Table \@ref(tab:intRateVsPastBankrModel).

```{r intRateVsPastBankrModel}
m_bankruptcy <- lm(interest_rate ~ bankruptcy, data = loans)
tidy(m_bankruptcy) %>%
  mutate(p.value = "<0.0001") %>%
  kable(caption = "Summary of a linear model for predicting interest rate based on whether the borrower has a bankruptcy in their record. Degrees of freedom for this model is 9998.") %>%
  kable_styling(full_width = FALSE)
```

::: {.example}
Interpret the coefficient for the past bankruptcy variable in the model. Is this coefficient significantly different from 0? 

---

The variable takes one of two values: 1 when the borrower has a bankruptcy in their history and 0 otherwise. A slope of 0.74 means that the model predicts a 0.74%
higher interest rate for those borrowers with a bankruptcy in their
record. 
(See Section \@ref(categprical-predictor-two-levels) for a review of the interpretation for two-level categorical predictor variables.) 
Examining the regression output in Table \@ref(tab:intRateVsPastBankrModel), we can see that the p-value for is very close to zero, indicating there is strong evidence the coefficient is different from zero when using this simple one-predictor model.
:::

Suppose we had fit a model using a 3-level categorical variable, such as `verified_income`. 
The output from software is shown in Table \@ref(tab:intRateVsVerIncomeModel). 
This regression output provides multiple rows for the variable. 
Each row represents the relative difference for each level of `verified_income`. 
However, we are missing one of the levels: `Not Verified`. 
The missing level is called the **reference level**, and it represents the default level that other levels are measured against.

```{r intRateVsVerIncomeModel}
m_verified_income <- lm(interest_rate ~ verified_income, data = loans)
tidy(m_verified_income) %>%
  mutate(p.value = "<0.0001") %>%
  kable(caption = "Summary of a linear model for predicting interest rate based on whether the borrower’s income source and amount has been verified. This predictor has three levels, which results in 2 rows in the regression output.") %>%
  kable_styling(full_width = FALSE)
```

::: {.example}
How would we write an equation for this regression model?

---

The equation for the regression model may be written as a model with two predictors: 

$$\widehat{rate} = 11.10 + 1.42 \times \text{verified_income}_{\text{Source Verified}} + 3.25 \times \text{verified_income}_{\text{Verified}}$$ 

We use the notation $\text{variable}_{\text{level}}$ to represent indicator variables for when the categorical variable takes a particular value. 
For example, $\text{verified_income}_{\text{Source Verified}}$ would take a value of 1 if
was for a loan, and it would take a value of 0 otherwise. 
Likewise, $\text{verified_income}_{\text{Verified}}$ would take a value of 1 if took a
value of `verified` and 0 if it took any other value.
:::

The notation $\text{variable}_{\text{level}}$ may feel a bit confusing. 
Let's figure out how to use the equation for each level of the `verified_income` variable.

::: {.example}
Using the model for predicting interest rate from income verification type, compute the average interest rate for borrowers whose income source and amount are both unverified. 

---

When `verified_income` takes a value of `Not Verified`, then both indicator functions in the equation for the linear model are set to 0:

$$\widehat{rate} = 11.10 + 1.42 \times 0 + 3.25 \times 0 = 11.10$$

The average interest rate for these borrowers is 11.1%. 
Because the level does not have its own coefficient and it is the reference value, the indicators for the other levels for this variable all drop out.
:::


::: {.example}
Using the model for predicting interest rate from income verification type, compute the average interest rate for borrowers whose income source and amount are both unverified. 

---

When `verified_income` takes a value of `Source Verified`, then the corresponding variable takes a value of 1 while the other ($\text{verified_income}_{\text{Verified}}$) is 0:

$$\widehat{rate} = = 11.10 + 1.42 \times 1 + 3.25 \times 0 = 12.52$$

The average interest rate for these borrowers is 12.52%.
:::


::: {.guidedpractice}
Compute the average interest rate for borrowers whose income source and amount are both verified.^[When `verified_income` takes a value of `Verified`, then the corresponding variable takes a value of 1 while the other is 0: $$11.10 + 1.42 \times 0 + 3.25 \times 1 = 14.35$$ The average interest rate for these borrowers is 14.35%.]
:::


::: {.important}
**Predictors with several categories.**

When fitting a regression model with a categorical variable that has $k$ levels where $k > 2$, software will provide a coefficient for $k - 1$ of those levels. 
For the last level that does not receive a coefficient, this is the , and the coefficients listed for the other levels are all considered relative to this reference level.
:::


::: {.guidedpractice}
Interpret the coefficients in the model.^[Each of the coefficients gives the incremental interest rate for the corresponding level relative to the `Not Verified` level, which is the reference level. For example, for a borrower whose income source and amount have been verified, the model predicts that they will have a 3.25% higher interest rate than a borrower who has not had their income source or amount verified.]
:::

The higher interest rate for borrowers who have verified their income source or amount is surprising. 
Intuitively, we'd think that a loan would look *less* risky if the borrower's income has been verified. 
However, note that the situation may be more complex, and there may be confounding variables that we didn't account for. 
For example, perhaps lender require borrowers with poor credit to verify their income. 
That is, verifying income in our data set might be a signal of some concerns about the borrower rather than a reassurance that the borrower will pay back the loan. 
For this reason, the borrower could be deemed higher risk, resulting in a higher interest rate. 
(What other confounding variables might explain this counter-intuitive relationship suggested by the model?)


::: {.guidedpractice}
How much larger of an interest rate would we expect for a borrower who has verified their income source and amount vs a borrower whose income source has only been verified?^[Relative to the `Not Verified` category, the `Verified` category has an interest rate of 3.25% higher, while the `Source Verified` category is only 1.42% higher. Thus, `Verified` borrowers will tend to get an interest rate about $3.25% - 1.42% = 1.83%$ higher than `Source Verified` borrowers.]
:::


### Many predictors in a model

### Adjusted R-squared

### Exercises

::: {.sectionexercise}

```{r intro, child="04-exercises/04-01-regression-multiple-predictors.Rmd"}
```

:::

## Model selection {#model-selection}

### Assessing helpfulness of predictors

### Model selection strategies

### Exercises

::: {.sectionexercise}

```{r intro, child="04-exercises/04-02-model-selection.Rmd"}
```

:::


## Model diagnostics {#model-diagnostics}

### Diagnostic plots

### Improving model fit

### Exercises

::: {.sectionexercise}

```{r intro, child="04-exercises/04-03-model-diagnostics.Rmd"}
```

:::


## Case study: Mario Kart {#case-study-mario-kart}

### Data and the full model

### Model selection

### Checking model conditions


## Logistic regression {#logistic-regression}

### Resume data

### Modeling the probability of an event

### Logistic model with many variables

### Model diagnostics

### Groups of different sizes

### Exercises

::: {.sectionexercise}

```{r intro, child="04-exercises/04-05-logistic-regression.Rmd"}
```

:::



## Chapter review {#chp4-review}

### Terms

We introduced the following terms in the chapter. 
If you're not sure what some of these terms mean, we recommend you go back in the text and review their definitions.
We are purposefully presenting them in alphabetical order, instead of in order of appearance, so they will be a little more challenging to locate. 
However you should be able to easily spot them as **bolded text**.

```{r eval=FALSE}
make_terms_table(terms_chp_4)
```

### Chapter exercises

::: {.sectionexercise}

```{r intro, child="04-exercises/04-06-chapter-review.Rmd"}
```

:::

### Interactive R tutorials

Navigate the concepts you've learned in this chapter in R using the following self-paced tutorials. 
All you need is your browser to get started!

::: {.alltutorials}
[Tutorial 4: Multiple and logistic regression](https://openintrostat.github.io/ims-tutorials/04-multivariable-and-logistic-models/)
:::

::: {.singletutorial}
[Tutorial 4 - Lesson 1: Parallel slopes](https://openintro.shinyapps.io/ims-04-multivariable-and-logistic-models-01/)
:::

::: {.singletutorial}
[Tutorial 4 - Lesson 2: Evaluating and extending parallel slopes model](https://openintro.shinyapps.io/ims-04-multivariable-and-logistic-models-02/)
:::

::: {.singletutorial}
[Tutorial 4 - Lesson 3: Multiple regression](https://openintro.shinyapps.io/ims-04-multivariable-and-logistic-models-03/)
:::

::: {.singletutorial}
[Tutorial 4 - Lesson 4: Logistic regression](https://openintro.shinyapps.io/ims-04-multivariable-and-logistic-models-04/)
:::

::: {.singletutorial}
[Tutorial 4 - Lesson 5: Case study - Italian restaurants in NYC](https://openintro.shinyapps.io/ims-04-multivariable-and-logistic-models-05/)
:::

You can also access the full list of tutorials supporting this book [here](https://openintrostat.github.io/ims-tutorials/).

### R labs

Further apply the concepts you've learned in this chapter in R with computational labs that walk you through a data analysis case study.

::: {.singlelab}
[Multiple linear regression - Grading the professor](http://openintrostat.github.io/oilabs-tidy/09_multiple_regression/multiple_regression.html)
:::

::: {.alllabs}
[Full list of labs supporting OpenIntro::Introduction to Modern Statistics](http://openintrostat.github.io/oilabs-tidy/)
:::
