# Inference for categorical data {#inference-cat}

```{block2, type="uptohere", echo=TRUE}
The content in this chapter is currently just placeholder. We will remove this banner once the chapter content has been updated and is ready for review.
```

```{block2, chp2-intro, type="chapterintro", echo=TRUE}
Focusing now on Statistical Inference for **categorical data**, we will revisit many of the foundational aspects of hypothesis testing from Chapter \@ref(inference-foundations).

The three data structures we detail are one binary variable, summarized using a single proportion; two binary variables, summarized using a difference of two proportions; and two categorical variables, summarized using a two-way table. 
When appropriate, each of the data structures will be analyzed using the three methods from Chapter \@ref(inference-foundations): randomization test, bootstrapping, and mathematical models.

As we build on the inferential ideas, we will visit new foundational concepts in statistical inference. 
For example, we will cover the conditions for when a normal model is appropriate; the two different error rates in hypothesis testing; and choosing the confidence level for a confidence interval.
```


```{r include=FALSE}
terms_chp_6 <- c("categorical data")
```

## Inference for a single proportion {#single-prop}

We encountered inference methods for a single proportion
in Chapter \@ref(inference-foundations),
exploring point estimates, confidence intervals,
and hypothesis tests.
In this section, we'll do a review of these topics
and also how to choose an appropriate sample size
when collecting data for single proportion contexts.

Note that there is only one variable being measured in a study which focuses on one proportion. 
For each observational unit, the single variable is measured as either a success or failure (e.g., "surgical complication" vs. "no surgical complication"). 
Because the nature of the research question at hand focuses on only a single variable, there is not a way to randomize the variable across a different (explanatory) variable. 
For this reason, we will not use randomization as an analysis tool when focusing on a single proportion.  Instead, we will apply bootstrapping techniques to test a given hypothesis, and we will also revisit the associated mathematical models.

### One proportion: bootstrap under $H_0$ {#one-prop-null-boot}

The bootstrap simulation concept when $H_0$ is true is similar to the ideas used
in the case studies presented in
Section \@ref(boot-ci) where we bootstrapped without an assumption about $H_0$.  Because we will be testing a hypothesized value of $p$ (referred to as $p_0$), the bootstrap simulation for hypothesis testing has a fantastic advantage that it can be used for any sample size (a huge benefit for small samples, a nice alternative for large samples).

We expand on the medical consultant example, see Section \@ref(sec-med-consult), where instead of finding an interval estimate for the true complication rate, we work to test a specific research claim.

Recall the set-up for the example:

People providing an organ for donation sometimes seek the help of a special "medical consultant". These consultants assist the patient in all aspects of the surgery, with the goal of reducing the possibility of complications during the medical procedure and recovery. Patients might choose a consultant based in part on the historical complication rate of the consultant's clients. One consultant tried to attract patients by noting the average complication rate for liver donor surgeries in the US is about 10%, but her clients have only had 3 complications in the 62 liver donor surgeries she has facilitated. She claims this is strong evidence that her work meaningfully contributes to reducing complications (and therefore she should be hired!).

```{block2, type="example", echo=TRUE}
Using the data, is it possible to assess the consultant's claim that her complication rate is less than 10%?

---

No. The claim is that there is a causal connection, but the data are observational. Patients who hire this medical consultant may have lower complication rates for other reasons.

While it is not possible to assess this causal claim, it is still possible to test for an association using these data. For this question we ask, could the low complication rate of $\hat{p} = 0.048$ be due to chance?
```



```{block2, type="guidedpractice", echo=TRUE}
Write out hypotheses in both plain and statistical language to test for the association between the consultant's work and the true complication rate, $p$, for the consultant's clients.^[$H_0$: There is no association between the consultant's contributions and the clients' complication rate. In statistical language, $p=0.10$. 
$H_A$: Patients who work with the consultant tend to have a complication rate lower than 10%, i.e., $p<0.10$.]
```


Because, as it turns out, the conditions of working with the normal distribution are not met (see Section \@ref(one-prop-norm)), the uncertainty associated with the sample proportion should not be modeled using the normal distribution. However, we would still like to assess the hypotheses from the previous Guided Practice in absence of the normal framework. To do so, we need to evaluate the possibility of a sample value ($\hat{p}$) as far below the null value, $p_0=0.10$ as what was observed. The deviation of the sample value from the hypothesized parameter is usually quantified with a p-value.

The p-value is computed based on the null distribution, which is the distribution of the test statistic if the null hypothesis is true. Supposing the null hypothesis is true, we can compute the p-value by identifying the chance of observing a test statistic that favors the alternative hypothesis at least as strongly as the observed test statistic. Here we will use a bootstrap simulation to measure the p-value.

#### Generating the null distribution and p-value by bootstrap simulation {-}

We want to identify the sampling distribution of the test statistic ($\hat{p}$) if the null hypothesis was true. In other words, we want to see how the sample proportion changes due to chance alone. Then we plan to use this information to decide whether there is enough evidence to reject the null hypothesis.

Under the null hypothesis, 10% of liver donors have complications during or after surgery. Suppose this rate was really no different for the consultant's clients (for *all* the consultant's clients, not just the 62 previously measured). If this was the case, we could *simulate* 62 clients to get a sample proportion for the complication rate from the null distribution.  Simulating observations using a hypothesized null parameter value is often called a **parametric bootstrap simulation**\index{parametric bootstrap}.


```{r include=FALSE}
terms_chp_6 <- c(terms_chp_6, "parametric bootstrap")
```

```{block2, type = "todo", echo = TRUE}
need a new physical object
```

Each client can be simulated using a single 10-sided die with one red side and nine black sides. Rolling the die once is one way of simulating the chance a patient has a complication *if the true complication rate is 10%* for the data. If we roll the die 62 times and compute the proportion of patients with complications in the simulation, $\hat{p}_{sim}$, then this sample proportion is exactly a sample from the null distribution.

An undergraduate student was paid $2 to complete this simulation. There were 5 simulated cases with a complication and 57 simulated cases without a complication, i.e., $\hat{p}_{sim} = 5/62 = 0.081$.

```{block2, type = "example", echo = TRUE}
Is this one simulation enough to determine whether or not we should reject the null hypothesis?

---
  
No. To assess the hypotheses, we need to see a distribution of many $\hat{p}_{sim}$, not just a *single* draw from this sampling distribution.
```

One simulation isn't enough to get a sense of the null distribution; many simulation studies are needed. Roughly 10,000 seems sufficient. However, paying someone to simulate 10,000 studies by hand is a waste of time and money. Instead, simulations are typically programmed into a computer, which is much more efficient.

Figure \@ref(fig:nullDistForPHatIfLiverTransplantConsultantIsNotHelpful) shows the results of 10,000 simulated studies. The proportions that are equal to or less than $\hat{p}=0.048$ are shaded. The shaded areas represent sample proportions under the null distribution that provide at least as much evidence as $\hat{p}$ favoring the alternative hypothesis. There were 1222 simulated sample proportions with $\hat{p}_{sim} \leq 0.048$. We use these to construct the null distribution's left-tail area and find the p-value:
\begin{align}
\text{left tail area }\label{estOfPValueBasedOnSimulatedNullForSingleProportion}
	&= \frac{\text{Number of observed simulations with }\hat{p}_{sim}\leq\text{ 0.048}}{10000}
\end{align}
Of the 10,000 simulated $\hat{p}_{sim}$, 1222 were equal to or smaller than $\hat{p}$. Since the hypothesis test is one-sided, the estimated p-value is equal to this tail area: 0.1222.


```{r nullDistForPHatIfLiverTransplantConsultantIsNotHelpful, fig.cap="The null distribution for $\\hat{p}$, created from 10,000 simulated studies. The left tail, representing the p-value for the hypothesis test, contains 12.22% of the simulations.", warning=FALSE, fig.width=10}

pHat <- rbinom(10^4, 62, 0.1) / 62
M    <- max(pHat) * 62
histPlot(pHat,
         breaks = (-1:(2 * M) + 0.75) / 2 / 62,
         xlab = expression(hat(p)[sim]*"    "),
         col = COL[7, 3],
         ylab = "")
histPlot(pHat[pHat < 0.05],
         breaks = (-1:(2 * M) + 0.75) / 2 / 62,
         col = COL[1],
         add = TRUE)
par(las = 0)
mtext("Number of simulations", 2, 3.3)
```

```{block2, type = "guidedpractice", echo = TRUE}
Because the estimated p-value is 0.1222, which is larger than the significance level 0.05, we do not reject the null hypothesis. Explain what this means in plain language in the context of the problem.^[There isn't sufficiently strong evidence to support an association between the consultant's work and fewer surgery complications.]
```

\index{data!medical consultant|)}

```{block2, type = "guidedpractice", echo = TRUE}
Does the conclusion in the previous Guided Practice imply there is no real association between the surgical consultant's work and the risk of complications? Explain.^[No. It might be that the consultant's work is associated with a reduction but that there isn't enough data to convincingly show this connection.]
```


```{block2, type="onebox", echo = TRUE}
**Null distribution of $\hat{p}$ with bootstrap simulation**
  
Regardless of the statistial method chosen, the p-value is always derived by analyzing the null distribution of the test statistic. The normal model poorly approximates the null distribution for $\hat{p}$ when the success-failure condition is not satisfied. As a substitute, we can generate the null distribution using simulated sample proportions and use this distribution to compute the tail area, i.e., the p-value.
```


In the previous Guided Practice, the p-value is *estimated*.
It is not exact because the simulated null distribution itself is not exact, only a close approximation.
An exact p-value can be generated using the binomial distribution, but that method will not be covered in this text.

<!--
#### Generating the exact null distribution and p-value  {-} {#exactNullDistributionUsingBinomialModel}

The number of successes in $n$ independent cases can be described using the binomial model, which was introduced in Section \ref{binomialModel}. Recall that the probability of observing exactly $k$ successes is given by
\begin{align} \label{binomialEquationShownForFindingNullDistributionInSmallSamplePropTest}
P(k\text{ successes}) = {n\choose k} p^{k}(1-p)^{n-k} = \frac{n!}{k!(n-k)!} p^{k}(1-p)^{n-k}
\end{align}
where $p$ is the true probability of success. The expression ${n\choose k}$ is read as \emph{$n$ choose $k$}, and the exclamation points represent factorials. For instance, $3!$ is equal to $3\times 2\times 1=6$, $4!$ is equal to $4\times 3\times 2\times 1 = 24$, and so on (see Section \ref{binomialModel}).

The tail area of the null distribution is computed by adding up the probability in Equation \eqref{binomialEquationShownForFindingNullDistributionInSmallSamplePropTest} for each $k$ that provides at least as strong of evidence favoring the alternative hypothesis as the data. If the hypothesis test is one-sided, then the p-value is represented by a single tail area. If the test is two-sided, compute the single tail area and double it to get the p-value, just as we have done in the past.

\begin{example}{Compute the exact p-value to check the consultant's claim that her clients' complication rate is below 105.}
Exactly $k=3$ complications were observed in the $n=62$ cases cited by the consultant. Since we are testing against the 10% national average, our null hypothesis is $p=0.10$. We can compute the p-value by adding up the cases where there are 3 or fewer complications:
\begin{align*}
\text{p-value}
	&= \sum_{j=0}^{3} {n\choose j} p^{j}(1-p)^{n-j} \\
	&= \sum_{j=0}^{3} {62\choose j} 0.1^{j}(1-0.1)^{62-j} \\
	&= {62\choose 0} 0.1^{0}(1-0.1)^{62-0} +
		{62\choose 1} 0.1^{1}(1-0.1)^{62-1} \\
	& \qquad + {62\choose 2} 0.1^{2}(1-0.1)^{62-2} +
		{62\choose 3} 0.1^{3}(1-0.1)^{62-3} \\
	&= 0.0015 + 0.0100 + 0.0340 + 0.0755 \\
	&= 0.1210
\end{align*}
This exact p-value is very close to the p-value based on the simulations (0.1222), and we come to the same conclusion. We do not reject the null hypothesis, and there is not statistically significant evidence to support the association.

If it were plotted, the exact null distribution would look almost identical to the simulated null distribution shown in Figure \ref{nullDistForPHatIfLiverTransplantConsultantIsNotHelpful} on page \pageref{nullDistForPHatIfLiverTransplantConsultantIsNotHelpful}.
\end{example}

-->


### One proportion: mathematical model {#one-prop-norm}

#### Identifying when the sample proportion is nearly normal {-}

In Section \@ref(normalDist), we introduced the normal distribution and showed how it can be used as a mathematical model to describe the variability of a statistic.
There are conditions under which a sample proportion $\hat{p}$ is well modeled using a normal distribution.
When the sample observations
are independent and the sample size is sufficiently
large, the normal model will describe the variability quite well; when the observations violate the conditions, the normal model can be inaccurate

```{block2, type = "onebox", echo = TRUE}
**Sampling distribution of
    $\hat{p}$**  

  The sampling distribution for $\hat{p}$ based on
  a sample of size $n$ from a population with a true
  proportion $p$ is nearly normal when:

1. The sample's observations are independent,
      e.g., are from a simple random sample.
2. We expected to see at least 10 successes and
      10 failures in the sample, i.e., $np\geq10$ and
      $n(1-p)\geq10$.
      This is called the **success-failure condition**.

  When these conditions are met, then the sampling
  distribution of $\hat{p}$ is nearly normal with mean
  $p$ and standard error of $\hat{p}$ as $SE = \sqrt{\frac{\ p(1-p)\ }{n}}$.
```

\index{success-failure condition}
\index{standard error (SE)!single proportion}
  
```{r include=FALSE}
terms_chp_6 <- c(terms_chp_6, "success-failure condition", "standard error of single proportion")
```

Typically we don't know the true proportion $p$,
so we substitute some value to check conditions
and estimate the standard error.
For confidence intervals, the sample proportion
$\hat{p}$ is used to check the success-failure condition
and compute the standard error.
For hypothesis tests, typically the null value --
that is, the proportion claimed in the null hypothesis --
is used in place of $p$.

The independence condition is a more nuanced requirement.
When it isn't met, it is important to understand how and why
it isn't met.
For example, if we took a cluster sample
(see Section \@ref(samp-methods)), the observations wouldn't be independent. 
However, suitable statistical methods are available for analyzing the data, but they are beyond the scope of even most second or third courses
in statistics.
On the other hand, we'd be stretched to find any method
that we could confidently apply to correct the inherent biases
of data from a convenience sample.

```{block2, type="example", echo=TRUE}
In the examples based on large sample theory, we modeled $\hat{p}$ using the normal distribution. Why is this not appropriate for the case study on the medical consultant?
  
---
  
The independence assumption may be reasonable if each of the surgeries is from a different surgical team. However, the success-failure condition is not satisfied. Under the null hypothesis, we would anticipate seeing $62\times 0.10=6.2$ complications, not the 10 required for the normal approximation.
```


While this book is scoped to well-constrained statistical
problems, do remember that this is just the first
book in what is a large library of statistical methods that
are suitable for a very wide range of data and contexts.


#### Confidence intervals for a proportion  {-}

\index{point estimate!single proportion}

A confidence interval provides a range of
plausible values for the parameter $p$,
and when $\hat{p}$ can be modeled using a
normal distribution, the confidence interval
for $p$ takes the form
\begin{align*}
\hat{p} \pm z^{\star} \times SE.
\end{align*}

We have seen $\hat{p}$ to be the sample proportion.  The value $z^{\star}$ determines the confidence level (previously set to be 1.96) and will be discussed in detail in the examples following.  The value of the standard error, $SE$, depends heavily on the sample size.

```{block2, type = "onebox", echo = TRUE}
**Standard Error of one proportion, $\hat{p}$**
  
When the conditions are met so that the distribution fo $\hat{p}$ is nearly normal, the **variability** of a single proportion, $\hat{p}$ is well described by:
  
$$SE(\hat{p}) = \sqrt{\frac{p(1-p)}{n}}$$
```

\index{data!Payday regulation poll|(}

<!--
\newcommand{\paydayN}{826}
\newcommand{\paydayNHalf}{413}
\newcommand{\paydayRegPerc}{70\%}
\newcommand{\paydayRegProp}{0.70}
\newcommand{\paydayRegSE}{0.016}
\newcommand{\paydayRegSEPerc}{1.6\%}
\newcommand{\paydayRegLower}{0.669}
\newcommand{\paydayRegUpper}{0.731}
\newcommand{\paydayRegLowerPerc}{66.9\%}
\newcommand{\paydayRegUpperPerc}{73.1\%}
% https://www.pewtrusts.org/-/media/assets/2017/04/payday-loan-customers-want-more-protections-methodology.pdf

did search and replace for each term above.  for example 826 for 826

-->

```{block2, type = "example", echo = TRUE}

A simple random sample of 826
    payday loan borrowers was surveyed to better
    understand their interests around regulation and costs.
    70% of the responses supported new
    regulations on payday lenders.
    Is it reasonable to model $\hat{p} = 0.70$
    using a normal distribution?
      
---
      
  The data are a random sample, so the observations are
  independent and representative of the population of
  interest.

  We also must check the success-failure condition,
  which we do using $\hat{p}$ in place
  of $p$ when computing a confidence interval:
  \begin{align*}
  \text{Support: }
      n p &
          \approx 826 \times 0.70
      = 578
  &\text{Not: }
      n (1 - p) &
        \approx 826 \times (1 - 0.70)
      = 248
  \end{align*}
  Since both values are at least 10, we can use the normal
  distribution to model $\hat{p}$.
```


```{block2, type = "guidedpractice", echo = TRUE}

Estimate the standard error of $\hat{p} = 0.70$.
Because $p$ is unknown and the standard error is for
a confidence interval, use $\hat{p}$ in place of $p$
in the formula.^[$SE = \sqrt{\frac{p(1-p)}{n}} \approx
    \sqrt{\frac{0.70 (1 - 0.70)}
        {826}} = 0.016$.]
```

```{block2 type = "example", echo = TRUE}

Construct a 95% confidence interval for $p$,
    the proportion of payday borrowers who support increased
    regulation for payday lenders.
    
---
      
  Using
  the point estimate 0.70,
  $z^{\star} = 1.96$ for a 95% confidence interval,
  and
  the standard error $SE = 0.016$ from the pervious
  Guided Practice,
  the confidence interval is
  \begin{eqnarray*}
  \text{point estimate} \ \pm\ z^{\star} \times SE
      \quad\to\quad
      0.70 \ \pm\ 1.96 \times 0.016
      \quad\to\quad
      (0.669, 0.731)
  \end{eqnarray*}
  We are 95% confident that the true proportion of
  payday borrowers who supported regulation at the time
  of the poll was between 0.669 and
  0.731.
```

```{block2, type = "onebox", echo = TRUE}
**Constructing a confidence interval for a single proportion**

There are three steps to constructing a confidence
  interval for $p$.

1. Check independence and the success-failure condition
      using $\hat{p}$.
      If the conditions are met, the sampling distribution
      of $\hat{p}$ may be well-approximated by the normal model.
2. Construct the standard error using $\hat{p}$
      in place of $p$ in the standard error formula.
3. Apply the general confidence interval formula.
```


For additional one-proportion confidence interval examples,
see Section \@ref(ConfidenceIntervals).

#### Changing the confidence level {-}

\index{confidence interval!confidence level|(}

Suppose we want to consider confidence intervals where the confidence level is somewhat higher than 95%: perhaps we would like a confidence level of 99%. Think back to the analogy about trying to catch a fish: if we want to be more sure that we will catch the fish, we should use a wider net. To create a 99% confidence level, we must also widen our 95% interval. On the other hand, if we want an interval with lower confidence, such as 90%, we could make our original 95% interval slightly slimmer.

The 95% confidence interval structure provides guidance in how to make intervals with new confidence levels. Below is a general 95% confidence interval for a point estimate that comes from a nearly normal distribution:
\begin{eqnarray}
\text{point estimate}\ \pm\ 1.96\times SE
\end{eqnarray}
There are three components to this interval: the point estimate, "1.96", and the standard error. The choice of $1.96\times SE$ was based on capturing 95% of the data since the estimate is within 1.96 standard errors of the true value about 95% of the time. The choice of 1.96 corresponds to a 95% confidence level. 

```{block2, type="guidedpractice", echo=TRUE}
If $X$ is a normally distributed random variable, how often will $X$ be within 2.58 standard deviations of the mean?^[This is equivalent to asking how often the $Z$ score will be larger than -2.58 but less than 2.58. (For a picture, see Figure \@ref(fig:choosingZForCI).) To determine this probability, look up -2.58 and 2.58 in the normal probability table (0.0049 and 0.9951). Thus, there is a $0.9951-0.0049 \approx 0.99$ probability that the unobserved random variable $X$ will be within 2.58 standard deviations of the mean.]
```


```{r choosingZForCI, fig.cap="The area between -$z^{\\star}$ and $z^{\\star}$ increases as $|z^{\\star}|$ becomes larger. If the confidence level is 99%, we choose $z^{\\star}$ such that 99% of the normal curve is between -$z^{\\star}$ and $z^{\\star}$, which corresponds to 0.5% in the lower tail and 0.5% in the upper tail: $z^{\\star}=2.58$.", warning=FALSE, fig.width=10}

par(mar=c(3.3,1,.5,1), mgp=c(2.1,0.6,0))
X <- rev(seq(-4, 4, 0.025))
Y <- dt(X, 10) # makes better visual
plot(X, Y, type='l', xlab='standard deviations from the mean', ylab='', axes=FALSE, xlim=3.3*c(-1,1), ylim=c(0, 0.59), col=COL[6])
axis(1, at=-3:3) #at=c(-4, -2, 0, 2, 4))
abline(h=0) #, col='#00000088')
yMax <- 0.41
X <- seq(-4, 4, 0.025)
Y <- dt(X, 10) # makes better visual
lines(X, Y, col=COL[6])

these <- (X < 2.58 & X > -2.58)
x <- c(-2.58, X[these], 2.58)
y <- c(0, dt(X[these], 10), 0)
polygon(x, y, col=COL[1,3], border='#00000000')
these <- (X < 1.96 & X > -1.96)
x <- c(-1.96, X[these], 1.96)
y <- c(0, dt(X[these], 10), 0)
polygon(x, y, col=COL[1], border='#00000000')

#x <- 1.96*c(-1,1,1,-1)
#y <- c(1,1,1.38,1.38)*yMax/3 
#polygon(x, y, border='#FFFFFF55', col='#FFFFFF77', lwd=7)
lines(1.96*c(-1,1), rep(yMax,2), lwd=2)
lines(rep(-1.96,2), c(0,yMax), lty=2, col=COL[6])
lines(rep( 1.96,2), c(0,yMax), lty=2, col=COL[6])
text(0, yMax, '95%, extends -1.96 to 1.96', pos=3)

yMax <- 0.53
#x <- 2.58*c(-1,1,1,-1)
#y <- c(1,1,1.19,1.19)*2*yMax/3 
#polygon(x, y, border='#FFFFFF55', col='#FFFFFF77', lwd=7)
lines(2.58*c(-1,1), rep(yMax,2), lwd=2)
lines(rep(-2.58,2), c(0,yMax), lty=2, col='#00000055')
lines(rep( 2.58,2), c(0,yMax), lty=2, col='#00000055')
text(0, yMax, '99%, extends -2.58 to 2.58', pos=3)

```

\index{confidence interval!confidence level|)}


To create a 99% confidence interval, change 1.96 in the 95% confidence interval formula to be $2.58$. The previous Guided Practice highlights that 99% of the time a normal random variable will be within 2.58 standard deviations of its mean. This approach -- using the Z scores in the normal model to compute confidence levels -- is appropriate when the point estimate is associated with a normal distribution and we can properly compute the standard error. Thus, the formula for a 99% confidence interval is:

\begin{eqnarray*}
\text{point estimate}\ \pm\ 2.58\times SE
\end{eqnarray*}

<!--
label for previous equation?
\label{99PercCIForMean}
\label{99PercCIForNormalPointEstimate}

%\Comment{I don't know where the equation number above gets referenced. Might drop the equation number.}
-->

The normal approximation is crucial to the precision of the $z^\star$ confidence intervals (in contrast to the bootstrap confidence intervals). When the normal model is not a good fit, we will use alternative distributions that better characterize the sampling distribution or we will use bootstrapping procedures.


```{block2, type="guidedpractice", echo=TRUE}
Create a 99% confidence interval for the impact of the stent on the risk of stroke using the data from Section \@ref(basic-stents-strokes). The point estimate is 0.090, and the standard error is $SE = 0.028$. It has been verified for you that the point estimate can reasonably be modeled by a normal distribution.^[Since the necessary conditions for applying the normal model have already been checked for us, we can go straight to the construction of the confidence interval: $\text{point estimate}\ \pm\ 2.58 \times SE \rightarrow (0.018, 0.162)$. We are 99% confident that implanting a stent in the brain of a patient who is at risk of stroke increases the risk of stroke within 30 days by a rate of 0.018 to 0.162 (assuming the patients are representative of the population).]
```



```{block2, type="onebox", echo=TRUE}
**Mathematical model confidence interval for any confidence level.**

If the point estimate follows the normal model with standard error $SE$, then a confidence interval for the population parameter is
\begin{eqnarray*}
\text{point estimate}\ \pm\ z^{\star} \times SE
\end{eqnarray*}
where $z^{\star}$ corresponds to the confidence level selected.
```

Figure \@ref(fig:choosingZForCI) provides a picture of how to identify $z^{\star}$ based on a confidence level. We select $z^{\star}$ so that the area between -$z^{\star}$ and $z^{\star}$ in the normal model corresponds to the confidence level. 



```{block2, type="onebox", echo=TRUE}
**Margin of error.**

In a confidence interval, $z^{\star}\times SE$ is called the **margin of error**\index{margin of error}.
```



```{r include=FALSE}
terms_chp_6 <- c(terms_chp_6, "margin of error")
```


```{block2, type="guidedpractice", echo=TRUE}
Previously, we found that implanting a stent in the brain of a patient at risk for a stroke *increased* the risk of a stroke. The study estimated a 9% increase in the number of patients who had a stroke, and the standard error of this estimate was about $SE = 2.8%$. Compute a 90% confidence interval for the effect.^[We must find $z^{\star}$ such that 90% of the distribution falls between -$z^{\star}$ and $z^{\star}$ in the standard normal model, $N(\mu=0, \sigma=1)$. We can look up -$z^{\star}$ in the normal probability table by looking for a lower tail of 5% (the other 5% is in the upper tail), thus $z^{\star}=1.65$. The 90% confidence interval can then be computed as $\text{point estimate}\ \pm\ 1.65\times SE \to (4.4\%, 13.6\%)$. (Note: the conditions for normality had earlier been confirmed for us.) That is, we are 90% confident that implanting a stent in a stroke patient's brain increased the risk of stroke within 30 days by 4.4% to 13.6%.]
```

#### Hypothesis testing for a proportion {-}

<!--
\label{htForPropSection}

\newcommand{\paydayCCPerc}{51\%}
\newcommand{\paydayCCProp}{0.51}
\newcommand{\paydayCCSE}{0.017}
\newcommand{\paydayCCSEPerc}{1.7\%}
\newcommand{\paydayCCZ}{0.59}
\newcommand{\paydayCCOneTail}{0.2776}
\newcommand{\paydayCCPvalue}{0.5552}
-->

One possible regulation for payday lenders is that they
would be required to do a credit check and evaluate debt
payments against the borrower's finances.
We would like to know: would borrowers support this form
of regulation?

<!--
\label{paydayCC_hypotheses_gp}%
-->

```{block2, type = "guidedpractice", echo = TRUE}
Set up hypotheses to evaluate whether borrowers
have a majority support or majority opposition for this
type of regulation.^[$H_0$: $p = 0.50$. $H_A$: $p \neq 0.50$.]
```

To apply the normal distribution framework in the context
of a hypothesis test for a proportion, the independence
and success-failure conditions must be satisfied.
In a hypothesis test, the success-failure condition is
checked using the null proportion:
we verify $np_0$ and $n(1-p_0)$ are at least 10,
where $p_0$ is the null value.



<!--
\label{paydayCC_conditions_gp}%
-->
```{block2, type = "guidedpractice", echo = TRUE}
Do payday loan borrowers support a regulation
that would
require lenders to pull their credit report
and evaluate their debt payments?
From a random sample of 826 borrowers,
51% said they would support such
a regulation.
Is it reasonable use a normal distribution to model $\hat{p}$
for a hypothesis test here?^[Independence holds since the poll
    is based on a random sample.
    The success-failure condition also holds,
    which is checked
    using the null value ($p_0 = 0.5$) from $H_0$:
    $np_0 = 826 \times 0.5 = 413$,
    $n(1 - p_0) = 826 \times 0.5 = 413$.]
```

    
```{block2, type = "example", echo = TRUE}
Using the hypotheses and data from the previous
    Guided Practices,
    evaluate whether the poll on lending regulations provides convincing evidence
    that a majority of payday loan borrowers support
    a new regulation that would
    require lenders to pull credit reports
    and evaluate debt payments.
    
---
      
  With hypotheses already set up and conditions checked,
  we can move onto calculations.
  The standard error in the context of a one-proportion
  hypothesis test is computed using the null value, $p_0$:
  \begin{align*}
  SE = \sqrt{\frac{p_0 (1 - p_0)}{n}}
      = \sqrt{\frac{0.5 (1 - 0.5)}{826}}
      = 0.017
  \end{align*}
  A picture of the normal model is shown below
  with the p-value represented by the shaded region.

  Based on the normal model, the test statistic can be
  computed as the Z-score of the point estimate:
  \begin{align*}
  Z = \frac{\text{point estimate} - \text{null value}}{SE}
      = \frac{0.51 - 0.50}{0.017}
      = 0.59
  \end{align*}
  The single tail area is 0.2776, and the p-value,
  represented by both tail areas together, is 0.5552.
  Because the p-value is larger than 0.05,
  we do not reject $H_0$.
  The poll does not provide convincing evidence that
  a majority of payday loan borrowers support or oppose
  regulations around credit checks and evaluation of
  debt payments.  In this situation, the conclusion is somewhat unsatisfactory because there is no conclusion.  That is, there is no resolution one way or the other about public opinion.  We cannot claim that exactly 50% of people support the regulation, but we cannot claim a majority in either direction.
```

  
```{r paydayCC-norm-pvalue, fig.cap="", warning=FALSE, fig.width=10}
normTail(0.5, 0.017, L = 0.49, U = 0.51, col = COL[1])  
```

<!--
\oneprophtsummary{}
-->

```{block2, type = "onebox", echo = TRUE}
**Mathematical model hypothesis test for a proportion.**
  
Set up hypotheses and verify the conditions using the null value, $p_0$, to ensure $\hat{p}$ is nearly normal under $H_0$. If the conditions hold, construct the standard error, again using $p_0$, and show the p-value in a drawing. Lastly, compute the p-value and evaluate the hypotheses.
```


For additional one-proportion hypothesis test examples,
see Section \@ref(HypothesisTesting).

\index{data!Payday regulation poll|)}
<!--
\CalculatorVideos{confidence intervals and hypothesis tests for a single proportion}
-->

#### When one or more conditions aren't met {-}

We've spent a lot of time discussing conditions for when
$\hat{p}$ can be reasonably modeled by a normal distribution.
What happens when the success-failure condition fails?
What about when the independence condition fails?
In either case, the general ideas of confidence intervals
and hypothesis tests remain the same, but the strategy
or technique used to generate the interval or p-value
change.

When the success-failure condition isn't met
for a hypothesis test, we can simulate the null distribution 
of $\hat{p}$ using the null value, $p_0$, as seen in Section \@ref(one-prop-null-boot).  Unfortunately, methods for dealing with observations which are not independent are outside the scope of this book.


#### Choosing a sample size when estimating a proportion {-}

\index{margin of error|(}

```{r include=FALSE}
terms_chp_6 <- c(terms_chp_6, "margin of error")
```

When collecting data, we choose a sample size suitable
for the purpose of the study. 
You might agree that the following interval estimate would not be particularly useful: a 95% confidence interval for the proportion of liver transplants with complications is between 0 and 1.
Often times "suitable for the study" means choosing a sample size large
enough that the **margin of error**\index{margin of error} --
which is the part we add and subtract from the point
estimate in a confidence interval --
is sufficiently small that the result is useful.
For example, our task might be to find a sample size
$n$ so that the sample proportion is within $\pm 0.04$
of the actual proportion in a 95% confidence interval.

<!--
% For example, the margin of error for a point estimate using 95% confidence can be written as $1.96\times SE$. We set up a general equation to represent the problem:
%\begin{align*}
%ME = z^{\star} \times SE \leq m
%\end{align*}
%where $ME$ represented the actual margin of error and $z^{\star}$ was chosen to correspond to the confidence level. The standard error formula is specified to correspond to the particular setting. For instance, in the case of means, the standard error was given as $\sigma / \sqrt{n}$. In the case of a single proportion, we use $\sqrt{p(1-p) / n\ }$ for the standard error.
-->

\index{data!Student football stadium|(}

```{block2, type = "example", echo = TRUE}
A university newspaper is conducting
    a survey to determine what fraction of students
    support a $200 per year increase in fees to pay
    for a new football stadium.
    How big of a sample is required to ensure the
    margin of error is smaller than 0.04 using a
    95% confidence level?

---
      
  The margin of error for a sample proportion is
  \begin{align*}
  z^{\star} \sqrt{\frac{p (1 - p)}{n}}
  \end{align*}
  Our goal is to find the smallest sample size $n$
  so that this margin of error is smaller than $0.04$.
  For a 95% confidence level, the value $z^{\star}$
  corresponds to 1.96:
  \begin{align*}
  1.96\times \sqrt{\frac{p(1-p)}{n}} \ < \ 0.04
  \end{align*}
  There are two unknowns in the equation: $p$ and $n$.
  If we have an estimate of $p$, perhaps from a prior
  survey, we could enter in that value and solve for $n$.
  If we have no such estimate, we must use some other
  value for $p$.
  It turns out that the margin of error is largest
  when $p$ is 0.5, so we typically use this
  \emph{worst case value} if no estimate of the
  proportion is available:
  \begin{align*}
	1.96\times \sqrt{\frac{0.5(1-0.5)}{n}} &\ < \ 0.04 \\
	1.96^2\times \frac{0.5(1-0.5)}{n} &\ < \ 0.04^2 \\
	1.96^2\times \frac{0.5(1-0.5)}{0.04^2} &\ < \ n \\
	600.25 &\ < \  n
  \end{align*}
  We would need over 600.25 participants, which means
  we need 601 participants or more, to ensure the
  sample proportion is within 0.04 of the true proportion
  with 95% confidence.
```

\index{data!Student football stadium|)}

When an estimate of the proportion is available, we use it in place of the worst case proportion value, 0.5.


\index{data!Tire failure rate|(}

```{block2, type = "todo", echo = TRUE}
not sure why the footnote didn't go into the footnote
```


```{block2, type = "guidedpractice", echo = TRUE}
A manager is about to oversee the mass
production of a new tire model in her factory,
and she would like to estimate what proportion of
these tires will be rejected through quality control.
The quality control team has monitored the last three
tire models produced by the factory,
failing 1.7% of tires in the first model,
6.2% of the second model,
and 1.3% of the third model.
The manager would like to examine enough tires
to estimate the failure rate of the new tire model
to within about 1% with a 90% confidence level.
There are three different failure rates to choose from.
Perform the sample size computation for each separately,
and identify three sample sizes to consider.^[For a 90% confidence interval, $z^{\star} = 1.65$,
  and since an estimate of the proportion 0.017 is available,
  we'll use it in the margin of error formula:
  \begin{align*}
  1.65\times \sqrt{\frac{0.017(1-0.017)}{n}} &\ < \ 0.01
    \qquad\to\qquad
      \frac{0.017(1-0.017)}{n} \ < \ 
          \left(\frac{0.01}{1.65}\right)^2
    \qquad\to\qquad
      454.96 \ < \ n
  \end{align*}
  For sample size calculations, we always round up,
  so the first tire model suggests 455 tires would
  be sufficient.

  A similar computation can be accomplished using 0.062
  and 0.013 for $p$, and you should verify that using these
  proportions results in minimum sample sizes of 1584
  and 350 tires, respectively.]
```

```{block2, type = "example", echo = TRUE}
The sample sizes vary widely in the previous
    Guided Practice.
    Which of the three would you suggest using?
    What would influence your choice?

---
      
  We could examine which of the old models is most
  like the new model, then choose the corresponding sample
  size.
  Or if two of the previous estimates are based on small
  samples while the other is based on a larger sample,
  we might consider the value corresponding to the larger
  sample.
  There are also other reasonable approaches.

  Also observe that the success-failure
  condition would need to be checked in the final sample.
  For instance, if we sampled $n = 1584$ tires and found
  a failure rate of 0.5%, the normal approximation would
  not be reasonable, and we would require more advanced
  statistical methods for creating the confidence interval.
```

\index{data!Tire failure rate|)}
\index{data!Payday regulation poll|(}

```{block2, type = "guidedpractice", echo = TRUE}
Suppose we want to continually track the support
of payday borrowers for regulation on lenders,
where we would conduct a new poll every month.
Running such frequent polls is expensive, so we decide
a wider margin of error of 5% for each individual survey
would be acceptable.
Based on a previous sample of borrowers where
70% supported some form of regulation,
how big should our monthly sample be for a margin
of error of 0.04 with 95% confidence?^[We complete the same computations as before,
   except now we use $0.70$ instead of $0.5$
   for $p$:
   \begin{align*}
   1.96\times \sqrt{\frac{p(1-p)}{n}}
       \approx 1.96\times
           \sqrt{\frac{0.70(1-0.70)}
               {n}}
       &\leq 0.05
     \qquad\to\qquad
       n \geq 322.7
  \end{align*}
  A sample size of 323 or more would be reasonable.
  (Reminder: always round up for sample size calculations!)
  Given that we plan to track this poll over time,
  we also may want to periodically repeat these calculations
  to ensure that we're being thoughtful in our sample
  size recommendations in case the baseline rate fluctuates.]
```

\index{data!Payday regulation poll|)}
\index{margin of error|)}

<!--
{\input{ch_inference_for_props/TeX/inference_for_a_single_proportion.tex}}
-->



## Inference for the difference of two proportions {#diff-two-prop}

We now extend the methods from Section \@ref(single-prop) to apply confidence intervals and hypothesis tests to differences in population proportions that come from two groups:$p_1 - p_2$.

<!--
%We consider three examples.
%In the first, we compare the utility of a blood thinner
%for heart attack patients.
%In the second application, we examine the efficacy of
%mammograms in reducing deaths from breast cancer.
%In the last example, a quadcopter company weighs whether
%to switch to a higher quality manufacturer of rotor blades.
-->

In our investigations, we'll identify a reasonable
point estimate of $p_1 - p_2$ based on the sample,
and you may have already guessed its form:
$\hat{p}_1 - \hat{p}_2$.
\index{point estimate!difference of proportions}
Then we'll look at the inferential analysis in three different ways: using a randomization test, applying bootstrapping for interval estimates, and, if
we verify that the point estimate
can be modeled using a normal distribution,
we compute the estimate's standard error, and
we apply the mathematical framework.

```{r include=FALSE}
terms_chp_6 <- c(terms_chp_6, "point estimate")
```

### Two proportions: randomization test {#two-prop-errors}


We consider a study on a new malaria vaccine
called PfSPZ.
In this study, volunteer patients were randomized
into one of two experiment groups:
14 patients received an experimental vaccine
or 6 patients received a placebo vaccine.
Nineteen weeks later, all 20 patients were exposed
to a drug-sensitive malaria virus strain;
the motivation of using a drug-sensitive strain
of virus here is for ethical considerations,
allowing any infections to be treated effectively.
The results are summarized in
Table \@ref(tab:malaria_vaccine_20_exp_summary),
where 9 of the 14 treatment patients remained free
of signs of infection while all of the 6 patients
in the control group patients showed some baseline
signs of infection.

```{r malaria_vaccine_20_exp_summary}
temptbl <- tribble(
 ~variable,    ~col1, ~col2, ~col3, ~col4, 
 "", "vaccine", "5", "9", "14",
 "`treatment`", "placebo", "6", "0","6",
 "", "Total", "11", "9", "20"
)

temptbl %>%
 kable(caption = "Summary results for the malaria vaccine experiment.",
    col.names = c("", "", "infection", "no infection", "Total")) %>%
 kable_styling() %>%
 add_header_above(c(" ", " ", "`outcome`" = 2, " "))
```

\newcommand{\malariaAA}{5}
\newcommand{\malariaAB}{9}
\newcommand{\malariaAD}{14}
\newcommand{\malariaBA}{6}
\newcommand{\malariaBB}{0}
\newcommand{\malariaBD}{6}
\newcommand{\malariaDA}{11}
\newcommand{\malariaDB}{9}
\newcommand{\malariaDD}{20}
\newcommand{\malariaVIR}{0.357}
\newcommand{\malariaVIRPerc}{35.7\%}
\newcommand{\malariaPIR}{1.000}
\newcommand{\malariaPIRPerc}{100\%}
\newcommand{\malariaIRDiff}{0.643}
\newcommand{\malariaIRDiffPerc}{64.3\%}


```{block2, type = "guidedpractice", echo = TRUE}
Is this an observational study or an experiment?
What implications does the study type have on what can
be inferred from the results?^[The
  study is an experiment, as patients were randomly
  assigned an experiment group.
  Since this is an experiment, the results can be used
  to evaluate a causal relationship between the malaria
  vaccine and whether patients showed signs
  of an infection.]
```

In this study, a smaller proportion of patients
who received the vaccine showed signs of an infection
(35.7% versus 100%).
However, the sample is very small,
and it is unclear whether the difference provides
*convincing evidence* that the vaccine is
effective.

<!--
```{block, type = "example", echo = TRUE}
Data scientists are sometimes called
    upon to evaluate the strength of evidence.
    When looking at the rates of infection for patients
    in the two groups in this study,
    what comes to mind as we try to determine whether
    the data show convincing evidence of a real difference?

---
  The observed infection rates
  (35.7% for the treatment group versus
  100% for the control group)
  suggest the vaccine may be effective.
  However, we cannot be sure if the observed difference
  represents the vaccine's efficacy or is just from
  random chance.
  Generally there is a little bit of fluctuation
  in sample data, and we wouldn't expect the sample
  proportions to be *exactly* equal,
  even if the truth was that the infection rates
  were independent of getting the vaccine.
  Additionally, with such small samples,
  perhaps it's common to observe such large differences
  when we randomly split a group due to chance alone!
```

The previous malaria example
is a reminder that the observed outcomes in the data
sample may not perfectly reflect the true relationships
between variables since there is \term{random noise}.
While the observed difference in rates of infection
is large, the sample size for the study is small,
making it unclear if this observed difference represents
efficacy of the vaccine or whether it is simply due to
chance.
We label these two competing claims, $H_0$ and $H_A$,
which are spoken as ``H-nought'' and ``H-A'':
\begin{itemize}
\setlength{\itemsep}{0mm}
\item[$H_0$:] \textbf{Independence model.}
    The variables \var{treatment} and \var{outcome}
    are independent.
    They have no relationship, and the observed difference
    between the proportion of patients who developed
    an infection in the two groups, 64.3%,
    was due to chance.
\item[$H_A$:] \textbf{Alternative model.}
    The variables are \emph{not} independent.
    The difference in infection rates of
    64.3%
    was not due to chance,
    and vaccine affected the rate of infection.
\end{itemize}

What would it mean if the independence model,
which says the vaccine had no influence on the
rate of infection, is true?
It would mean 11 patients were going to
develop an infection \emph{no matter which group
they were randomized into},
and 9 patients would not develop an infection
\emph{no matter which group they were randomized
into}.
That is, if the vaccine did not affect the rate
of infection, the difference in the infection rates
was due to chance alone in how the patients were
randomized.

Now consider the alternative model:
infection rates were influenced by whether a patient
received the vaccine or not.
If this was true, and especially if this influence
was substantial, we would expect to see some difference
in the infection rates of patients in the groups.

We choose between these two competing claims
by assessing if the data conflict so much with
$H_0$ that the independence model cannot be deemed
reasonable.
If this is the case, and the data support $H_A$,
then we will reject the notion of independence
and conclude there was discrimination.


\subsection{Simulating the study}
\label{simulatingTheStudy}

We're going to implement
\termsub{simulations}{simulation},
where we will pretend we know that the malaria
vaccine being tested does \emph{not} work.
Ultimately, we want to understand if the large
difference we observed is common in these
simulations.
If it is common, then maybe the difference
we observed was purely due to chance.
If it is very uncommon, then the possibility
that the vaccine was helpful seems more plausible.

Table \@ref(tab:malaria_vaccine_20_exp_summary)
shows that 11 patients developed infections and 9 did not.
For our simulation, we will suppose the infections
were independent of the vaccine and we were able to
\emph{rewind} back to when the researchers randomized
the patients in the study.
If we happened to randomize the patients differently,
we may get a different result in this hypothetical
world where the vaccine doesn't influence the infection.
Let's complete another \term{randomization} using
a simulation.

\D{\newpage}

In this \term{simulation}, we take 20 notecards to
represent the 20 patients, where we write down ``infection''
on 11 cards and ``no infection'' on 9 cards.
In this hypothetical world, we believe each patient
that got an infection was going to get it regardless
of which group they were in, so let's see what happens
if we randomly assign the patients to the treatment
and control groups again.
We thoroughly shuffle the notecards and deal 14 into
a \resp{vaccine} pile and 6 into a \resp{placebo} pile.
Finally, we tabulate the results, which are shown in
Figure \ref{malaria_vaccine_20_exp_summary_rand_1}.

\begin{figure}[ht]
\centering
\begin{tabular}{l l cc rr}
  & & \multicolumn{2}{c}{\var{outcome}} \\
  \cline{3-4}
  &  &  {infection} & {no infection} & Total & \hspace{3mm}  \\ 
  \cline{2-5}
  treatment & {vaccine} & 7 & 7 & 14 \\ 
  (simulated) & {placebo} & 4 & 2 & 6 \\ 
  \cline{2-5}
  & Total & 11 & 9 & 20 \\
  \cline{2-5}
\end{tabular}
\caption{Simulation results, where any difference
    in infection rates is purely due to chance.}
\label{malaria_vaccine_20_exp_summary_rand_1}
\end{figure}

\begin{exercisewrap}
\begin{nexercise}
\label{malaria_vaccine_20_exp_summary_rand_1_diff}
What is the difference in infection rates between
the two simulated groups in
Figure \ref{malaria_vaccine_20_exp_summary_rand_1}?
How does this compare to the observed
64.3% difference
in the actual data?\footnotemark{}
\end{nexercise}
\end{exercisewrap}
\footnotetext{$4 / 6 - 7 / 14 = 0.167$
  or about 16.7\% in favor of the vaccine.
  This difference due to chance is much smaller than the
  difference observed in the actual groups.}
-->

As we saw in Section \@ref(inf-rand), we can randomize the responses (`infection` or `no infection`) to the treatment conditions under the null hypothesis of independence and compute possible differences in proportions.
<!--
We computed one possible difference under the
independence model in Guided
Practice \ref{malaria_vaccine_20_exp_summary_rand_1_diff},
which represents one difference due to chance.
While in this first simulation, we physically dealt
out notecards to represent the patients,
it is more efficient to perform this simulation
using a computer.
Repeating the simulation on a computer, we get another
difference due to chance:
\begin{align*}
\frac{2}{\malariaBD{}} - \frac{9}{\malariaAD{}} = -0.310
\end{align*}
And another:
\begin{align*}
\frac{3}{\malariaBD{}} - \frac{8}{\malariaAD{}} = -0.071
\end{align*}
And so on until we repeat the simulation enough times
that we have a good idea of what represents the
\emph{distribution of differences from chance alone}.
-->

Figure \@ref(fig:malaria_rand_dot_plot) shows a stacked plot
of the differences found from 100 randomization simulations,
where each dot represents a simulated difference between
the infection rates (control rate minus treatment rate).

```{r malaria_rand_dot_plot, fig.cap="A stacked dot plot of differences from 100 simulations produced under the independence model $H_0$, where in these simulations infections are unaffected by the vaccine. Two of the 100 simulations had a difference of at least 64.3%, the difference observed in the study.", warning=FALSE, fig.width=10}

set.seed(47)

exp_gp  <- rep(c("vaccine", "placebo"), c(14, 6))
outcome <- c(rep(c('infection', 'no infection'), c(5, 9)),
             rep(c('infection', 'no infection'), c(6, 0)))

nsim    <- 100
n       <- length(exp_gp)
success <- "infection"

SimulateTable <- function(exp_gp, outcome, ...) {
  table(sample(exp_gp), outcome)
}
# SimulateTable(exp_gp, outcome)

sim_tables <-
    lapply(1:nsim, SimulateTable,
        exp_gp = exp_gp,
        outcome = outcome)
result <- sim_tables %>%
    lapply(function(x) {
    	  x[1, 1] / sum(x[1, ]) - x[2, 1] / sum(x[2, ])
    	}) %>%
    	unlist()


pval <- sum(result >= 0.64) / nsim
values <- table(result)
diffs <- unique(result)

X <- c()
Y <- c()
for (i in 1:length(diffs)) {
  x   <- diffs[i]
  rec <- sum(result == x)
  X   <- append(X, rep(x, rec))
  Y   <- append(Y, 1:rec)
}

plot(X, Y,
     xlim = range(diffs) + c(-1, 1) * sd(diffs) / 4,
     xlab = "Difference in Infection Rates",
     axes = FALSE,
     ylim = c(0, max(Y)),
     col = COL[1],
     pch = 20)
# at <- seq(-0.4, 0.4, 0.1)
# labels <- c(-0.4, "", -0.2, "", 0, "", 0.2, "", 0.4)
axis(1) #, at, labels)
abline(h = 0)
points(X[X > 0.64], Y[X > 0.64], lwd=3, col = COL[4], cex=0.4)
```

Note that the distribution of these simulated differences
is centered around 0.
We simulated the differences assuming that the independence
model was true, and under this condition,
we expect the difference to be near zero with some random
fluctuation, where *near* is pretty generous in this
case since the sample sizes are so small in this study.

```{block2, type = "example", echo = TRUE}
How often would you observe a difference
    of at least 64.3% (0.643)
    according to Figure \@ref(fig:malaria_rand_dot_plot)?
    Often, sometimes, rarely, or never?
---
      
  It appears that a difference of at least
  64.3% due to chance alone would only
  happen about 2% of the time according to
  Figure \@fig(fig:malaria_rand_dot_plot).
  Such a low probability indicates a rare event.
```

The difference of 64.3% being
a rare event suggests two possible interpretations
of the results of the study:

> $H_0$ Independence model. The vaccine has no effect on infection rate, and we just happened to observe a difference that would only occur on a rare occasion.
> $H_A$ Alternative model. The vaccine has an effect on infection rate, and the difference we observed was actually due to the vaccine being effective at combatting malaria, which explains the large difference of 64.3%.

Based on the simulations, we have two options.
(1) We conclude that the study results do not provide
strong evidence against the independence model.
That is, we do not have sufficiently strong evidence
to conclude the vaccine had an effect in this
clinical setting.
(2) We conclude the evidence is sufficiently strong
to reject $H_0$ and assert that the vaccine was useful.
When we conduct formal studies, usually we reject the
notion that we just happened to observe a rare
event.\footnote{This reasoning does not generally extend
    to anecdotal observations.
    Each of us observes incredibly rare events every day,
    events we could not possibly hope to predict.
    However, in the non-rigorous setting of anecdotal
    evidence, almost anything may appear to be a rare event,
    so the idea of looking for rare events in day-to-day
    activities is treacherous.
    For example, we might look at the lottery:
    there was only a 1 in 292 million chance that the
    Powerball numbers for the largest jackpot in history
    (January 13th, 2016) would be (04, 08, 19, 27, 34)
    with a Powerball of (10),
    but nonetheless those numbers came up!
    However, no matter what numbers had turned up,
    they would have had the same incredibly rare odds.
    That is, \emph{any set of numbers we could have
    observed would ultimately be incredibly rare}.
    This type of situation is typical of our daily lives:
    each possible event in itself seems incredibly rare,
    but if we consider every alternative, those outcomes
    are also incredibly rare.
    We should be cautious not to misinterpret such
    anecdotal evidence.}
So in this case, we reject the independence model in favor
of the alternative.
That is, we are concluding the data provide strong evidence
that the vaccine provides some protection against malaria
in this clinical setting.

\index{data!malaria vaccine|)}

One field of statistics, statistical inference, is built
on evaluating whether such differences are due to chance.
In statistical inference, data scientists evaluate which
model is most reasonable given the data.
Errors do occur, just like rare events, and we might choose
the wrong model.
While we do not always choose correctly, statistical
inference gives us tools to control and evaluate how
often these errors occur.
In Chapter \ref{foundationsForInference},
we give a formal introduction to the problem of model
selection.
We spend the next two chapters building a foundation
of probability and theory necessary to make that
discussion rigorous.






#### Decision errors {-}

\index{hypothesis testing!decision errors|(}

Hypothesis tests are not flawless. Just think of the court system: innocent people are sometimes wrongly convicted and the guilty sometimes walk free. 
Similarly, data can point to the wrong conclusion. 
However, what distinguishes statistical hypothesis tests from a court system is that our framework allows us to quantify and control how often the data lead us to the incorrect conclusion.

There are two competing hypotheses: the null and the alternative. 
In a hypothesis test, we make a statement about which one might be true, but we might choose incorrectly. There are four possible scenarios in a hypothesis test, which are summarized in Table \@ref(tab:fourHTScenarios).


```{r fourHTScenarios}
temptbl <- tribble(
  ~variable,    ~col1, ~col2, ~col3,
 "", "$H_0$ true", "okay", "Type 1 Error",
 "**Truth**", "$H_A$ true", "Type 2 Error", "okay"
)

temptbl %>%
 kable(caption = "Four different scenarios for hypothesis tests.", 
    col.names = c("", "", "", "")) %>%
 add_header_above( c("", "", "**Test conclusion**" = 2)) %>%
 kable_styling()
```


A **Type 1 Error**\index{Type 1 Error} is rejecting the null hypothesis when $H_0$ is actually true. 
Since we rejected the null hypothesis in the gender discrimination and opportunity cost studies, it is possible that we made a Type 1 Error in one or both of those studies. 
A **Type 2 Error**\index{Type 2 Error} is failing to reject the null hypothesis when the alternative is actually true.

```{r include=FALSE}
terms_chp_6 <- c(terms_chp_6, "Type 1 Error", "Type 2 Error")
```


```{block2, type="example", echo=TRUE}
In a US court, the defendant is either innocent ($H_0$) or guilty ($H_A$). 
What does a Type 1 Error represent in this context? 
What does a Type 2 Error represent? 
Table \@ref(tab:fourHTScenarios) may be useful.

---
 
If the court makes a Type 1 Error, this means the defendant is innocent ($H_0$ true) but wrongly convicted. 
A Type 2 Error means the court failed to reject $H_0$ (i.e., failed to convict the person) when she was in fact guilty ($H_A$ true).
```


```{block2, type="guidedpractice", echo=TRUE}
Consider the opportunity cost study where we concluded students were less likely to make a DVD purchase if they were reminded that money not spent now could be spent later. What would a Type 1 Error represent in this context?^[Making a Type 1 Error in this context would mean that reminding students that money not spent now can be spent later does not affect their buying habits, despite the strong evidence (the data suggesting otherwise) found in the experiment. Notice that this does *not* necessarily mean something was wrong with the data or that we made a computational mistake. Sometimes data simply point us to the wrong conclusion, which is why scientific studies are often repeated to check initial findings.]
```


```{block2, type="example", echo=TRUE}
How could we reduce the Type 1 Error rate in US courts? 
What influence would this have on the Type 2 Error rate?

---
 
To lower the Type 1 Error rate, we might raise our standard for conviction from "beyond a reasonable doubt" to "beyond a conceivable doubt" so fewer people would be wrongly convicted. However, this would also make it more difficult to convict the people who are actually guilty, so we would make more Type 2 Errors.
```


```{block2, type="guidedpractice", echo=TRUE}
How could we reduce the Type 2 Error rate in US courts? 
What influence would this have on the Type 1 Error rate?^[To lower the Type 2 Error rate, we want to convict more guilty people. We could lower the standards for conviction from "beyond a reasonable doubt" to "beyond a little doubt". Lowering the bar for guilt will also result in more wrongful convictions, raising the Type 1 Error rate.]
```


\index{hypothesis testing!decision errors|)}

The example and guided practice above provide an important lesson: if we reduce how often we make one type of error, we generally make more of the other type.

<!--
%Hypothesis testing is built around rejecting or failing to reject the null hypothesis. That is, we do not reject $H_0$ unless the data provide strong evidence against it. But what precisely does *strong evidence* mean? As a general rule of thumb, for those cases where the null hypothesis is actually true, we do not want to incorrectly reject $H_0$ more than 5% of the time. This corresponds to our default significance level of $\alpha = 0.05$, which we use as a comparison with the p-value. In the next section, we discuss the appropriateness of different significance levels.
-->

#### Choosing a significance level {-}

\index{hypothesis testing!significance level|(}
\index{significance level|(}

Choosing a significance level for a test is important in many contexts, and the traditional level is 0.05. 
However, it is sometimes helpful to adjust the significance level based on the application. 
We may select a level that is smaller or larger than 0.05 depending on the consequences of any conclusions reached from the test.

If making a Type 1 Error is dangerous or especially costly, we should choose a small significance level (e.g., 0.01 or 0.001). 
Under this scenario, we want to be very cautious about rejecting the null hypothesis, so we demand very strong evidence favoring the alternative $H_A$ before we would reject $H_0$.

If a Type 2 Error is relatively more dangerous or much more costly than a Type 1 Error, then we should choose a higher significance level (e.g., 0.10). 
Here we want to be cautious about failing to reject $H_0$ when the null is actually false.


```{block2, type="tipbox", echo=TRUE}
**Significance levels should reflect consequences of errors.**

The significance level selected for a test should reflect the real-world consequences associated with making a Type 1 or Type 2 Error.
```

#### Introducing two-sided hypotheses {-}

<!--
%_________________
%\section[Case study: CPR and blood thinner (randomization)]{Case study: blood thinner and CPR\\(randomization)}
-->

\index{hypothesis testing!two tails|(}

So far we have explored whether women were discriminated against and whether a simple trick could make students a little thriftier. 
In these two case studies, we've actually ignored some possibilities:

* What if *men* are actually discriminated against?
* What if the money trick actually makes students *spend more*?

These possibilities weren't considered in our hypotheses or analyses. 
This may have seemed natural since the data pointed in the directions in which we framed the problems. However, there are two dangers if we ignore possibilities that disagree with our data or that conflict with our world view:

1. Framing an alternative hypothesis simply to match the direction that the data point will generally inflate the Type 1 Error rate. After all the work we've done (and will continue to do) to rigorously control the error rates in hypothesis tests, careless construction of the alternative hypotheses can disrupt that hard work. We'll explore this topic further in Section \@ref(HypothesisTesting).

2. If we only use alternative hypotheses that agree with our worldview, then we're going to be subjecting ourselves to **confirmation bias**\index{confirmation bias}, which means we are looking for data that supports our ideas. That's not very scientific, and we can do better!


The previous hypotheses we've seen are called **one-sided hypothesis tests**\index{one-sided hypothesis test} because they only explored one direction of possibilities. 
Such hypotheses are appropriate when we are exclusively interested in the single direction, but usually we want to consider all possibilities. 
To do so, let's learn about **two-sided hypothesis tests**\index{two-sided hypothesis test} in the context of a new study that examines the impact of using blood thinners on patients who have undergone CPR.


```{r include=FALSE}
terms_chp_6 <- c(terms_chp_6, "confirmation bias", "one-sided hypothesis test", "two-sided hypothesis test")
```

\index{data!CPR and blood thinner|(}

Cardiopulmonary resuscitation (CPR) is a procedure used on individuals suffering a heart attack when other emergency resources are unavailable. 
This procedure is helpful in providing some blood circulation to keep a person alive, but CPR chest compressions can also cause internal injuries. Internal bleeding and other injuries that can result from CPR complicate additional treatment efforts. 
For instance, blood thinners may be used to help release a clot that is causing the heart attack once a patient arrives in the hospital. However, blood thinners negatively affect internal injuries.

Here we consider an experiment with patients who underwent CPR for a heart attack and were subsequently admitted to a hospital.^[B$\ddot{\text{o}}$ttiger et al. "Efficacy and safety of thrombolytic therapy after initially unsuccessful cardiopulmonary resuscitation: a prospective clinical trial." The Lancet, 2001.] 
Each patient was randomly assigned to either receive a blood thinner (treatment group) or not receive a blood thinner (control group). 
The outcome variable of interest was whether the patient survived for at least 24 hours.

<!--
% The way p_c and p_t are described make it sound like we are only considering the samples
-->


```{block2, type="example", echo=TRUE}
Form hypotheses for this study in plain and statistical language. 
Let $p_c$ represent the true survival rate of people who do not receive a blood thinner (corresponding to the control group) and $p_t$ represent the survival rate for people receiving a blood thinner (corresponding to the treatment group).

---
 
We want to understand whether blood thinners are helpful or harmful. 
We'll consider both of these possibilities using a two-sided hypothesis test.

* $H_0$: Blood thinners do not have an overall survival effect, i.e., the survival proportions are the same in each group. $p_t - p_c = 0$.

* $H_A$: Blood thinners have an impact on survival, either positive or negative, but not zero. $p_t - p_c \neq 0$.
```

There were 50 patients in the experiment who did not receive a blood thinner and 40 patients who did. 
The study results are shown in Table \@ref(tab:resultsForCPRStudyInSmallSampleSection).

```{r resultsForCPRStudyInSmallSampleSection}
temptbl <- tribble(
 ~variable,    ~col1, ~col2, ~col3, 
 "Control", "11", "39", "50",
 "Treatment", "14", "26","40",
 "Total", "25", "65", "90"
)

temptbl %>%
 kable(caption = "Results for the CPR study. Patients in the treatment group were given a blood thinner, and patients in the control group were not",
    col.names = c("", "Survived", "Died", "Total")) %>%
 kable_styling() 
```

```{block2, type="guidedpractice", echo=TRUE}
What is the observed survival rate in the control group? 
And in the treatment group? 
Also, provide a point estimate of the difference in survival proportions of the two groups: $\hat{p}_t - \hat{p}_c$.^[Observed control survival rate: $p_c = \frac{11}{50} = 0.22$. 
Treatment survival rate: $p_t = \frac{14}{40} = 0.35$. 
Observed difference: $\hat{p}_t - \hat{p}_c = 0.35 - 0.22 = 0.13$.]
```

According to the point estimate, for patients who have undergone CPR outside of the hospital, an additional 13% of these patients survive when they are treated with blood thinners. 
However, we wonder if this difference could be easily explainable by chance.

As we did in our past two studies this chapter, we will simulate what type of differences we might see from chance alone under the null hypothesis. 
By randomly assigning "simulated treatment" and "simulated control" stickers to the patients' files, we get a new grouping. 
If we repeat this simulation 10,000 times, we can build a **null distribution**\index{null distribution} of the differences shown in Figure \@ref(fig:CPR-study-right-tail).


```{r include=FALSE}
terms_chp_6 <- c(terms_chp_6, "null distribution")
```


```{r CPR-study-right-tail, fig.cap="Null distribution of the point estimate for the difference in proportions, $\\hat{p}_t - \\hat{p}_c$. The shaded right tail shows observations that are at least as large as the observed difference, 0.13.", warning=FALSE, fig.width=10}

tr <- rep(1:2, c(50, 40))
su <- c(rep(c("s", "d"), c(11, 39)), rep(c("s", "d"), c(14, 26)))

N <- 10^5
d <- rep(NA, N)
for(i in 1:N){
	trf  <- sample(tr)
	p1   <- sum(su[trf == 1] == "s") / 50
	p2   <- sum(su[trf == 2] == "s") / 40
	d[i] <- p2 - p1
}
#sum((d) > 0.1299) / N

udiffs    <- sort(unique(d))
bin.width <- diff(udiffs)[1]
breaks    <- c(udiffs - bin.width / 4, udiffs + bin.width / 4)

hist(d, breaks = breaks, col=COL[7,3], main="", xlab="Differences under the null hypothesis", ylab="", axes=FALSE)
axis(1)
axis(2, at=(0:3)*N/20, labels=c(0, NA, NA, 3)/20)
hist(d[d > 0.1299], breaks = breaks, col=COL[1], add=TRUE)
abline(h=0)
lines(rep(0.13, 2), c(0, 3)*N/25, lty=3, lwd=1.7)
text(0.13, 3*N/25, "0.13", pos=3, cex=0.7)

```

The right tail area is about 0.13. (Note: it is only a coincidence that we also have $\hat{p}_t - \hat{p}_c=0.13$.) 
However, contrary to how we calculated the p-value in previous studies, the p-value of this test is not 0.13!

The p-value is defined as the chance we observe a result at least as favorable to the alternative hypothesis as the result (i.e., the difference) we observe. 
In this case, any differences less than or equal to -0.13 would also provide equally strong evidence favoring the alternative hypothesis as a difference of 0.13. 
A difference of -0.13 would correspond to 13% higher survival rate in the control group than the treatment group. 
In Figure \@ref(fig:CPR-study-p-value) we've also shaded these differences in the left tail of the distribution. 
These two shaded tails provide a visual representation of the p-value for a two-sided test.

<!--
%There is something different in this study than in the past studies: in this study, we are particularly interested in whether blood thinners increase *or* decrease the risk of death in patients who undergo CPR before arriving at the hospital.\footnote{Realistically, we probably are interested in either direction in the past studies as well, and so we should have used the approach we now discuss in this section. However, for simplicity and the sake of not introducing too many concepts at once, we skipped over these details in earlier sections.} For example, there are chance differences of $\hat{p}_t - \hat{p}_c = -0.14$, that would have been stronger evidence against the null hypothesis as our observed difference of +0.13. Likewise, anything less than or equal -0.13 would provide as much evidence against the null hypothesis as +0.13, and for this reason, we must count both tails towards the p-value, as shown in Figure \ref{CPR-study-p-value}.
-->


```{r CPR-study-p-value, fig.cap="Null distribution of the point estimate for the difference in proportions, $\\hat{p}_t - \\hat{p}_c$.. All values that are at least as extreme as +0.13 but in either direction away from 0 are shaded.", warning=FALSE, fig.width=10}

hist(d, breaks = breaks, col=COL[7,3], main="", xlab="Differences under the null hypothesis", ylab="", axes=FALSE)
axis(1)
axis(2, at=(0:3)*N/20, labels=c(0, NA, NA, 3)/20)
hist(d[d > 0.1299], breaks = breaks, col=COL[1], add=TRUE)
abline(h=0)
hist(d[d < -0.1299], breaks = breaks, col=COL[1], add=TRUE)
abline(h=0)
lines(rep(0.13, 2), c(0, 3)*N/25, lty=3, lwd=1.7)
text(0.13, 3*N/25, "0.13", pos=3, cex=0.7)
```


For a two-sided test, take the single tail (in this case, 0.13) and double it to get the p-value: 0.26. 
Since this p-value is larger than 0.05, we do not reject the null hypothesis. 
That is, we do not find statistically significant evidence that the blood thinner has any influence on survival of patients who undergo CPR prior to arriving at the hospital. 

<!--%Once again, we can discuss the causal conclusion since this is an experiment.
-->

\index{data!CPR and blood thinner|)}

```{block2, type="onebox", echo=TRUE}
**Default to a two-sided test.**

We want to be rigorous and keep an open mind when we analyze data and evidence. 
Use a one-sided hypothesis test only if you truly have interest in only one direction.
```

```{block2, type="onebox", echo=TRUE}
**Computing a p-value for a two-sided test.**

First compute the p-value for one tail of the distribution, then double that value to get the two-sided p-value. 
That's it!
```

```{block2, type="example", echo=TRUE}
Consider the situation of the medical consultant. 
Now that you know about one-sided and two-sided tests, which type of test do you think is more appropriate?

---

The setting has been framed in the context of the consultant being helpful (which is what led us to a one-sided test originally), but what if the consultant actually performed *worse* than the average? 
Would we care? 
More than ever! 
Since it turns out that we care about a finding in either direction, we should run a two-sided test. 
The p-value for the two-sided test is double that of the one-sided test, here the simulated p-value would be 0.2444.
```


Generally, to find a two-sided p-value we double the single tail area, which remains a reasonable approach even when the sampling distribution is asymmetric.
However, the approach can result in p-values larger than 1 when the point estimate is very near the mean in the null distribution; in such cases, we write that the p-value is 1. 
Also, very large p-values computed in this way (e.g., 0.85), may also be slightly inflated. 
Typically, we do not worry too much about the precision of very large p-values because they lead to the same analysis conclusion, even if the value is slightly off.

#### Controlling the Type 1 Error rate {-}

It is never okay to change two-sided tests to one-sided tests after observing the data. 
We explore the consequences of ignoring this advice in the next example.

```{block2, type="example", echo=TRUE}
Using $\alpha=0.05$, we show that freely switching from two-sided tests to one-sided tests will lead us to make twice as many Type 1 Errors as intended.

---
 
Suppose we are interested in finding any difference from 0. 
We've created a smooth-looking **null distribution** representing differences due to chance in Figure \@ref{fig:type1ErrorDoublingExampleFigure}.

Suppose the sample difference was larger than 0. 
Then if we can flip to a one-sided test, we would use $H_A$: difference $> 0$. 
Now if we obtain any observation in the upper 5% of the distribution, we would reject $H_0$ since the p-value would just be a the single tail. 
Thus, if the null hypothesis is true, we incorrectly reject the null hypothesis about 5% of the time when the sample mean is above the null value, as shown in Figure \@ref(fig:type1ErrorDoublingExampleFigure).

Suppose the sample difference was smaller than 0. 
Then if we change to a one-sided test, we would use $H_A$: difference $< 0$. 
If the observed difference falls in the lower 5% of the figure, we would reject $H_0$. 
That is, if the null hypothesis is true, then we would observe this situation about 5% of the time.

By examining these two scenarios, we can determine that we will make a Type 1 Error $5\%+5\%=10\%$ of the time if we are allowed to swap to the "best" one-sided test for the data. 
This is twice the error rate we prescribed with our significance level: $\alpha=0.05$ (!).
```



```{r type1ErrorDoublingExampleFigure, fig.cap="The shaded regions represent areas where we would reject $H_0$ under the bad practices considered in when $\\alpha = 0.05$.", warning=FALSE, fig.width=10}
par(las=1, mar=c(2,0,0.5,0), mgp=c(3,0.9,0))
X <- seq(-4,4,0.01)
Y <- dnorm(X)
plot(X, Y, type='l', axes=F, xlim=c(-3.4,3.4))
axis(1, at=c(-5, 0, 5), label=expression(0, 0, 0), cex.axis=1.2)
these <- which(X >= 1.65)
polygon(c(X[these[1]], X[these],X[rev(these)[1]]), c(0,Y[these],0), col=COL[1])
these <- which(X <= -1.65)
polygon(c(X[these[1]], X[these],X[rev(these)[1]]), c(0,Y[these],0), col=COL[1])

arrows(-2,0.17, -1.9, 0.08, length=0.07, col=COL[1], lwd=1.5)
text(-1.92, 0.21, '5%', cex=1.2, col=COL[1])
arrows(2,0.17, 1.9, 0.08, length=0.07, col=COL[1], lwd=1.5)
text(2.08, 0.21, '5%', cex=1.2, col=COL[1])

lines(X, Y, lwd=1.5)
abline(h=0)
```

```{block2, type = "cautionbox", echo = TRUE}
**Hypothesis tests should be set up *before* seeing the data.**

After observing data, it is tempting to turn a two-sided test into a one-sided test. 
Avoid this temptation. 
Hypotheses should be set up *before* observing the data.
```

<!--
%\Comment{Should we scrap this subsection and example and just leave the caution box? Downside: weakens item 1 near the start of Section \ref{IntroducingTwoSidedHypotheses}.}
-->

\index{hypothesis testing!two tails|)}



### Two proportions: bootstrap {#two-prop-boot-ci}

The key will be to use two different dice to simulate from the original data.

### Two proportions: mathematical model 

#### Sampling distribution of the difference of two proportions {-}

Like with $\hat{p}$, the difference of two sample
proportions $\hat{p}_1 - \hat{p}_2$ can be modeled
using a normal distribution when certain conditions
are met.
First, we require a broader independence condition,
and secondly,
the success-failure condition must be met by both groups.

```{block2, type = "onebox", echo = TRUE}
**Conditions for the sampling distribution of $\hat{p}_1 -\hat{p}_2$ to be normal**
  
The difference $\hat{p}_1 - \hat{p}_2$ can be modeled
  using a normal distribution when

1. Independence, extended.  
    The data are independent within and between
    the two groups.
    Generally this is satisfied if the data come
    from two independent random samples
    or if the data come from a randomized experiment.
2. *Success-failure condition.*
    The success-failure condition holds for both
    groups, where we check successes and failures
    in each group separately.

    
  When these conditions are satisfied,
  the standard error of $\hat{p}_1 - \hat{p}_2$ is

  \begin{eqnarray*}
  SE = \sqrt{\frac{p_1(1-p_1)}{n_1} + \frac{p_2(1-p_2)}{n_2}}
  \end{eqnarray*}
  where $p_1$ and $p_2$ represent the population proportions,
  and $n_1$ and $n_2$ represent the sample sizes.
```

\index{standard error (SE)!difference in proportions}

```{block2, type = "todo", echo = TRUE}
SE reference above?
    \label{seForDiffOfProp}
```

```{r include=FALSE}
terms_chp_6 <- c(terms_chp_6, "standard error for difference in proportions")
```



#### Confidence intervals for $p_1 - p_2$ {-}

<!--
{Confidence intervals for $\pmb{p_1 - p_2}$}
-->

\index{data!CPR and blood thinner|(}

We can apply the generic confidence interval formula
for a difference of two proportions,
where we use $\hat{p}_1 - \hat{p}_2$ as the point
estimate and substitute the $SE$ formula:
\begin{align*}
&\text{point estimate} \ \pm\  z^{\star} \times SE
&&\to
&&\hat{p}_1 - \hat{p}_2 \ \pm\ 
    z^{\star} \times
   \sqrt{\frac{p_1(1-p_1)}{n_1} + \frac{p_2(1-p_2)}{n_2}}
\end{align*}
We can also follow the same
Prepare, Check, Calculate, Conclude steps for
computing a confidence interval
or completing a hypothesis test.
The details change a little,
but the general approach remain the same.
Think about these steps when you apply statistical methods.

```{block2, type = "onebox", echo = TRUE}
**Standard Error of the difference in two proportions, $\hat{p}_1 -\hat{p}_2$.**
  
When the conditions are met so that the distribution fo $\hat{p}$ is nearly normal, the **variability** of the difference in proportions, $\hat{p}_1 -\hat{p}_2$, is well described by:
\begin{eqnarray*}
  SE($\hat{p}_1 -\hat{p}_2$) = \sqrt{\frac{p_1(1-p_1)}{n_1} + \frac{p_2(1-p_2)}{n_2}}
  \end{eqnarray*}
```

<!--
```{block2, type = "onebox", echo = TRUE}
**Confidence interval for a difference of two proportions**
  Once you've determined a confidence interval for the
  difference of two proportions would be helpful for an
  application, there are four steps to constructing the interval:

* **Prepare.**
      Identify the sample proportions and sample sizes
      for each of the two groups,
      determine what confidence level you wish to use.
* **Check.**
      Verify the conditions to ensure each sample
      proportion is nearly normal.
      The success-failure condition should be checked
      for each group.
* **Calculate.**
      If the conditions hold, compute $SE$,
      find $z^{\star}$, and construct the interval.
* **Conclude.**
      Interpret the confidence interval in the context
      of the problem.
```
-->



```{block2, type = "example", echo = TRUE}
We consider an experiment for patients
    who underwent cardiopulmonary resuscitation (CPR)
    for a heart attack and were
    subsequently admitted to a
    hospital.
    These patients were randomly divided into a treatment
    group where they received a blood thinner or the control
    group where they did not receive a blood thinner.
    The outcome variable of interest was whether the
    patients survived for at least 24 hours.
    The results are shown in
    Figure \@ref(tab:resultsForCPRStudyInSmallSampleSection).
    Check whether we can model the difference in
    sample proportions using the normal distribution.

---
      
  We first check for independence:
  since this is a randomized experiment,
  this condition is satisfied.
  
  Next, we check the success-failure condition for
  each group.
  We have at least 10 successes and 10 failures in
  each experiment arm (11, 14, 39, 26),
  so this condition is also satisfied.

  With both conditions satisfied,
  the difference in sample proportions can be
  reasonably modeled using a normal distribution
  for these data.
```

<!--
```{r resultsForCPRStudyInSmallSampleSectionDup}
temptbl <- tribble(
 ~variable,    ~col1, ~col2, ~col3,
 "Control", "11", "39", "50",
 "Treatment", "14", "26", "40",
 "Total", "25", "65", "90"
)

temptbl %>%
 kable(caption = "Results for the CPR study.
    Patients in the treatment group were given
    a blood thinner, and patients in the control
    group were not.", 
    col.names = c("", "Survived", "Died", "Total")) %>%
 kable_styling()
```
-->

```{block2, type = "example", echo = TRUE}
Create and interpret a 90% confidence interval of the
    difference for the survival rates in the CPR study.

---
      
We'll use $p_t$ for the survival
  rate in the treatment group and $p_c$ for the control
  group:
  \begin{align*}
  \hat{p}_{t} - \hat{p}_{c}
    = \frac{14}{40} - \frac{11}{50}
    = 0.35 - 0.22
    = 0.13
  \end{align*}
  We use the standard error formula previously provided.
  As with the one-sample proportion case,
  we use the sample estimates of each proportion
  in the formula in the confidence interval context:
  \begin{align*}
  SE \approx \sqrt{\frac{0.35 (1 - 0.35)}{40} +
      \frac{0.22 (1 - 0.22)}{50}}
    = 0.095
  \end{align*}
  For a 90% confidence interval, we use $z^{\star} = 1.65$:
  \begin{align*}
  \text{point estimate} \ \pm\ z^{\star} \times SE
    \quad \to \quad 0.13 \ \pm\ 1.65 \times  0.095
    \quad \to \quad (-0.027, 0.287)
  \end{align*}
  We are 90% confident that blood thinners have
  a difference of -2.7% to +28.7% percentage point
  impact on survival rate for patients who are like
  those in the study.
  Because 0% is contained in the interval,
  we do not have enough information to say
  whether blood thinners help or harm
  heart attack patients who have been admitted after
  they have undergone CPR.
```

\index{data!CPR and blood thinner|)}



```{block2, type = "todo", echo = TRUE}
not sure why the footnote in the guidedpractice below doesn't show up?

should be right after: "Also interpret the interval in the context of the study."
```

```{block2, type = "guidedpractice", echo = TRUE}
A 5-year experiment
was conducted to evaluate the effectiveness
of fish oils on reducing cardiovascular events,
where each subject was randomized into one of two
treatment groups.
We'll consider heart attack outcomes in the patients listed in the table below:

Create a 95% confidence interval for the effect of fish oils
on heart attacks for patients who are well-represented by
those in the study.
Also interpret the interval in the context of the
study.^[Because the patients were randomized,
  the subjects are independent, both within and between
  the two groups.
  The success-failure condition is also met for both
  groups as all counts are at least 10.
  This satisfies the conditions necessary to model
  the difference in proportions using a normal distribution.

  Compute the sample proportions
  ($\hat{p}_{\text{fish oil}} = 0.0112$,
    $\hat{p}_{\text{placebo}} = 0.0155$),
  point estimate of the difference ($0.0112 - 0.0155 = -0.0043$),
  and standard error 
$SE = \sqrt{\frac{0.0112 \times 0.9888}{12933} +
      \frac{0.0155 \times 0.9845}{12938}}
    = 0.00145$. 

Next, plug the values into the general formula for
  a confidence interval, where we'll use a 95%
  confidence level with $z^{\star} = 1.96$:
  \begin{align*}
  -0.0043 \pm 1.96 \times 0.00145
      \quad \to \quad
      (-0.0071, -0.0015)
  \end{align*}
  We are 95% confident that fish oils decreases
  heart attacks by
  0.15 to 0.71 percentage points
  (off of a baseline of about 1.55%)
  over a 5-year period for subjects who are similar
  to those in the study.
  Because the interval is entirely below 0,
  the data provide strong evidence
  that fish oil supplements reduce heart attacks
  in patients like those in the study.]
```

```{r}
fish_oil_18[[3]] %>% 
  data.frame() %>%
  mutate(group = c("fish oil", "placebo")) %>%
  mutate(total = myocardioal_infarction + no_event) %>%
  column_to_rownames(var = "group") %>%
 kable(caption = "Results for the study on n-3 fatty acid supplement and related health benefits.", 
    col.names = c( "heart attack", "no event", "Total"),
    row.names = TRUE) %>%
 kable_styling()
```

#### Hypothesis tests for the difference of two proportions {-}

\index{data!mammography|(}
\index{data!breast cancer|(}

A mammogram is an X-ray procedure used to check for
breast cancer.
Whether mammograms should be used is part of a
controversial discussion, and it's the topic of our
next example where we learn about 2-proportion
hypothesis tests when $H_0$ is $p_1 - p_2 = 0$
(or equivalently, $p_1 = p_2$).

A 30-year study was conducted with nearly 90,000 female participants. During a 5-year screening period, each woman was randomized to one of two groups: in the first group, women received regular mammograms to screen for breast cancer, and in the second group, women received regular non-mammogram breast cancer exams. No intervention was made during the following 25 years of the study, and we'll consider death resulting from breast cancer over the full 30-year period. Results from the study are summarized in Figure \@ref(tab:mammogramStudySummaryTable).

If mammograms are much more effective than non-mammogram breast cancer exams, then we would expect to see additional deaths from breast cancer in the control group. On the other hand, if mammograms are not as effective as regular breast cancer exams, we would expect to see an increase in breast cancer deaths in the mammogram group.


```{r mammogramStudySummaryTable}
temptbl <- tribble(
 ~variable,    ~col1, ~col2,
 "Mammogram", "500", "44,425",
 "Control", "505", "44,405"
)

temptbl %>%
 kable(caption = "Summary results for breast cancer study.", 
    col.names = c( "", "Yes", "No")) %>%
 add_header_above( c("", "Death from breast cancer?" = 2)) %>%
 kable_styling()
```


```{block2, type = "guidedpractice", echo = TRUE}
Is this study an experiment or an observational study?^[This is an experiment. Patients were randomized
    to receive mammograms or a standard breast cancer exam.
    We will be able to make causal conclusions based on this study.]
```


```{block2, type = "guidedpractice", echo = TRUE}
Set up hypotheses to test whether there was a difference
in breast cancer deaths in the mammogram and control groups.^[$H_0$: the breast cancer death rate for patients
    screened using mammograms is the same as the breast cancer
    death rate for patients in the control,
    $p_{mgm} - p_{ctrl} = 0$.  
    $H_A$: the breast cancer death rate for patients screened
    using mammograms is different than the breast cancer death
    rate for patients in the control,
    $p_{mgm} - p_{ctrl} \neq 0$.]
```

Using the previous example,
we will check the conditions for using a normal distribution to
analyze the results of the study.
The details are very similar to that of confidence intervals.
However, when the null hypothesis is that $p_1 - p_2 = 0$,
we use a special proportion called the
**pooled proportion**\index{pooled proportion} to check the success-failure condition:
\begin{align*}
\hat{p}_{\textit{pooled}}
    &= \frac
        {\text{\# of patients who died from breast cancer in the
            entire study}}
        {\text{\# of patients in the entire study}} \\
	&= \frac{500 + 505}{500 + \text{44,425} + 505 + \text{44,405}} \\
	&= 0.0112
\end{align*}
This proportion is an estimate of the breast cancer death rate
across the entire study, and it's our best estimate of the
proportions $p_{mgm}$ and $p_{ctrl}$
*if the null hypothesis is true that $p_{mgm} = p_{ctrl}$*.
We will also use this pooled proportion when computing
the standard error.

```{r include=FALSE}
terms_chp_6 <- c(terms_chp_6, "pooled proportion")
```

```{block2, type = "example", echo = TRUE}
Is it reasonable to model the difference
    in proportions using a normal distribution in this
    study?
  
---
      
  Because the patients are randomized, they can be treated
  as independent, both within and between groups.
  We also must check the success-failure condition for each group.
  Under the null hypothesis, the proportions $p_{mgm}$
  and $p_{ctrl}$ are equal, so we check the success-failure
  condition with our best estimate of these values under $H_0$,
  the \hiddenterm{pooled proportion} from the two samples,
  $\hat{p}_{\textit{pooled}} = 0.0112$:
  \begin{align*}
  \hat{p}_{\textit{pooled}} \times n_{mgm}
      &= 0.0112 \times \text{44,925} = 503
    & (1 - \hat{p}_{\textit{pooled}}) \times n_{mgm}
      &= 0.9888 \times \text{44,925} = \text{44,422} \\
  \hat{p}_{\textit{pooled}} \times n_{ctrl}
      &= 0.0112 \times \text{44,910} = 503
    & (1 - \hat{p}_{\textit{pooled}}) \times n_{ctrl}
      &= 0.9888 \times \text{44,910} = \text{44,407}
  \end{align*}
  The success-failure condition is satisfied since
  all values are at least 10.
  With both conditions satisfied, we can safely model
  the difference in proportions using a normal
  distribution.
```

<!--
#no pooled estimates

```{block2, type = "onebox", echo = TRUE}
**Use the pooled proportion when $\pmb{H_0}$ is $p_1 - p_2 = 0$**
  
  When the null hypothesis is that the proportions are equal,
  use the pooled proportion ($\hat{p}_{\textit{pooled}}$)
  to verify the
  success-failure condition and estimate the standard error:
  \begin{eqnarray*}
  \hat{p}_{\textit{pooled}}
    = \frac{\text{number of ``successes''}}
      {\text{number of cases}}
    = \frac{\hat{p}_1 n_1 + \hat{p}_2 n_2}{n_1 + n_2}
  \end{eqnarray*}
  Here $\hat{p}_1 n_1$ represents the number of successes in
  sample 1 since
  \begin{eqnarray*}
  \hat{p}_1
    = \frac{\text{number of successes in sample 1}}{n_1}
  \end{eqnarray*}
  Similarly, $\hat{p}_2 n_2$ represents the number
  of successes in sample 2.
```

In the previous example,
the pooled proportion was used to check the success-failure
condition.
-->


```{block, type = "todo", echo = TRUE}
I don't think the malaria example is in the book yet...
^[For an example of a two-proportion
  hypothesis test that does not require the
  success-failure condition to be met, see
  Section \ref{caseStudyMalariaVaccine}.]
```


<!--
In the next example, we see the second place where the pooled
proportion comes into play: the standard error calculation.


```{block2, type = "example", echo = TRUE}
Compute the point estimate of the difference
    in breast cancer death rates in the two groups,
    and use the pooled proportion
    $\hat{p}_{\textit{pooled}} = 0.0112$ to calculate
    the standard error.
    
---
      
  The point estimate of the difference in breast cancer death
  rates is
  \begin{align*}
  \hat{p}_{mgm} - \hat{p}_{ctrl}
    &= \frac{500}{500 + 44,425} - \frac{505}{505 + 44,405} \\
    &= 0.01113 - 0.01125 \\
    &= -0.00012
  \end{align*}
  The breast cancer death rate in the mammogram group
  was 0.012% less than in the control group.
  Next, the standard error is calculated
  \emph{using the pooled proportion}, $\hat{p}_{\textit{pooled}}$:
\begin{align*}
SE = \sqrt{
      \frac{\hat{p}_{\textit{pooled}}(1-\hat{p}_{\textit{pooled}})}
          {n_{mgm}}
      + \frac{\hat{p}_{\textit{pooled}}(1-\hat{p}_{\textit{pooled}})}
          {n_{ctrl}}
    }
	= 0.00070
\end{align*}
```
-->



```{block2, type = "example", echo = TRUE}
Using the point estimate $\hat{p}_{mgm} - \hat{p}_{ctrl} = -0.00012$ and standard error $SE = 0.00070$, calculate a p-value for the hypothesis test and write a conclusion.

---
  
Just like in past tests, we first compute a test statistic and draw a picture:
\begin{align*}
Z = \frac{\text{point estimate} - \text{null value}}{SE}
	= \frac{-0.00012 - 0}{0.00070}
	= -0.17
\end{align*}

The lower tail area is 0.4325, which we double to get the p-value: 0.8650. Because this p-value is larger than 0.05, we do not reject the null hypothesis. That is, the difference in breast cancer death rates is reasonably explained by chance, and we do not observe benefits or harm from mammograms relative to a regular breast exam.
```

```{r echo=FALSE}
normTail(L = -0.17, U = 0.17,
        col = COL[1],
        axes = FALSE,
        xlim = c(-3.2, 3.2))
at <- c(-10, -2, 0, 2, 10)
labels <- c(0, -0.0014, 0, 0.0014, 0)
axis(1, at, labels, cex.axis = 0.9)
```

Can we conclude that mammograms have no benefits or harm?
Here are a few considerations to keep in mind when reviewing
the mammogram study as well as any other medical study:


* We do not accept the null hypothesis, which means
    we don't have sufficient evidence to conclude that
    mammograms reduce or increase breast cancer deaths.
* If mammograms are helpful or harmful, the data
    suggest the effect isn't very large.
* Are mammograms more or less expensive than
    a non-mammogram breast exam?
    If one option is much more expensive than the
    other and doesn't offer clear benefits,
    then we should lean towards the less expensive
    option.
* The study's authors also found that mammograms
    led to over-diagnosis of breast cancer,
    which means some breast cancers were found
    (or thought to be found) but that these cancers
    would not cause symptoms during patients' lifetimes.
    That is, something else would kill the patient
    before breast cancer symptoms appeared.
    This means some patients may have been treated
    for breast cancer unnecessarily, and this
    treatment is another cost to consider.
    It is also important to recognize that
    over-diagnosis can cause unnecessary physical
    or emotional harm to patients.

These considerations highlight the complexity around medical care and treatment recommendations. Experts and medical boards who study medical treatments use considerations like those above to provide their best recommendation based on the current evidence.

\index{data!breast cancer|)}
\index{data!mammography|)}

<!--
```{block2, type = "onebox", echo=TRUE}
**Hypothesis testing when ${H_0}$ is $p_1 - p_2 = 0$.**
  
  Once you've determined a hypothesis test for the difference
  of two proportions is the correct procedure, there are four
  steps to completing the test:

* **Prepare.**
      Identify the parameter of interest,
      list out hypotheses,
      identify the significance level,
      and compute summary statistics for each group.
* **Check.**
      Verify the conditions to ensure
      $\hat{p}_1 - \hat{p}_2$ is nearly normal under $H_0$.
      When the null hypothesis is that the difference is 0,
      use a pooled proportion to check the success-failure
      condition for each group.
* **Calculate.**
      If the conditions hold, compute the standard
      error, again using the pooled proportion,
      compute the Z-score, and identify the p-value.
* **Conclude.**
      Evaluate the hypothesis test by comparing the p-value
      to $\alpha$, and provide a conclusion in the context
      of the problem.
```
-->

<!--
\subsection{More on 2-proportion hypothesis tests (special topic)}

When we conduct a 2-proportion hypothesis test,
usually $H_0$ is $p_1 - p_2 = 0$. However, there are rare
situations where we want to check for some difference in
$p_1$ and $p_2$ that is some value other than 0.
For example, maybe we care about checking a null hypothesis
where $p_1 - p_2 = 0.1$. %\footnote{We can
%  also encounter a similar situation with a difference of
%  two means, though no such example is given in
%  Chapter \ref{inferenceForNumericalData} since the methods
%  remain exactly the same in the context of sample means.
%  On the other hand, the success-failure condition and the
%  calculation of the standard error vary slightly in different
%  proportion contexts.}
In contexts like these, we generally use $\hat{p}_1$ and
$\hat{p}_2$ to check the success-failure condition and
construct the standard error.

\begin{exercisewrap}
\begin{nexercise}
\label{carWheelBladeManufacturer}%
A quadcopter company is considering a new manufacturer
for rotor blades.
The new manufacturer would be more expensive,
but they claim
their higher-quality blades are more reliable,
with 3% more blades passing inspection than their
competitor.
Set up appropriate hypotheses for the test.\footnotemark
\end{nexercise}
\end{exercisewrap}
\footnotetext{$H_0$: The higher-quality blades will pass
  inspection 3% more frequently than the standard-quality blades.
  $p_{highQ} - p_{standard} = 0.03$.
  $H_A$: The higher-quality blades will pass inspection
  some amount different than 3% more often than the
  standard-quality blades.
  $p_{highQ} - p_{standard} \neq 0.03$.}

\setlength{\captionwidth}{85mm}

\begin{figure}[h]
\centering
\Figures{0.6}{quadcopter}{quadcopter_david_j}
\caption{A Phantom quadcopter.\vspace{-1mm} \\
   -----------------------------\vspace{-2mm}\\
   {\footnotesize Photo by David J
   (\oiRedirect{textbook-quadcopter_david_j}
       {http://flic.kr/p/oiWLNu}).
   \oiRedirect{textbook-CC_BY_2}{CC-BY 2.0 license.}
   This photo has been cropped and a border has been added.}}
\label{quadcopter_david_j}
\end{figure}

\setlength{\captionwidth}{\mycaptionwidth}

\D{\newpage}

%\Add{In Guided Practice \ref{qualityCtrlEngHypothesisEval}, the null difference is 0.03. However, in the vast majority of applications for differences in means or proportions, the null difference is 0. While the details for a difference of means does not change if the null difference is zero or non-zero, that is not the case for a difference in proportions. As we'll see in Section \ref{}, a hypothesis test for a difference in proportions where the null value is 0 requires additional care.}

\begin{examplewrap}
\begin{nexample}{The quality control engineer from
    Guided Practice \ref{carWheelBladeManufacturer}
    collects a sample of blades, examining 1000 blades
    from each company, and she finds that 899 blades pass
    inspection from the current supplier and 958 pass
    inspection from the prospective supplier.
    Using these data, evaluate the hypotheses from
    Guided Practice \ref{carWheelBladeManufacturer}
    with a significance level of 5%.}
  \label{qualityCtrlEngHypothesisEval}%
  First, we check the conditions.
  The sample is not necessarily random, so to proceed
  we must assume the blades are all independent;
  for this sample we will suppose this assumption
  is reasonable, but the engineer would be more knowledgeable
  as to whether this assumption is appropriate.
  The success-failure condition also holds for each sample.
  Thus, the difference in sample proportions,
  $0.958 - 0.899 = 0.059$, can be said to come from a nearly
  normal distribution.

  The standard error is computed using the two sample
  proportions since we do not use a pooled proportion
  for this context:
  \begin{align*}
  SE
    = \sqrt{\frac{0.958(1-0.958)}{1000} +
        \frac{0.899(1-0.899)}{1000}}
    = 0.0114
  \end{align*}
  In this hypothesis test, because the null is that
  $p_1 - p_2 = 0.03$, the sample proportions were used
  for the standard error calculation rather than a pooled
  proportion.

  Next, we compute the test statistic and use it to find the
  p-value, which is depicted in
  Figure \ref{bladesTwoSampleHTPValueQC}.
  \begin{align*}
  Z = \frac{\text{point estimate} - \text{null value}}{SE}
    = \frac{0.059 - 0.03}{0.0114} = 2.54
  \end{align*}
  Using a standard normal distribution for this test statistic,
  we identify the right tail area as 0.006,
  and we double it to get the p-value: 0.012.
  We reject the null hypothesis because 0.012 is less than 0.05.
  Since we observed a larger-than-3% increase in blades
  that pass inspection, we have statistically significant
  evidence that the higher-quality blades pass inspection
  \emph{more than} 3% as often as the currently used blades,
  exceeding the company's claims.
\end{nexample}
\end{examplewrap}

\begin{figure}[h]
  \centering
  \Figure{0.45}{bladesTwoSampleHTPValueQC}
  \caption{Distribution of the test statistic if the null
      hypothesis was true.
      The p-value is represented by the shaded areas.}
  \label{bladesTwoSampleHTPValueQC}
\end{figure}


-->








<!--
## Testing for goodness of fit using chi-square (special topic, include simulation version)
-->


## Testing for independence in two-way tables

Note that with two-way tables, there is not an obvious single parameter of interest. 
Instead, research questions usually focus on how the proportions of the response variable changes (or not) across the different levels of the explanatory variable. 
Because there is not a population parameter to estimate, bootstrapping to find the standard error of the estimate is not meaningful.
As such, for two-way tables, we will focus on the randomization test and corresponding mathematical approximation (and not bootstrapping).

### Two-way table: randomization test

### Two-way table: mathematical model


## Chapter 6 review {#chp6-review}

### Terms

We introduced the following terms in the chapter. 
If you're not sure what some of these terms mean, we recommend you go back in the text and review their definitions.
We are purposefully presenting them in alphabetical order, instead of in order of appearance, so they will be a little more challenging to locate. 
However you should be able to easily spot them as **bolded text**.

```{r}
make_terms_table(terms_chp_6)
```

