# Inference for regression {#inference-reg}

::: {.uptohere}
This chapter is currently under construction, however the content to be presented in this chapter is covered in the R tutorials. We encourage you to review the content there in the meantime. Click [here](https://openintrostat.github.io/ims-tutorials/08-inference-for-regression/) to take a look! 
:::

::: {.chapterintro}
Chapter intro
:::


## Inference for linear regression

### Randomization test for $H_0: \beta_1= 0$ {#randslope}


#### Observed data {-}

#### Variability of the statistic {-}

#### Observed statistic vs. null statistics {-}


### Bootstrap confidence interval for $\beta_1$ {#bootbeta1}



#### Observed data {-}

#### Variability of the statistic {-}



### Mathematical model {#mathslope}

t-distribution for regression coefficients

#### Observed data {-}

#### Variability of the statistic {-}

#### Observed statistic vs. null statistics {-}


#### Observed data {-}

#### Variability of the statistic {-}


<!--
- Model Comparison: Occam’s Razor and $R^2 > R^2_{adj}$
-->


## Checking model assumptions using graphs

- L-I-N-E

## Inference for multiple regression

- residuals vs. fitted instead of residuals vs. x

## Inference for logistic regression

- t-dist only

## Chapter review {#chp8-review}

### Terms

We introduced the following terms in the chapter. 
If you're not sure what some of these terms mean, we recommend you go back in the text and review their definitions.
We are purposefully presenting them in alphabetical order, instead of in order of appearance, so they will be a little more challenging to locate. 
However you should be able to easily spot them as **bolded text**.

```{r eval=FALSE}
make_terms_table(terms_chp_8)
```

### Chapter exercises

::: {.sectionexercise}

```{r intro, child="08-exercises/08-05-chapter-review.Rmd", eval=FALSE}
```

:::

### Interactive R tutorials

Navigate the concepts you've learned in this chapter in R using the following self-paced tutorials. 
All you need is your browser to get started!

::: {.alltutorials}
[Tutorial 8: Inference for regression](https://openintrostat.github.io/ims-tutorials/08-inference-for-regression/)
:::

::: {.singletutorial}
[Tutorial 8 - Lesson 1: Inference in regression](https://openintro.shinyapps.io/ims-08-inference-for-regression-01/)
:::

::: {.singletutorial}
[Tutorial 8 - Lesson 2: Randomization test for slope](https://openintro.shinyapps.io/ims-08-inference-for-regression-02/)
:::

::: {.singletutorial}
[Tutorial 8 - Lesson 3: t-test for slope](https://openintro.shinyapps.io/ims-08-inference-for-regression-03/)
:::

::: {.singletutorial}
[Tutorial 8 - Lesson 4: Checking technical conditions for slope inference](https://openintro.shinyapps.io/ims-08-inference-for-regression-04/)
:::

::: {.singletutorial}
[Tutorial 8 - Lesson 5: Inference beyond the simple linear regression model](https://openintro.shinyapps.io/ims-08-inference-for-regression-05/)
:::

You can also access the full list of tutorials supporting this book [here](https://openintrostat.github.io/ims-tutorials/).

### R labs

Further apply the concepts you've learned in this chapter in R with computational labs that walk you through a data analysis case study.

::: {.singlelab}
[Multiple linear regression - Grading the professor](http://openintrostat.github.io/oilabs-tidy/09_multiple_regression/multiple_regression.html)
:::

::: {.alllabs}
[Full list of labs supporting OpenIntro::Introduction to Modern Statistics](http://openintrostat.github.io/oilabs-tidy/)
:::
