# Exercise solutions {#exercise-solutions}

::: {.section-exercise-solution}
<!-- Chapter 1 -->

## Chapter 1

### Case study: stents and strokes

```{r child="01/short-answers/01-01-case-study-short-answers.Rmd"}
```

### Data basics

```{r child="01/short-answers/01-02-data-basics-short-answers.Rmd"}
```

### Sampling principles and strategies

```{r child="01/short-answers/01-03-sampling-short-answers.Rmd"}
```

### Experiments

```{r child="01/short-answers/01-04-experiments-short-answers.Rmd"}
```

### Chapter review

```{r child="01/short-answers/01-05-chapter-review-short-answers.Rmd"}
```

<!-- Chapter 2 -->

## Chapter 2

### Exploring numerical data

```{r child="02/short-answers/02-01-numerical-data-short-answers.Rmd"}
```

### Exploring categorical data

```{r child="02/short-answers/02-02-categorical-data-short-answers.Rmd"}
```

### Effective data visualization

```{r child="02/short-answers/02-03-effective-dataviz-short-answers.Rmd", eval=FALSE}
```

### Case study: malaria vaccine

```{r child="02/short-answers/02-04-case-study-short-answers.Rmd"}
```

### Chapter review

```{r child="02/short-answers/02-05-chapter-review-short-answers.Rmd"}
```

<!-- Chapter 3 -->

## Chapter 3

### Fitting a line, residuals, and correlation

```{r child="03/short-answers/03-01-fit-line-res-cor-short-answers.Rmd"}
```

### Least squares regression

```{r child="03/short-answers/03-02-least-squares-regression-short-answers.Rmd"}
```

### Outliers in linear regression

```{r child="03/short-answers/03-03-outliers-in-regression-short-answers.Rmd"}
```

### Chapter review

<!-- Chapter 4 -->

## Chapter 4

### Regression with multiple predictors

### Model selection

### Model diagnostics

### Case study: Mario Kart

### Logistic regression

### Chapter review

<!-- Chapter 5 -->

## Chapter 5

### Randomization tests

### Bootstrap confidence intervals

### Mathematical models

### Chapter review

<!-- Chapter 6 -->

## Chapter 6

### One proportion

### Difference of two proportions

### Independence in two-way tables

### Chapter review

<!-- Chapter 7 -->

## Chapter 7

### One mean

### Difference of two neabs

### Paired difference

### t-procedures

### Comparing many means with ANOVA

### Chapter review

<!-- Chapter 8 -->

## Chapter 8

### Inference for linear regression

### Checking model assumptions

### Inference for multiple regression

### Inference for logistic regression

### Chapter review
:::
