\addtocontents{toc}{\protect\setcounter{tocdepth}{0}}

# (APPENDIX) Appendix {.unnumbered}

```{r, include = FALSE}
source("_common.R")
```

# Exercise solutions {#exercise-solutions}

## Chapter 1 {#exercise-solutions-01 .unlisted}

::: {.exercises-solution}
```{r sa-01, child = "exercises/01-sa-data-hello.Rmd"}
```
:::

## Chapter 2 {#exercise-solutions-02 .unlisted}

::: {.exercises-solution}
```{r sa-02, child = "exercises/02-sa-data-design.Rmd"}
```
:::

## Chapter 3 {#exercise-solutions-03 .unlisted}

Application chapter, no exercises.

## Chapter 4 {#exercise-solutions-04 .unlisted}

::: {.exercises-solution}
```{r sa-04, child = "exercises/04-sa-explore-categorical.Rmd"}
```
:::

## Chapter 5 {#exercise-solutions-05 .unlisted}

::: {.exercises-solution}
```{r sa-05, child = "exercises/05-sa-explore-numerical.Rmd"}
```
:::

## Chapter 6 {#exercise-solutions-06 .unlisted}

Application chapter, no exercises.

## Chapter 7 {#exercise-solutions-07 .unlisted}

::: {.exercises-solution}
```{r sa-07, child = "exercises/07-sa-model-slr.Rmd"}
```
:::

## Chapter 8 {#exercise-solutions-08 .unlisted}

::: {.exercises-solution}
```{r sa-08, child = "exercises/08-sa-model-mlr.Rmd"}
```
:::

## Chapter 9 {#exercise-solutions-09 .unlisted}

::: {.exercises-solution}
```{r sa-09, child = "exercises/09-sa-model-logistic.Rmd"}
```
:::

## Chapter 10 {#exercise-solutions-10 .unlisted}

Application chapter, no exercises.

## Chapter 11 {#exercise-solutions-11 .unlisted}

::: {.exercises-solution}
```{r sa-11, child = "exercises/11-sa-foundations-randomization.Rmd"}
```
:::

## Chapter 12 {#exercise-solutions-12 .unlisted}

::: {.exercises-solution}
```{r sa-12, child = "exercises/12-sa-foundations-bootstrapping.Rmd"}
```
:::

## Chapter 13 {#exercise-solutions-13 .unlisted}

::: {.exercises-solution}
```{r sa-13, child = "exercises/13-sa-foundations-mathematical.Rmd"}
```
:::

## Chapter 14 {#exercise-solutions-14 .unlisted}

::: {.exercises-solution}
```{r sa-14, child = "exercises/14-sa-foundations-errors.Rmd", eval = FALSE}
```
:::

## Chapter 15 {#exercise-solutions-15 .unlisted}

Application chapter, no exercises.

## Chapter 16 {#exercise-solutions-16 .unlisted}

::: {.exercises-solution}
```{r sa-16, child = "exercises/16-sa-inference-one-prop.Rmd"}
```
:::

## Chapter 17 {#exercise-solutions-17 .unlisted}

::: {.exercises-solution}
```{r sa-17, child = "exercises/17-sa-inference-two-props.Rmd", eval = FALSE}
```
:::

## Chapter 18 {#exercise-solutions-18 .unlisted}

::: {.exercises-solution}
```{r sa-18, child = "exercises/18-sa-inference-tables.Rmd", eval = FALSE}
```
:::

## Chapter 19 {#exercise-solutions-19 .unlisted}

::: {.exercises-solution}
```{r sa-19, child = "exercises/19-sa-inference-one-mean.Rmd"}
```
:::

## Chapter 20 {#exercise-solutions-20 .unlisted}

::: {.exercises-solution}
```{r sa-20, child = "exercises/20-sa-inference-two-means.Rmd"}
```
:::

## Chapter 21 {#exercise-solutions-21 .unlisted}

::: {.exercises-solution}
```{r sa-21, child = "exercises/21-sa-inference-paired-means.Rmd"}
```
:::

## Chapter 22 {#exercise-solutions-22 .unlisted}

::: {.exercises-solution}
```{r sa-22, child = "exercises/22-sa-inference-many-means.Rmd"}
```
:::

## Chapter 23 {#exercise-solutions-23 .unlisted}

Application chapter, no exercises.

## Chapter 24 {#exercise-solutions-24 .unlisted}

::: {.exercises-solution}
```{r sa-24, child = "exercises/24-sa-inf-model-slr.Rmd"}
```
:::

## Chapter 25 {#exercise-solutions-25 .unlisted}

::: {.exercises-solution}
```{r sa-25, child = "exercises/25-sa-inf-model-mlr.Rmd"}
```
:::

## Chapter 26 {#exercise-solutions-26 .unlisted}

::: {.exercises-solution}
```{r sa-26, child = "exercises/26-sa-inf-model-logistic.Rmd"}
```
:::
