# (APPENDIX) Appendix {.unnumbered}

```{r, include = FALSE}
source("_common.R")
```

# Exercise solutions {#exercise-solutions}

## Chapter 1 {.unlisted .unnumbered}

::: {.exercises-solution}
```{r sa-01, child = "exercises/01-sa-data-hello.Rmd"}
```
:::

## Chapter 2 {.unlisted .unnumbered}

::: {.exercises-solution}
```{r sa-02, child = "exercises/02-sa-data-design.Rmd"}
```
:::

## Chapter 3 {.unlisted .unnumbered}

Application chapter, no exercises.

## Chapter 4 {.unlisted .unnumbered}

::: {.exercises-solution}
```{r sa-04, child = "exercises/04-sa-explore-categorical.Rmd"}
```
:::

## Chapter 5 {.unlisted .unnumbered}

::: {.exercises-solution}
```{r sa-05, child = "exercises/05-sa-explore-numerical.Rmd"}
```
:::

## Chapter 6 {.unlisted .unnumbered}

::: {.exercises-solution}
```{r sa-06, child = "exercises/20-sa-inference-one-mean.Rmd", eval = FALSE}
```
:::

## Chapter 7 {.unlisted .unnumbered}

Application chapter, no exercises.

## Chapter 8 {.unlisted .unnumbered}

::: {.exercises-solution}
```{r sa-08, child = "exercises/08-sa-model-slr.Rmd"}
```
:::

## Chapter 9 {.unlisted .unnumbered}

::: {.exercises-solution}
```{r sa-09, child = "exercises/09-sa-model-mlr.Rmd"}
```
:::

## Chapter 10 {.unlisted .unnumbered}

::: {.exercises-solution}
```{r sa-10, child = "exercises/10-sa-model-logistic.Rmd"}
```
:::

## Chapter 11 {.unlisted .unnumbered}

Application chapter, no exercises.

## Chapter 12 {.unlisted .unnumbered}

::: {.exercises-solution}
```{r sa-12, child = "exercises/20-sa-inference-one-mean.Rmd", eval = FALSE}
```
:::

## Chapter 13 {.unlisted .unnumbered}

::: {.exercises-solution}
```{r sa-13, child = "exercises/20-sa-inference-one-mean.Rmd", eval = FALSE}
```
:::

## Chapter 14 {.unlisted .unnumbered}

::: {.exercises-solution}
```{r sa-14, child = "exercises/20-sa-inference-one-mean.Rmd", eval = FALSE}
```
:::

## Chapter 15 {.unlisted .unnumbered}

::: {.exercises-solution}
```{r sa-15, child = "exercises/20-sa-inference-one-mean.Rmd", eval = FALSE}
```
:::

## Chapter 16 {.unlisted .unnumbered}

Application chapter, no exercises.

## Chapter 17 {.unlisted .unnumbered}

::: {.exercises-solution}
```{r sa-17, child = "exercises/17-sa-inference-one-prop.Rmd"}
```
:::

## Chapter 18 {.unlisted .unnumbered}

::: {.exercises-solution}
```{r sa-18, child = "exercises/20-sa-inference-one-mean.Rmd", eval = FALSE}
```
:::

## Chapter 19 {.unlisted .unnumbered}

::: {.exercises-solution}
```{r sa-19, child = "exercises/20-sa-inference-one-mean.Rmd", eval = FALSE}
```
:::

## Chapter 20 {.unlisted .unnumbered}

::: {.exercises-solution}
```{r sa-20, child = "exercises/20-sa-inference-one-mean.Rmd"}
```
:::

## Chapter 21 {.unlisted .unnumbered}

::: {.exercises-solution}
```{r sa-21, child = "exercises/21-sa-inference-two-means.Rmd"}
```
:::

## Chapter 22 {.unlisted .unnumbered}

::: {.exercises-solution}
```{r sa-22, child = "exercises/22-sa-inference-paired-means.Rmd"}
```
:::

## Chapter 23 {.unlisted .unnumbered}

::: {.exercises-solution}
```{r sa-23, child = "exercises/20-sa-inference-one-mean.Rmd"}
```
:::

## Chapter 24 {.unlisted .unnumbered}

Application chapter, no exercises.

## Chapter 25 {.unlisted .unnumbered}

::: {.exercises-solution}
```{r sa-25, child = "exercises/20-sa-inference-one-mean.Rmd"}
```
:::

## Chapter 26 {.unlisted .unnumbered}

::: {.exercises-solution}
```{r sa-26, child = "exercises/20-sa-inference-one-mean.Rmd"}
```
:::

## Chapter 27 {.unlisted .unnumbered}

::: {.exercises-solution}
```{r sa-27, child = "exercises/20-sa-inference-one-mean.Rmd"}
```
:::
