---
knit: bookdown::render_book
site: bookdown::bookdown_site
title: "Introduction to Modern Statistics"
author: "Mine Çetinkaya-Rundel and Johanna Hardin"
description: "This is IMS, aka ISRS2!"
github-repo: openintrostat/ims
twitter-handle: OpenIntroOrg
url: 'https\://openintro-ims.netlify.app/'
link-citations: yes
bibliography: [book.bib, packages.bib, eoce.bib]
biblio-style: apalike
monofont: "Source Code Pro"
always_allow_html: true
documentclass: book
classoption: 
  - 10pt
  - openany
---

```{r, include = FALSE}
source("_common.R")
```

```{r, include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(.packages(), 
                   'bookdown', 'knitr', 
                   'rmarkdown'), 'packages.bib')
```

::: {.welcome}
```{asis, echo = knitr::is_html_output()}
# Welcome to IMS {.unnumbered}
```

\chapter*{}

\vfill

```{asis, echo = knitr::is_html_output()}
This is the website for **Introduction to Modern Statistics** by Çetinkaya-Rundel and Hardin. Introduction to Modern Statistics, which we'll refer to as IMS going forward, is a textbook from the [OpenIntro](https://www.openintro.org/) project. The book will always be available for free here and will soon also be available in PDF (for free or for the amount you choose to donate to the OpenIntro project) on Leanpub and in hard copy for purchase.

---
```

```{asis, echo = knitr::is_html_output()}
Copyright &#169;	 2021.
```

```{asis, echo = knitr::is_latex_output()}
Copyright $\copyright$ 2021.
```

First Edition.

Version date: `r format(Sys.Date(), "%B %d, %Y")`.

This textbook and its supplements, including slides, labs, and interactive tutorials, may be downloaded for free at\
[**openintro.org/book/ims**](http://openintro.org/book/ims).

This textbook is a derivative of *OpenIntro Statistics* 4th Edition and *Introduction to Statistics with Randomization and Simulation* 1st Edition by Diez, Barr, and Çetinkaya-Rundel, and it's available under a Creative Commons Attribution-ShareAlike 3.0 Unported United States License.
License details are available at the Creative Commons website:\
[**creativecommons.org**](https://www.openintro.org/go/?id=creativecommons_org&referrer=ims1_pdf).

Source files for this book may be found on GitHub at\
[**github.com/openintrostat/ims**](https://github.com/openintrostat/ims).
:::
