---
knit: bookdown::render_book
site: bookdown::bookdown_site
title: "Introduction to Modern Statistics"
author: "Mine Çetinkaya-Rundel and Johanna Hardin"
description: "This is IMS, aka ISRS2!"
date: "`r Sys.Date()`"
github-repo: openintrostat/ims
twitter-handle: OpenIntroOrg
url: 'https\://openintro-ims.netlify.app/'
link-citations: yes
bibliography: [book.bib, packages.bib, eoce.bib]
biblio-style: apalike
monofont: "Source Code Pro"
always_allow_html: true
documentclass: book
classoption: 
  - 10pt
  - openany
---

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

```{r, include = FALSE}
source("_common.R")
```

::: {.welcome}
```{asis, echo = knitr::is_html_output()}
# Welcome {.unnumbered}
```

\chapter*{}

\vfill

```{r}
library(r2symbols)
```

```{asis, echo = knitr::is_html_output()}
Copyright &#169;	 2021.
```

```{asis, echo = knitr::is_latex_output()}
Copyright $\copyright$ 2021.
```

First Edition .

Version date: `r format(Sys.Date(), "%B %d, %Y")`.

This textbook and its supplements, including slides, labs, and interactive tutorials, may be downloaded for free at\
[**openintro.orh/book/ims**](http://openintro.org/book/ims).

This textbook is a derivative of *OpenIntro Statistics* 4th Edition and *Introduction to Statistics with Randomization and Simulation* 1st Edition by Diez, Barr, and Çetinkaya-Rundel, and it's available under a Creative Commons Attribution-ShareAlike 3.0 Unported United States License.
License details are available at the Creative Commons website:\
[**creativecommons.org**](https://www.openintro.org/go/?id=creativecommons_org&referrer=ims1_pdf).

Source files for this book may be found on GitHub at\
[**github.com/openintrostat/ims**](https://github.com/openintrostat/ims).
:::
