1.  **Visualize the residuals.** 
The scatterplots shown below each have a superimposed regression line. If we were to construct a residual plot (residuals versus $x$) for each, describe what those plots would look like.

    ```{r}
    library(openintro)
    library(tidyverse)
    
    set.seed(84628)
    
    df <- tibble(
      x = seq(1, 100, 1),
      y_linear = 3 * x + 5 + rnorm(length(x), mean = 0, sd = 20),
      y_fan_back = 4 * x + 5 + rnorm(length(x), mean = 0, sd = sort(1.3 * x,     decreasing = TRUE))
    ) %>%
      pivot_longer(cols = -x, names_to = "type", values_to = "y") %>%
      arrange(type) %>%
      mutate(type_label = if_else(type == "y_linear", "(a)", "(b)"))
    
    ggplot(df, aes(x = x, y = y)) +
      geom_point(color = COL["blue", "full"], size = 2, alpha = 0.8) +
      geom_smooth(method = "lm", se = FALSE, color = "black", size = 0.8) +
      facet_wrap(~type_label) +
      theme_minimal() +
      theme(axis.text = element_blank()) +
      labs(x = NULL, y = NULL)
    ```

1.  **Trends in the residuals.** 
Shown below are two plots of residuals remaining after fitting a linear model to two different sets of data. 
Describe important features and determine if a linear model would be appropriate for these data. 
Explain your reasoning.

    ```{r}
    library(openintro)
    library(tidyverse)
    library(broom)
    
    set.seed(8313)
    
    df <- tibble(
        x = seq(1:300),
        y_fan = 4*x + 5 + rnorm(length(x), mean = 0, sd = sort(1.3*x)),
        y_log = log(x) + rnorm(300, mean = 0, sd = 0.5)
    )
    
    m_fan_aug <- augment(lm(y_fan ~ x, data = df)) %>% mutate(type = "fan")
    m_log_aug <- augment(lm(y_log ~ x, data = df)) %>% mutate(type = "log")
    df        <- bind_rows(m_fan_aug, m_log_aug) %>%
      select(x, y = .resid, type) %>%
      mutate(type_label = if_else(type == "fan", "(a)", "(b)"))
    
    ggplot(df, aes(x = x, y = y)) +
      geom_point(color = COL["blue","full"], size = 2, alpha = 0.8) +
      geom_hline(yintercept = 0, linetype = "dashed") +
      facet_wrap(~type_label, scales = "free_y") +
      theme_minimal() +
      theme(axis.text = element_blank()) +
      labs(x = NULL, y = NULL)
    ```

1.  **Identify relationships, Part I.** 
For each of the six plots, identify the strength of the relationship (e.g. weak, moderate, or strong) in the data and whether fitting a linear model would be reasonable.

    ```{r out.width="100%"}
    library(openintro)
    library(tidyverse)
    library(patchwork)
    
    set.seed(9271)
    
    df1 <- tibble(
      x              = seq(0, 6, by = 0.05),
      u              = (x-3)^2 - 4 + rnorm(length(x), mean = 0, sd = 1),
      lin_pos_strong = 3*x + 10 + rnorm(length(x), mean = 0, sd = 2),
      lin_pos_weak   = 3*x + 10 + rnorm(length(x), mean = 0, sd = 20)
      ) %>%
      pivot_longer(cols = -x, names_to = "type", values_to = "y") %>%
      arrange(type) %>%
      mutate(type_label = case_when(
          type == "u"              ~ "(a)",
          type == "lin_pos_strong" ~ "(b)",
          type == "lin_pos_weak"   ~ "(c)"
      ))
    
    df2 <- tibble(
      x              = seq(-8, -2, by = 0.05),
      n              = -1 * (x + 5)^2 + 1 + rnorm(length(x), mean = 0, sd = 2),
      lin_neg_strong = -5 * x + 3 + rnorm(length(x), mean = 0, sd = 2),
      none           = rnorm(length(x), mean = 0, sd = 10)
      ) %>%
      pivot_longer(cols = -x, names_to = "type", values_to = "y") %>%
      arrange(type) %>%
      mutate(type_label = case_when(
          type == "n"              ~ "(d)",
          type == "lin_neg_strong" ~ "(e)",
          type == "none"           ~ "(f)"
      ))
        
    p_1 <- ggplot(df1, aes(x = x, y = y)) +
      geom_point(color = COL["blue", "full"], size = 2, alpha = 0.8) +
      facet_wrap(~type_label, scales = "free_y") +
      theme_minimal() +
      theme(axis.text = element_blank()) +
      labs(x = NULL, y = NULL)
    
    p_2 <- ggplot(df2, aes(x = x, y = y)) +
      geom_point(color = COL["blue", "full"], size = 2, alpha = 0.8) +
      facet_wrap(~type_label, scales = "free_y") +
      theme_minimal() +
      theme(axis.text = element_blank()) +
      labs(x = NULL, y = NULL)
    
    p_1 / p_2
    ```

1.  **Identify relationships, Part II.** 
For each of the six plots, identify the strength of the relationship (e.g. weak, moderate, or strong) in the data and whether fitting a linear model would be reasonable.

    ```{r}
    library(openintro)
    library(tidyverse)
    
    set.seed(9271)
    
    df <- tibble(
      x              = seq(-3, 4, 0.05),
      s              = -0.5 * x^3 + x^2 + x + rnorm(length(x), mean = 0, sd = 2),
      hockey_stick   = 2 * x^4 + -0.5 * x^3 + x^2 + x + rnorm(length(x), mean = 0, sd = 30),
      pos_lin_strong = 3 * x + rnorm(length(x), mean = 0, sd = 2),
      pos_weak       = 3 * x + rnorm(length(x), mean = 0, sd = 20),
      pos_weaker     = -3 * x + rnorm(length(x), mean = 0, sd = 10),
      neg_lin_weak   = -3 * x + rnorm(length(x), mean = 0, sd = 5) 
      ) %>%
      pivot_longer(cols = -x, names_to = "type", values_to = "y") %>%
      arrange(type) %>%
      mutate(type_label = case_when(
          type == "s"              ~ "(a)",
          type == "hockey_stick"   ~ "(b)",
          type == "pos_lin_strong" ~ "(c)",
          type == "pos_weak"       ~ "(d)",
          type == "pos_weaker"     ~ "(e)",
          type == "neg_lin_weak"   ~ "(f)",
      ))
        
    ggplot(df, aes(x = x, y = y)) +
      geom_point(color = COL["blue", "full"], size = 2, alpha = 0.8) +
      facet_wrap(~type_label, scales = "free_y", nrow = 2) +
      theme_minimal() +
      theme(axis.text = element_blank()) +
      labs(x = NULL, y = NULL)
    ```

1.  **Midterms and final.** 
The two scatterplots below show the relationship between the overall course average and two midterm exams (Exam 1 and Exam 2) recorded for 233 students during several years for a statistics course at a university.^[The data used in this exercise can be found in the **openintro** package: [`exam_grades`](http://openintrostat.github.io/openintro/reference/exam_grades.html).]

    ```{r}
    library(openintro)
    library(tidyverse)
    library(patchwork)
    
    p_1 <- ggplot(exam_grades, aes(x = exam1, y = course_grade)) +
      geom_point(color = COL["blue", "full"], size = 2, alpha = 0.8) +
      theme_minimal() +
      labs(x = "Exam 1 grade", y = "Course grade")
    
    p_2 <- ggplot(exam_grades, aes(x = exam2, y = course_grade)) +
      geom_point(color = COL["blue", "full"], size = 2, alpha = 0.8) +
      theme_minimal() +
      labs(x = "Exam 2 grade", y = "Course grade")
    
    p_1 + p_2
    ```

    a.  Based on these graphs, which of the two exams has the strongest correlation with the course grade? Explain.

    b.  Can you think of a reason why the correlation between the exam you chose in part (a) and the course grade is higher?

1.  **Partners' ages and heights.** 
The Great Britain Office of Population Census and Surveys collected data on a random sample of 170 married couples in Britain, recording the age (in years) and heights (converted here to inches) of the partners. The scatterplot on the left shows the heights of the partners plotted against each other and the plot on the right shows the ages of the partners plotted against each other.^[The data used in this exercise can be found in the **openintro** package: [`husbands_wives`](http://openintrostat.github.io/openintro/reference/husbands_wives.html).]

    ```{r}
    library(tidyverse)
    library(openintro)
    library(patchwork)
    
    p_height <- ggplot(husbands_wives, aes(x = ht_husband / 25.4, y = ht_wife / 25.4)) +
      geom_point(color = COL["blue","full"], alpha = 0.8) +
      theme_minimal() +
      labs(x = "One partner's height (inches)", y = "The other partner's height (inches)")
    
    p_age <- ggplot(husbands_wives, aes(x = age_husband, y = age_wife)) +
      geom_point(color = COL["blue","full"], alpha = 0.8) +
      theme_minimal() +
      labs(x = "One partner's age (years)", y = "The other partner's age (years)")
    
    p_height + p_age
    ```

    a.  Describe the relationship between partners' ages.

    b.  Describe the relationship between partners' heights.

    c.  Which plot shows a stronger correlation? Explain your reasoning.

    d.  Data on heights were originally collected in centimeters, and then converted to inches. Does this conversion affect the correlation between partners' heights?

1.  **Match the correlation, Part I.** 
Match each correlation to the corresponding scatterplot.^[The data used in this exercise can be found in the **openintro** package: [`corr_match`](http://openintrostat.github.io/openintro/reference/corr_match.html).]

    ```{r out.width="100%", fig.asp=0.3}
    library(openintro)
    library(tidyverse)
    library(glue)
    
    corr_match %>%
      select(x, y1:y4) %>%
      pivot_longer(cols = -x, names_to = "type", values_to = "y") %>%
      mutate(
        type_label = str_remove(type, "y"),
        type_label = glue("({type_label})")
        ) %>%
      ggplot(aes(x = x, y = y)) +
      geom_point(color = COL["blue","full"], alpha = 0.8) +
      facet_wrap(~type_label, scales = "free_y", nrow = 1) +
      theme_minimal() +
      labs(x = NULL, y = NULL) +
      theme(axis.text = element_blank())
    ```

    a.  $r = -0.7$

    b.  $r = 0.45$

    c.  $r = 0.06$

    d.  $r = 0.92$

1.  **Match the correlation, Part II.** 
Match each correlation to the corresponding scatterplot.^[The data used in this exercise can be found in the **openintro** package: [`corr_match`](http://openintrostat.github.io/openintro/reference/corr_match.html).]

    ```{r out.width="100%", fig.asp=0.3}
    library(openintro)
    library(tidyverse)
    library(glue)
    
    corr_match %>%
      select(x, y5:y8) %>%
      pivot_longer(cols = -x, names_to = "type", values_to = "y") %>%
      mutate(
        type_label = str_remove(type, "y"),
        type_label = glue("({type_label})")
        ) %>%
      ggplot(aes(x = x, y = y)) +
      geom_point(color = COL["blue","full"], alpha = 0.8) +
      facet_wrap(~type_label, scales = "free_y", nrow = 1) +
      theme_minimal() +
      labs(x = NULL, y = NULL) +
      theme(axis.text = element_blank())
    ```

    a.  $r = 0.49$

    b.  $r = -0.48$

    c.  $r = -0.03$

    d.  $r = -0.85$

1.  **Body measurements, correlation.** 
Researchers studying anthropometry collected body and skeletal diameter measurements, as well as age, weight, height and sex for 507 physically active individuals. 
The scatterplot below shows the relationship between height and shoulder girth (circumference of shoulders measured over deltoid muscles), both measured in centimeters.^[The data used in this exercise can be found in the **openintro** package: [`bdims`](http://openintrostat.github.io/openintro/reference/bdims.html).]

    ```{r}
    library(tidyverse)
    library(openintro)
    
    ggplot(bdims, aes(x = sho_gi, y = hgt)) +
      geom_point(color = COL["blue","full"], size = 2, alpha = 0.8) +
      theme_minimal() +
      labs(x = "Shoulder girth (cm)", y = "Height (cm)")
    ```

    a.  Describe the relationship between shoulder girth and height.

    b.  How would the relationship change if shoulder girth was measured in inches while the units of height remained in centimeters?

1.  **Guess the correlation.** 
Eduardo and Rosie are both collecting data on number of rainy days in a year and the total rainfall for the year. Eduardo records rainfall in inches and Rosie in centimeters. How will their correlation coefficients compare?

1.  **The Coast Starlight, correlation.** 
The Coast Starlight Amtrak train runs from Seattle to Los Angeles. 
The scatterplot below displays the distance between each stop (in miles) and the amount of time it takes to travel from one stop to another (in minutes).^[The data used in this exercise can be found in the **openintro** package: [`coast_starlight`](http://openintrostat.github.io/openintro/reference/coast_starlight.html).]

    ```{r}
    ggplot(coast_starlight, aes(x = dist, y = travel_time)) +
      geom_point(color = COL["blue","full"], size = 2) +
      theme_minimal() +
      labs(x = "Distance (miles)", y = "Travel Time (minutes)")
    ```

    a.  Describe the relationship between distance and travel time.

    b.  How would the relationship change if travel time was instead measured in hours, and distance was instead measured in kilometers?

    c.  Correlation between travel time (in miles) and distance (in minutes) is $r = 0.636$. What is the correlation between travel time (in kilometers) and distance (in hours)?

1.  **Crawling babies, correlation.** 
A study conducted at the University of Denver investigated whether babies take longer to learn to crawl in cold months, when they are often bundled in clothes that restrict their movement, than in warmer months. 
Infants born during the study year were split into twelve groups, one for each birth month. 
We consider the average crawling age of babies in each group against the average temperature when the babies are six months old (that's when babies often begin trying to crawl). 
Temperature is measured in degrees Fahrenheit (F) and age is measured in weeks.^[The data used in this exercise can be found in the **openintro** package: [`babies_crawl`](http://openintrostat.github.io/openintro/reference/babies_crawl.html).]

    ```{r}
    library(tidyverse)
    library(openintro)
    
    ggplot(babies_crawl, aes(x = temperature, y = avg_crawling_age)) +
      geom_point(color = COL["blue","full"], size = 2) +
      theme_minimal() +
      labs(x = "Temperature (F)", y = "Average crawling age (weeks)")
    ```

    a.  Describe the relationship between temperature and crawling age.

    b.  How would the relationship change if temperature was measured in degrees Celsius (C) and age was measured in months?

    c.  The correlation between temperature in F and age in weeks was $r=-0.70$. If we converted the temperature to C and age to months, what would the correlation be?

1.  **Partners' ages.** 
What would be the correlation between the ages of partners if people always dated others who are 

    1.  3 years younger than themselves?

    2.  2 years older than themselves?

    3.  half as old as themselves?

1.  **Graduate degrees and salaries.** 
What would be the correlation between the annual salaries of people with and without a graduate degree at a company if for a certain type of position someone with a graduate degree always made 

    a.  \$5,000 more than those without a graduate degree?

    b.  25% more than those without a graduate degree?

    c.  15% less than those without a graduate degree?
