1.  **Outliers, Part I.** 
Identify the outliers in the scatterplots shown below, and determine what type of outliers they are. Explain your reasoning.

    ```{r fig.asp=0.33}
    library(openintro)
    library(tidyverse)
    library(patchwork)
    
    set.seed(83629)
    
    df <- tibble(
      x = seq(1,50,1),
      y = -2 * x + 20 + rnorm(length(x), mean = 0, sd = 10),
      x_influential = c(x[1:49], 200),
      y_leverage = c(y[1:49], -370),
      y_outlier = c(y[1:25], y[26]+100, y[27:50])
    )
    
    p_1 <- ggplot(df, aes(x = x_influential, y = y)) +
      geom_smooth(method = "lm", color = "darkgray", se = FALSE) +
      geom_point(col = COL["blue", "full"]) +
      theme_minimal() +
      labs(x = NULL, y = NULL) +
      theme(axis.text = element_blank())
    
    p_2 <- ggplot(df, aes(x = x_influential, y = y_leverage)) +
      geom_smooth(method = "lm", color = "darkgray", se = FALSE) +
      geom_point(col = COL["blue", "full"]) +
      theme_minimal() +
      labs(x = NULL, y = NULL) +
      theme(axis.text = element_blank())
    
    p_3 <- ggplot(df, aes(x = x, y = y_outlier)) +
      geom_smooth(method = "lm", color = "darkgray", se = FALSE) +
      geom_point(col = COL["blue", "full"]) +
      theme_minimal() +
      labs(x = NULL, y = NULL) +
      theme(axis.text = element_blank())
    
    p_1 + p_2 + p_3
    ```

1.  **Outliers, Part II.** 
Identify the outliers in the scatterplots shown below and determine what type of outliers they are. Explain your reasoning.

    ```{r fig.asp=0.33}
    library(openintro)
    library(tidyverse)
    library(patchwork)
    
    set.seed(83629)
    
    df <- tibble(
      x = seq(1,50,1),
      y = 3 * x + 3 + rnorm(length(x), mean = 0, sd = 10),
      x_influential = c(x[1:49], -50),
      y_influential = c(y[1:49], -300),
      y_outlier = c(y[1:25], y[26]+100, y[27:50])
    )
    
    p_1 <- ggplot(df, aes(x = x_influential, y = y)) +
      geom_smooth(method = "lm", color = "darkgray", se = FALSE) +
      geom_point(col = COL["blue", "full"]) +
      theme_minimal() +
      labs(x = NULL, y = NULL) +
      theme(axis.text = element_blank())
    
    p_2 <- ggplot(df, aes(x = x_influential, y = y_influential)) +
      geom_smooth(method = "lm", color = "darkgray", se = FALSE) +
      geom_point(col = COL["blue", "full"]) +
      theme_minimal() +
      labs(x = NULL, y = NULL) +
      theme(axis.text = element_blank())
    
    p_3 <- ggplot(df, aes(x = x, y = y_outlier)) +
      geom_smooth(method = "lm", color = "darkgray", se = FALSE) +
      geom_point(col = COL["blue", "full"]) +
      theme_minimal() +
      labs(x = NULL, y = NULL) +
      theme(axis.text = element_blank())
    
    p_1 + p_2 + p_3
    ```

1.  **Urban homeowners, outliers.** 
The scatterplot below shows the percent of families who own their home vs. the percent of the population living in urban areas. 
There are 52 observations, each corresponding to a state in the US. Puerto Rico and District of Columbia are also included.^[The data used in this exercise can be found in the **openintro** package: [`urban_owner`](http://openintrostat.github.io/openintro/reference/urban_owner.html).]

    ```{r}
    library(tidyverse)
    library(openintro)
    library(scales)
    
    ggplot(urban_owner, aes(x = poppct_urban, y = pct_owner_occupied)) +
      geom_point(color = COL["blue", "full"]) +
      theme_minimal() +
      scale_x_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
      scale_y_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
      labs(
        x = "Percent urban population",
        y = "Percent who own their home"
      )
    ```

    a.  Describe the relationship between the percent of families who own their home and the percent of the population living in urban areas.

    b.  The outlier at the bottom right corner is District of Columbia, where 100% of the population is considered urban. What type of an outlier is this observation?

1.  **Crawling babies, outliers.**
A study conducted at the University of Denver investigated whether babies take longer to learn to crawl in cold months, when they are often bundled in clothes that restrict their movement, than in warmer months. 
The plot below shows the relationship between average crawling age of babies born in each month and the average temperature in the month when the babies are six months old.
The plot reveals a potential outlying month when the average temperature is about 53F and average crawling age is about 28.5 weeks. 
Does this point have high leverage? Is it an influential point?^[The data used in this exercise can be found in the **openintro** package: [`babies_crawl`](http://openintrostat.github.io/openintro/reference/babies_crawl.html).]

    ```{r}
    library(tidyverse)
    library(openintro)
    
    ggplot(babies_crawl, aes(x = temperature, y = avg_crawling_age)) +
      geom_point(color = COL["blue","full"], size = 2) +
      theme_minimal() +
      labs(x = "Temperature (F)", y = "Average crawling age (weeks)")
    ```
