1.  **Units of regression.** 
Consider a regression predicting the number of calories (cal) from width (cm) for a sample of square shaped chocolate brownies. What are the units of the correlation coefficient, the intercept, and the slope?

1.  **Which is higher?**
Determine if (I) or (II) is higher or if they are equal: *"For a regression line, the uncertainty associated with the slope estimate, $b_1$, is higher when  (I)  there is a lot of scatter around the regression line or (II)  there is very little scatter around the regression line."* Explain your reasoning.

1.  **Over-under, Part I.** 
Suppose we fit a regression line to predict the shelf life of an apple based on its weight. For a particular apple, we predict the shelf life to be 4.6 days. The apple's residual is -0.6 days. Did we over or under estimate the shelf-life of the apple? Explain your reasoning.

1.  **Over-under, Part II.** 
Suppose we fit a regression line to predict the number of incidents of skin cancer per 1,000 people from the number of sunny days in a year. 
For a particular year, we predict the incidence of skin cancer to be 1.5 per 1,000 people, and the residual for this year is 0.5. 
Did we over or under estimate the incidence of skin cancer? Explain your reasoning.

1.  **Tourism spending.** 
The Association of Turkish Travel Agencies reports the number of foreign tourists visiting Turkey and tourist spending by year. 
Three plots are provided: scatterplot showing the relationship between these two variables along with the least squares fit, residuals plot, and histogram of residuals.^[The data used in this exercise can be found in the **openintro** package: [`tourism`](http://openintrostat.github.io/openintro/reference/tourism.html).]

    ```{r}
    library(openintro)
    library(ggplot2)
    library(broom)
    library(patchwork)
    
    p_1 <- ggplot(tourism, aes(x = visitor_count_tho, y = tourist_spending)) +
      geom_smooth(method = "lm", color = "darkgray", se = FALSE) +
      geom_point(col = COL["blue","full"], size = 3, alpha = 0.6) +
      labs(
        x = "Number of tourists (thousands)",
        y = "Spending (million $)"
        ) +
      theme_minimal()

    m <- lm(tourist_spending ~ visitor_count_tho, data = tourism)
    m_aug <- augment(m)
    
    p_2 <- ggplot(m_aug, aes(x = visitor_count_tho, y = .resid)) +
      geom_hline(yintercept = 0, linetype = "dashed") +
      geom_point(col = COL["blue","full"], size = 3, alpha = 0.6) +
      labs(
        x = "Number of tourists (thousands)",
        y = "Residuals"
        ) +
      theme_minimal()
    
    p_3 <- ggplot(m_aug, aes(x = .resid)) +
      geom_histogram(binwidth = 250, fill = COL["blue", "full"], color = "#4287aa") +
      labs(
        x = "Residuals",
        y = NULL
        ) +
      theme_minimal()
    
    p_1 + p_2 + p_3
    ```

    a.  Describe the relationship between number of tourists and spending.

    b.  What are the explanatory and response variables?

    c.  Why might we want to fit a regression line to these data?

    d.  Do the data meet the conditions required for fitting a least squares line? In addition to the scatterplot, use the residual plot and histogram to answer this question.

1.  **Starbucks, calories and carbs.** 
The scatterplot below shows the relationship between the number of calories and amount of carbohydrates (in grams) Starbucks food menu items contain. Since Starbucks only lists the number of calories on the display items, we are interested in predicting the amount of carbs a menu item has based on its calorie content.^[The data used in this exercise can be found in the **openintro** package: [`starbucks`](http://openintrostat.github.io/openintro/reference/starbucks.html).]

    ```{r}
    library(openintro)
    library(ggplot2)
    library(broom)
    library(patchwork)
    
    p_1 <- ggplot(starbucks, aes(x = carb, y = calories)) +
      geom_smooth(method = "lm", color = "darkgray", se = FALSE) +
      geom_point(col = COL["blue","full"], size = 3) +
      labs(
        x = "Calories",
        y = "Carbs (grams)"
        ) +
      theme_minimal()

    m <- lm(carb ~ calories, data = starbucks)
    m_aug <- augment(m)
    
    p_2 <- ggplot(m_aug, aes(x = calories, y = .resid)) +
      geom_hline(yintercept = 0, linetype = "dashed") +
      geom_point(col = COL["blue","full"], size = 3) +
      labs(
        x = "Calories",
        y = "Residuals"
        ) +
      theme_minimal()
    
    p_3 <- ggplot(m_aug, aes(x = .resid)) +
      geom_histogram(binwidth = 5, fill = COL["blue", "full"], color = "#4287aa") +
      labs(
        x = "Residuals",
        y = NULL
        ) +
      theme_minimal()
    
    p_1 + p_2 + p_3
    ```

    a.  Describe the relationship between number of calories and amount of carbohydrates (in grams) that Starbucks food menu items contain.

    b.  In this scenario, what are the explanatory and response variables?

    c.  Why might we want to fit a regression line to these data?

    d.  Do these data meet the conditions required for fitting a least squares line?

1.  **The Coast Starlight, regression.** 
The Coast Starlight Amtrak train runs from Seattle to Los Angeles. 
The scatterplot below displays the distance between each stop (in miles) and the amount of time it takes to travel from one stop to another (in minutes).
The mean travel time from one stop to the next on the Coast Starlight is 129 mins, with a standard deviation of 113 minutes. The mean distance traveled from one stop to the next is 108 miles with a standard deviation of 99 miles. 
The correlation between travel time and distance is 0.636.^[The data used in this exercise can be found in the **openintro** package: [`coast_starlight`](http://openintrostat.github.io/openintro/reference/coast_starlight.html).]

    ```{r}
    ggplot(coast_starlight, aes(x = dist, y = travel_time)) +
      geom_point(color = COL["blue","full"], size = 2) +
      theme_minimal() +
      labs(x = "Distance (miles)", y = "Travel Time (minutes)")
    ```

    a.  Write the equation of the regression line for predicting travel time.

    b.  Interpret the slope and the intercept in this context.

    c.  Calculate $R^2$ of the regression line for predicting travel time from distance traveled for the Coast Starlight, and interpret $R^2$ in the context of the application.

    d.  The distance between Santa Barbara and Los Angeles is 103 miles. Use the model to estimate the time it takes for the Starlight to travel between these two cities.

    e.  It actually takes the Coast Starlight about 168 mins to travel from Santa Barbara to Los Angeles. Calculate the residual and explain the meaning of this residual value.

    f.  Suppose Amtrak is considering adding a stop to the Coast Starlight 500 miles away from Los Angeles. Would it be appropriate to use this linear model to predict the travel time from Los Angeles to this point?

1.  **Body measurements, regression.** 
Researchers studying anthropometry collected body and skeletal diameter measurements, as well as age, weight, height and sex for 507 physically active individuals. 
The scatterplot below shows the relationship between height and shoulder girth (circumference of shoulders measured over deltoid muscles), both measured in centimeters.
The mean shoulder girth is 107.20 cm with a standard deviation of 10.37 cm. 
The mean height is 171.14 cm with a standard deviation of 9.41 cm. 
The correlation between height and shoulder girth is 0.67.^[The data used in this exercise can be found in the **openintro** package: [`bdims`](http://openintrostat.github.io/openintro/reference/bdims.html).]

    ```{r}
    library(tidyverse)
    library(openintro)
    
    ggplot(bdims, aes(x = sho_gi, y = hgt)) +
      geom_point(color = COL["blue","full"], size = 2, alpha = 0.8) +
      theme_minimal() +
      labs(x = "Shoulder girth (cm)", y = "Height (cm)")
    ```

    a.  Write the equation of the regression line for predicting height.

    b.  Interpret the slope and the intercept in this context.

    c.  Calculate $R^2$ of the regression line for predicting height from shoulder girth, and interpret it in the context of the application.

    d.  A randomly selected student from your class has a shoulder girth of 100 cm. Predict the height of this student using the model.

    e.  The student from part (d) is 160 cm tall. Calculate the residual, and explain what this residual means.

    f.  A one year old has a shoulder girth of 56 cm. Would it be appropriate to use this linear model to predict the height of this child?

1.  **Murders and poverty, regression.** 
The following regression output is for predicting annual murders per million (`annual_murders_per_mil`) from percentage living in poverty (`perc_pov`) in a random sample of 20 metropolitan areas.^[The data used in this exercise can be found in the **openintro** package: [`murders`](http://openintrostat.github.io/openintro/reference/murders.html).]

    ```{r}
    library(openintro)
    library(broom)
    library(knitr)
    library(ggplot2)
    library(scales)
    
    ggplot(murders, aes(x = perc_pov, y = annual_murders_per_mil)) +
      geom_point(col = COL["blue","full"], size = 3) +
      labs(
        x = "Percent in poverty",
        y = "Number of murders per year per million people"
        ) +
      theme_minimal() +
      scale_x_continuous(labels = label_percent(scale = 1, accuracy = 1))

    m <- lm(annual_murders_per_mil ~ perc_pov, data = murders)
    tidy(m) %>%
      kable(digits = 3)
    
    glance(m) %>%
      transmute(
        s = round(sigma, 3),
        `R-squared` = paste0(round(r.squared * 100, 2), "%"),
        `Adjusted R-squared` = paste0(round(adj.r.squared * 100, 2), "%")
      ) %>%
      kable(align = "rrr")
    ```

    a.  Write out the linear model.

    b.  Interpret the intercept.

    c.  Interpret the slope.

    d.  Interpret $R^2$.

    e.  Calculate the correlation coefficient.

1.  **Cats, regression.** 
The following regression output is for predicting the heart weight (`Hwt`, in g) of cats from their body weight (`Bwt`, in kg). The coefficients are estimated using a dataset of 144 domestic cats.^[The data used in this exercise can be found in the **MASS** package: [`cats`](https://cran.r-project.org/web/packages/MASS/MASS.pdf).]

    ```{r}
    library(openintro)
    library(MASS)
    library(broom)
    library(knitr)
    library(ggplot2)
    
    ggplot(cats, aes(x = Bwt, y = Hwt)) +
      geom_point(col = COL["blue","full"], size = 3) +
      labs(
        x = "Body weight (kg)",
        y = "Heart weight (g)"
        ) +
      theme_minimal()

    m <- lm(Hwt ~ Bwt, data = cats)
    tidy(m) %>%
      kable(digits = 3)
    
    glance(m) %>%
      transmute(
        s = round(sigma, 3),
        `R-squared` = paste0(round(r.squared * 100, 2), "%"),
        `Adjusted R-squared` = paste0(round(adj.r.squared * 100, 2), "%")
      ) %>%
      kable(align = "rrr")
    ```

    a.  Write out the linear model.

    b.  Interpret the intercept.

    c.  Interpret the slope.

    d.  Interpret $R^2$.

    e.  Calculate the correlation coefficient.
