1. **Identify the parameter, I**
For each of the following situations, state whether the parameter of interest is a mean or a proportion. It may be helpful to examine whether individual responses are numerical or categorical.

    a. In a survey, one hundred college students are asked how many hours per week they spend on the Internet.

    b. In a survey, one hundred college students are asked: "What percentage of the time you spend on the Internet is part of your course work?"

    c. In a survey, one hundred college students are asked whether or not they cited information from Wikipedia in their papers.

    d. In a survey, one hundred college students are asked what percentage of their total weekly spending is on alcoholic beverages.

    e. In a sample of one hundred recent college graduates, it is found that 85 percent expect to get a job within one year of their graduation date.

1. **Identify the parameter, II.**
For each of the following situations, state whether the parameter of interest is a mean or a proportion.

    a. A poll shows that 64% of Americans personally worry a great deal about federal spending and the budget deficit.

    b. A survey reports that local TV news has shown a 17% increase in revenue within a two year period while newspaper revenues decreased by 6.4% during this time period.

    c. In a survey, high school and college students are asked whether or not they use geolocation services on their smart phones.

    d. In a survey, smart phone users are asked whether or not they use a web-based taxi service.

    e. In a survey, smart phone users are asked how many times they used a web-based taxi service over the last year.

1. **Hypotheses.**
For each of the research statements below, note whether it represents a null hypothesis claim or an alternative hypothesis claim.

    a. The number of hours that grade-school children spend doing homework predicts their future success on standardized tests.
    
    b. King cheetahs on average run the same speed as standard spotted cheetahs.
    
    c. For a particular student, the probability of correctly answering a 5-option multiple choice test is larger than 0.2 (i.e., better than guessing).
    
    d. The mean length of African elephant tusks has changed over the last 100 years.
    
    e. The risk of facial clefts is equal for babies born to mothers who take folic acid supplements compared with those from mothers who do not.
    
    f. Caffeine intake during pregnancy affects mean birth weight.
    
    g. The probability of getting in a car accident is the same if using a cell phone than if not using a cell phone.

1. **True null hypothesis.**
Unbeknownst to you, let's say that the null hypothesis is actually true in the population. You plan to run a study anyway.

    a. If the level of significance you choose (i.e., the cutoff for your p-value) is 0.05, how likely is it that you will mistakenly reject the null hypothesis?
    
    b. If the level of significance you choose (i.e., the cutoff for your p-value) is 0.01, how likely is it that you will mistakenly reject the null hypothesis?
        
    c. If the level of significance you choose (i.e., the cutoff for your p-value) is 0.10, how likely is it that you will mistakenly reject the null hypothesis?

1. **Identify hypotheses, I.**
Write the null and alternative hypotheses in words and then symbols for each of the following situations.

    a.  New York is known as "the city that never sleeps". A random sample of 25 New Yorkers were asked how much sleep they get per night. Do these data provide convincing evidence that New Yorkers on average sleep less than 8 hours a night?

    b.  Employers at a firm are worried about the effect of March Madness, a basketball championship held each spring in the US, on employee productivity. They estimate that on a regular business day employees spend on average 15 minutes of company time checking personal email, making personal phone calls, etc. They also collect data on how much company time employees spend on such non- business activities during March Madness. They want to determine if these data provide convincing evidence that employee productivity decreases during March Madness.

1. **Identify hypotheses, II.**
Write the null and alternative hypotheses in words and using symbols for each of the following situations.

    a.  Since 2008, chain restaurants in California have been required to display calorie counts of each menu item. Prior to menus displaying calorie counts, the average calorie intake of diners at a restaurant was 1100 calories. After calorie counts started to be displayed on menus, a nutritionist collected data on the number of calories consumed at this restaurant from a random sample of diners. Do these data provide convincing evidence of a difference in the average calorie intake of a diners at this restaurant?

    b.  Based on the performance of those who took the GRE exam between July 1, 2004 and June 30, 2007, the average Verbal Reasoning score was calculated to be 462. In 2021 the average verbal score was slightly higher. Do these data provide convincing evidence that the average GRE Verbal Reasoning score has changed since 2021?

1. **Side effects of Avandia.** 
Rosiglitazone is the active ingredient in the controversial type 2 diabetes medicine Avandia and has been linked to an increased risk of serious cardiovascular problems such as stroke, heart failure, and death. A common alternative treatment is Pioglitazone, the active ingredient in a diabetes medicine called Actos. In a nationwide retrospective observational study of 227,571 Medicare beneficiaries aged 65 years or older, it was found that 2,593 of the 67,593 patients using Rosiglitazone and 5,386 of the 159,978 using Pioglitazone had serious cardiovascular problems. These data are summarized in the contingency table below.^[The [`avandia`](http://openintrostat.github.io/openintro/reference/avandia.html) data used in this exercise can be found in the **openintro** R package.] [@Graham:2010]

    ```{r}
    library(tools)
    library(openintro)
    library(tidyverse)
    library(janitor)
    library(knitr)
    library(kableExtra)
    
    avandia %>%
      count(treatment, cardiovascular_problems) %>%
      mutate(cardiovascular_problems = toTitleCase(as.character(cardiovascular_problems))) %>%
      pivot_wider(names_from = cardiovascular_problems, values_from = n) %>%
      adorn_totals(where = c("row", "col")) %>%
      kbl(linesep = "\\addlinespace", booktabs = TRUE, format.args = list(big.mark = ",")) %>%
      kable_styling(bootstrap_options = c("striped", "condensed"), 
                    latex_options = c("striped", "HOLD_position"),
                    full_width = FALSE) %>%
      column_spec(1:4, width = "7em")
    ```

    a.  Determine if each of the following statements is true or false. If false, explain why. *Be careful:* The reasoning may be wrong even if the statement's conclusion is correct. In such cases, the statement should be considered false.
    
        i. Since more patients on Pioglitazone had cardiovascular problems (5,386 vs. 2,593), we can conclude that the rate of cardiovascular problems for those on a Pioglitazone treatment is higher.
        
        ii. The data suggest that diabetic patients who are taking Rosiglitazone are more likely to have cardiovascular problems since the rate of incidence was (2,593 / 67,593 = 0.038) 3.8\% for patients on this treatment, while it was only (5,386 / 159,978 = 0.034) 3.4\% for patients on Pioglitazone. 
        
        iii. The fact that the rate of incidence is higher for the Rosiglitazone group proves that Rosiglitazone causes serious cardiovascular problems. 
        
        iv. Based on the information provided so far, we cannot tell if the difference between the rates of incidences is due to a relationship between the two variables or due to chance.
        
    b.  What proportion of all patients had cardiovascular problems?

    c.  If the type of treatment and having cardiovascular problems were independent, about how many patients in the Rosiglitazone group would we expect to have had cardiovascular problems?

    d.  We can investigate the relationship between outcome and treatment in this study using a randomization technique. While in reality we would carry out the simulations required for randomization using statistical software, suppose we actually simulate using index cards. In order to simulate from the independence model, which states that the outcomes were independent of the treatment, we write whether or not each patient had a cardiovascular problem on cards, shuffled all the cards together, then deal them into two groups of size 67,593 and 159,978. We repeat this simulation 100 times and each time record the difference between the proportions of cards that say "Yes" in the Rosiglitazone and Pioglitazone groups. Use the histogram of these differences in proportions to answer the following questions.
    
        i. What are the claims being tested? 
        
        ii. Compared to the number calculated in part (b), which would provide more support for the alternative hypothesis, *higher* or *lower* proportion of patients with cardiovascular problems in the Rosiglitazone group? 
        
        iii. What do the simulation results suggest about the relationship between taking Rosiglitazone and having cardiovascular problems in diabetic patients?
    
    ```{r}
    library(infer)
    set.seed(25)
    avandia %>%
      specify(response = cardiovascular_problems, explanatory = treatment, success = "yes") %>%
      hypothesize(null = "independence") %>%
      generate(reps = 100, type = "permute") %>%
      calculate(stat = "diff in props", order = c("Rosiglitazone", "Pioglitazone")) %>%
      ggplot(aes(x = stat)) +
      geom_histogram(binwidth = 0.001/4,  fill = IMSCOL["green", "full"]) +
      labs(y = NULL, x = "Difference in cardiovascular problem rates (Rosiglitazone - Pioglitazone)") +
      scale_y_continuous(breaks = seq(0, 16, 2))
    ```

1.  **Heart transplants.** 
The Stanford University Heart Transplant Study was conducted to determine whether an experimental heart transplant program increased lifespan. Each patient entering the program was designated an official heart transplant candidate, meaning that they were gravely ill and would most likely benefit from a new heart. Some patients got a transplant and some did not. The variable `transplant` indicates which group the patients were in; patients in the treatment group got a transplant and those in the control group did not. Of the 34 patients in the control group, 30 died. Of the 69 people in the treatment group, 45 died. Another variable called `survived` was used to indicate whether or not the patient was alive at the end of the study.^[The [`heart_transplant`](http://openintrostat.github.io/openintro/reference/heart_transplant.html) data used in this exercise can be found in the **openintro** R package.] [@Turnbull+Brown+Hu:1974]

    ```{r}
    library(tidyverse)
    library(openintro)
    library(patchwork)
    
    heart_transplant <- heart_transplant %>%
      mutate(survived_better_wording = if_else(survived == "dead", "deceased", as.character(survived)))
    
    p_bar <- ggplot(heart_transplant, aes(x = transplant, fill = survived_better_wording)) +
      geom_bar(position = "fill") +
      scale_fill_openintro("two") +
      labs(x = NULL, y = NULL, fill = "Outcome")
    
    p_box <- ggplot(heart_transplant, aes(x = transplant, y = survtime)) +
      geom_boxplot() +
      labs(x = NULL, y = "Survival time (days)")
    
    p_bar + p_box
    ```

    a.  Does the stacked bar plot indicate that survival is independent of whether or not the patient got a transplant? Explain your reasoning.

    b.  What do the box plots above suggest about the efficacy (effectiveness) of the heart transplant treatment.

    c.  What proportion of patients in the treatment group and what proportion of patients in the control group died?

    d.  One approach for investigating whether or not the treatment is effective is to use a randomization technique.
    
        i. What are the claims being tested?
        
        ii. The paragraph below describes the set up for such approach, if we were  to do it without using statistical software. Fill in the blanks with a number  or phrase, whichever is appropriate.
        
        > We write *alive* on $\rule{2cm}{0.5pt}$ cards representing patients who were  alive at the end of the study, and *deceased* on $\rule{2cm}{0.5pt}$ cards  representing patients who were not. Then, we shuffle these cards and split them into two groups: one group of size $\rule{2cm}{0.5pt}$ representing treatment, and another group of size $\rule{2cm}{0.5pt}$ representing control. We calculate the difference between the proportion of \textit{deceased} cards in the treatment and control groups (treatment - control) and record this value. We repeat this 100 times to build a distribution centered at $\rule{2cm}{0.5pt}$. Lastly, we calculate the fraction of simulations where the simulated differences in proportions are $\rule{2cm}{0.5pt}$. If this fraction is low, we conclude that it is unlikely to have observed such an outcome by chance and that the null hypothesis should be rejected in favor of the alternative.

         iii. What do the simulation results shown below suggest about the effectiveness  of the transplant program?
         
    ```{r}
    library(infer)
    set.seed(40)
    heart_transplant %>%
      specify(response = survived_better_wording, explanatory = transplant, success = "deceased") %>%
      hypothesize(null = "independence") %>%
      generate(reps = 100, type = "permute") %>%
      calculate(stat = "diff in props", order = c("treatment", "control")) %>%
      ggplot(aes(x = stat)) +
      geom_histogram(binwidth = 0.05, fill = IMSCOL["green", "full"]) +
      labs(y = NULL, x = "Difference in proportions of deceased (treatment - control)") +
      scale_y_continuous(breaks = seq(0, 24, 4), minor_breaks = seq(0, 24, 2))
    ```
