1. **Cats like enclosed spaces, Part I.**
A citizen science project on which type of enclosed spaces cats are most likely to sit in compared (among other options) two different spaces taped to the ground.  The first was a square, and the second was a shape known as Kanizsa square illusion.  When comparing the two options given to 7 cats, 5 chose the square, and 2 chose the Kanizsa.  For this question, we are interested to know whether there is evidence that the cats have a preference for one of the shapes, or not. [@Smith:2021]

    ```{r}
    library(tidyverse)
    library(infer)
    
    cats <- data.frame(choice = c(rep("square", 5), rep("Kanizsa", 2)))
    
    set.seed(47)
    cats %>%
      specify(response = choice, success = "square") %>%
      hypothesize(null = "point", p = 0.5) %>%
      generate(reps = 1000, method = "simulate") %>%
      calculate(stat = "prop") %>%
      ggplot() +
      geom_histogram(aes(x = stat)) + 
      xlab("Bootstrapped sample proportions of square, p = 0.5")
    ```

    a. Calculate the standard error of the sample proportion using the mathematical model.
    
    b. A parametric bootstrap simulation (1000 reps) was run and is displayed in the histogram above.  What is the approximate standard error of the sample proportion given the parametric bootstrap?
    
    c. Do the mathematical model and parametric bootstrap give similar standard errors?
    
    d. In order to approach the problem using the mathematical model, are the success-failure conditions met for this study?  Explain.
    
    e. What is evident in the parametric bootstrap distribution which indicates that the mathematical model should probably not be used?

1. **Cats like enclosed spaces, Part II.**
A citizen science project on which type of enclosed spaces cats are most likely to sit in compared (among other options) two different spaces taped to the ground.  The first was a square, and the second was a shape known as Kanizsa square illusion.  When comparing the two options given to 7 cats, 5 chose the square, and 2 chose the Kanizsa.  For this question, we are interested to know whether there is evidence that the cats have a preference for one of the shapes, or not. [@Smith:2021]

    ```{r}
    library(tidyverse)
    library(infer)
    
    cats <- data.frame(choice = c(rep("square", 5), rep("Kanizsa", 2)))
    
    set.seed(47)
    cats %>%
      specify(response = choice, success = "square") %>%
      hypothesize(null = "point", p = 0.5) %>%
      generate(reps = 1000, method = "simulate") %>%
      calculate(stat = "prop") %>%
      ggplot() +
      geom_histogram(aes(x = stat)) + 
      xlab("Bootstrapped sample proportions of square, p = 0.5")
    ```
    
    
    a. What are the null and alternative hypotheses for the question on whether cats have a preference over the two spaces?
    
    b. Using the parametric bootstrap sampling distribution above (on 1000 reps), find the p-value and conclude the hypothesis test in the context of the problem.

1. **Legalization of marijuana, SE Part I.**
The General Social Survey asked a random sample of 1,563 US adults: "Do you think the use of marijuana should be made legal, or not?" 60% of the respondents said it should be made legal. [@data:gssgrass]  Consider a scenario where, in order to become legal, 55% (or more) of voters must approve.

    A parametric bootstrap sampling distribution is given below with 100 repetitions to show the variability of the sample proportion when 55% of voters approve legalizing marijuana.

    ```{r}
    library(tidyverse)
    library(infer)
    
    mj <- data.frame(opinion = c(rep("legal", 938),rep("not legal", 625)))
    
    set.seed(47)
    mj %>%
      specify(response = opinion, success = "legal") %>%
      hypothesize(null = "point", p = 0.55) %>%
      generate(reps = 1000, type = "simulate") %>%
      calculate(stat = "prop") %>%
      ggplot() + 
      geom_histogram(aes(x = stat)) +
      xlab("Bootstrapped sample proportions of legal support, p = 0.55")
    
    ```
    
    a. Calculate the standard error of the sample proportion using the mathematical model.
    
    b. Use the parametric bootstrap simulation to approximate the standard error of the sample proportion.
    
    c. In this setting (to test whether the true underlying population proportion is greater than 0.55), would there be a strong reason to choose the mathematical model over the parametric bootstrap (or vice versa)?

1. **Legalization of marijuana, SE Part II.**
The General Social Survey asked a random sample of 1,563 US adults: "Do you think the use of marijuana should be made legal, or not?" 60% of the respondents said it should be made legal. [@data:gssgrass] Consider a scenario where, in order to become legal, 55% (or more) of voters must approve.

    A bootstrap sampling distribution is given below with 100 repetitions to show the variability of the sample proportion using the data provided from the General Social Survey.

    ```{r}
    library(tidyverse)
    library(infer)
    
    mj <- data.frame(opinion = c(rep("legal", 938),rep("not legal", 625)))
    
    set.seed(47)
    mj %>%
      specify(response = opinion, success = "legal") %>%
      generate(reps = 1000, type = "bootstrap") %>%
      calculate(stat = "prop") %>%
      ggplot() + 
      geom_histogram(aes(x = stat)) +
      xlab("Bootstrapped sample proportions of legal support")
    
    ```
    
    a. Calculate the standard error of the sample proportion using the mathematical model.
    
    b. Use the bootstrap simulation to approximate the standard error of the sample proportion.
    
    c. In this setting (to test whether the true underlying population proportion is greater than 0.55), would there be a strong reason to choose the mathematical model over the bootstrap taken repeatedly on the dataset (or vice versa)?
    

1. **Statistics and employment, Part I.**
A large university knows that about 70% of the full-time students are employed at least 5 hours per week.  The members of the Statistics Department wonder if the same proportion of their students work at least 5 hours per week.  They sample 25 majors and find that 15 of the students work 5 or more hours each week.

    Two bootstrap sampling distributions are created to describe the variability in the proportion of statistics majors who work at least 5 hours per week.  The parametric bootstrap imposes a true population proportion of $p=0.7$ while the data bootstrap resamples from the actual data (which has 60% of the observations who work at least 5 hours per week).

    ```{r}
    library(tidyverse)
    library(infer)
    
    students <- data.frame(outside = c(rep("work", 15), rep("don't work", 10)))
    
    set.seed(47)
    students %>%
      specify(response = outside, success = "work") %>%
      hypothesize(null = "point", p = 0.7) %>%
      generate(reps = 1000, type = "simulate") %>%
      calculate(stat = "prop") %>%
      ggplot() + 
      geom_histogram(aes(x = stat), binwidth = 0.02) +
      xlab("Parametric bootstrap, proportion who work, p = 0.7")

    set.seed(47)
    students %>%
      specify(response = outside, success = "work") %>%
      generate(reps = 1000, type = "bootstrap") %>%
      calculate(stat = "prop") %>%
      ggplot() + 
      geom_histogram(aes(x = stat), binwidth = 0.02) +
      xlab("Data bootstrap, proportion who work")
    ```
    
    a. Where are the two distributions centered?  Are they centered at roughly the same place?
    
    b. Do the two distributions indicate that the standard error of the sample proportion is roughly equal?  Provide the standard error values given by each histogram.
    
    c. Is the shape of the two histograms roughly the same?  Describe their shapes.
    
    d. The bootstrap sampling was done under two different settings.  Describe the two different settings.

1. **Statistics and employment, Part II.**
A large university knows that about 70% of the full-time students are employed at least 5 hours per week.  The members of the Statistics Department wonder if the same proportion of their students work at least 5 hours per week.  They sample 25 majors and find that 15 of the students work 5 or more hours each week.

    Two bootstrap sampling distributions are created to describe the variability in the proportion of statistics majors who work at least 5 hours per week.  The parametric bootstrap imposes a true population proportion of $p=0.7$ while the data bootstrap resamples from the actual data (which has 60% of the observations who work at least 5 hours per week).

    ```{r}
    library(tidyverse)
    library(infer)
    
    students <- data.frame(outside = c(rep("work", 15), rep("don't work", 10)))
    
    set.seed(47)
    students %>%
      specify(response = outside, success = "work") %>%
      hypothesize(null = "point", p = 0.7) %>%
      generate(reps = 1000, type = "simulate") %>%
      calculate(stat = "prop") %>%
      ggplot() + 
      geom_histogram(aes(x = stat), binwidth = 0.02) +
      xlab("Parametric bootstrap, proportion who work, p = 0.7")

    set.seed(47)
    students %>%
      specify(response = outside, success = "work") %>%
      generate(reps = 1000, type = "bootstrap") %>%
      calculate(stat = "prop") %>%
      ggplot() + 
      geom_histogram(aes(x = stat), binwidth = 0.02) +
      xlab("Data bootstrap, proportion who work")
    ```
    
    a. Which bootstrap distribution should be used to test whether the proportion of all statistics majors who work is 70%?  And which bootstrap distribution should be used to find a confidence interval for the true poportion of statistics majors who work at least 5 hours per week?
    
    b. Using the appropriate histogram, test the claim that 70% of statistics majors, like their peers, work at least 5 hours per week.  State the null and alternative hypotheses, find the p-value, and conclude the test in the context of the problem.
    
    c. Using the appropriate histogram, find a 98% bootstrap percentile confidence interval for the true proportion of statistics majors who work at least 5 hours per week.  Interpret the confidence interval in the context of the problem.
    
    d. Using the appropriate historgram, find a 98% bootstrap SE confidence interval for the true proportion of statistics majors who work at least 5 hours per week.  Interpret the confidence interval in the context of the problem.

1. **Assisted Reproduction.**
Assisted Reproductive Technology (ART) is a collection of techniques that help facilitate pregnancy (e.g., in vitro fertilization). A 2008 report by the Centers for Disease Control and Prevention estimated that ART has been successful in leading to a live birth in 31% of cases [@web:art]. A new fertility clinic claims that their success rate is higher than average. A random sample of 30 of their patients yielded a success rate of 40%. A consumer watchdog group would like to determine if this provides strong evidence to support the company's claim.

    ```{r}
    library(tidyverse)
    library(openintro)
    library(infer)
    
    set.seed(47)
    data.frame(fertility = c(rep("success", 12), rep("failure", 18))) %>%
      specify(response = fertility, success = "success") %>%
      hypothesize(null = "point", p = 0.31) %>%
      generate(reps = 10000, type = "simulate") %>%
      calculate(stat = "prop") %>% 
      ggplot() +
      geom_histogram(aes(x = stat), bins = 20) +
        labs(
          x = expression(hat(p)[sim]),
          y = "Count"
          ) +
      geom_vline(xintercept = 0.4, color = IMSCOL["red", 1])
    ```
    
     (a) Write the hypotheses to test if the success rate for ART at this clinic is significantly higher than the success rate reported by the CDC.

     (b) Describe a setup for a simulation that would be appropriate in this situation and how the p-value can be calculated using the simulation results.

     (c) Above is a histogram showing the distribution of $\hat{p}_{sim}$ in 10,000 simulations under the null hypothesis. Estimate the p-value using the plot and use it to evaluate the hypotheses.

     (d) After performing this analysis, the consumer group releases the following news headline: "Infertility clinic falsely advertises better success rates". Comment on the appropriateness of this statement.

1. **The Egyptian Revolution.**
A popular uprising that started on January 25, 2011 in Egypt led to the 2011 Egyptian Revolution. Polls show that about 69% of American adults followed the news about the political crisis and demonstrations in Egypt closely during the first couple weeks following the start of the uprising. Among a random sample of 30 high school students, it was found that only 17 of them followed the news about Egypt closely during this time. [@data:egypt]

    ```{r}
    library(tidyverse)
    library(openintro)
    library(infer)
    
    set.seed(47)
    data.frame(news = c(rep("follow", 17), rep("ignore", 13))) %>%
      specify(response = news, success = "follow") %>%
      hypothesize(null = "point", p = 0.69) %>%
      generate(reps = 10000, type = "simulate") %>%
      calculate(stat = "prop") %>% 
      ggplot() +
      geom_histogram(aes(x = stat), bins = 20) +
        labs(
          x = expression(hat(p)[sim]),
          y = "Count"
          ) +
      geom_vline(xintercept = 0.567, color = IMSCOL["red", 1])
    ```

    a. Write the hypotheses for testing if the proportion of high school students who followed the news about Egypt is different than the proportion of American adults who did.

    b. Calculate the proportion of high schoolers in this sample who followed the news about Egypt closely during this time.

    c. Describe how to perform a simulation and, once you had results, how to estimate the p-value.

    d. Above is a histogram showing the distribution of $\hat{p}_{sim}$ in 10,000 simulations under the null hypothesis. The red line is at the sample proportion $\hat{p} = 17/30.$  Estimate the p-value using the plot and determine the conclusion of the hypothesis test.

1. **National Health Plan.** 
A *Kaiser Family Foundation* poll for a random sample of US adults in 2019 found that 79% of Democrats, 55% of Independents, and 24% of Republicans supported a generic "National Health Plan". 
There were 347 Democrats, 298 Republicans, and 617 Independents surveyed. [@data:KFF2019nathealthplan]

    a. A political pundit on TV claims that a majority of Independents support a National Health Plan. Do these data provide strong evidence to support this type of statement?

    b. Would you expect a confidence interval for the proportion of Independents who oppose the public option plan to include 0.5? Explain.

1. **National Health Plan, Part II.** 
A *Kaiser Family Foundation* poll for a random sample of US adults in 2019 found that 79% of Democrats, 55% of Independents, and 24% of Republicans supported a generic "National Health Plan". 
There were 347 Democrats, 298 Republicans, and 617 Independents surveyed. [@data:KFF2019_nat_health_plan]

    A political pundit on TV claims that a majority of Independents support a National Health Plan. Do these data provide strong evidence to support this type of statement?  One approach to assessing the question of whether a majority of Independents support a National Health Plan is to simulate 1000 parametric bootstrap samples with p = 0.5 as the proportion of Independents in support.

    ```{r}
    library(tidyverse)
    library(infer)
    
    nhp <- data.frame(opinion = c(rep("yes plan", 339),rep("no plan", 278)))
    
    set.seed(47)
    nhp %>%
      specify(response = opinion, success = "yes plan") %>%
      hypothesize(null = "point", p = 0.50) %>%
      generate(reps = 1000, type = "simulate") %>%
      calculate(stat = "prop") %>%
      ggplot() + 
      geom_histogram(aes(x = stat)) +
      xlab("Bootstrapped sample proportions of national health care support, p = 0.50")
    
    ```
    
    a. The histogram above displays 1000 values of what?  

    b. Is the observed proportion of Independents consistent with the parametric bootstrap proportions under the setting where $p=0.5?$
    
    c. In order to test the claim that "a majority of Independents support a National Health Plan" what are the null and alternative hypotheses?
    
    d. Using the parametric bootstrap distribution, find the p-value and conclude the hypothesis test in the context of the problem.
    

1.  **CLT for proportions** Define the term "sampling distribution\" of
    the sample proportion, and describe how the shape, center, and
    spread of the sampling distribution change as the sample size
    increases when $p = 0.1$.

1. **Vegetarian college students.**
Suppose that 8% of college students are vegetarians. Determine if the following statements are true or false, and explain your reasoning.

    a. The distribution of the sample proportions of vegetarians in random samples of size 60 is approximately normal since $n \ge 30$.

    b. The distribution of the sample proportions of vegetarian college students in random samples of size 50 is right skewed.

    c. A random sample of 125 college students where 12% are vegetarians would be considered unusual.

    d. A random sample of 250 college students where 12% are vegetarians would be considered unusual.

    e. The standard error would be reduced by one-half if we increased the sample size from 125 to 250.

1. **Young Americans, Part I.** 
About 77% of young adults think they can achieve the American dream. 
Determine if the following statements are true or false, and explain your reasoning. [@news:youngAmericans1]

    a. The distribution of sample proportions of young Americans who think they can achieve the American dream in random samples of size 20 is left skewed.

    b. The distribution of sample proportions of young Americans who think they can achieve the American dream in random samples of size 40 is approximately normal since $n \ge 30$.

    c. A random sample of 60 young Americans where 85% think they can achieve the American dream would be considered unusual.

    d. A random sample of 120 young Americans where 85% think they can achieve the American dream would be considered unusual.

1. **Orange tabbies.** 
Suppose that 90% of orange tabby cats are male.
Determine if the following statements are true or false, and explain your reasoning.

    a. The distribution of sample proportions of random samples of size 30 is left skewed.

    b. Using a sample size that is 4 times as large will reduce the standard error of the sample proportion by one-half.

    c. The distribution of sample proportions of random samples of size 140 is approximately normal.

    d. The distribution of sample proportions of random samples of size 280 is approximately normal.

1. **Young Americans, Part II.**
About 25% of young Americans have delayed starting a family due to the continued economic slump.
Determine if the following statements are true or false, and explain your reasoning. [@news:youngAmericans2]

    a. The distribution of sample proportions of young Americans who have delayed starting a family due to the continued economic slump in random samples of size 12 is right skewed.

    b. In order for the distribution of sample proportions of young Americans who have delayed starting a family due to the continued economic slump to be approximately normal, we need random samples where the sample size is at least 40.

    c. A random sample of 50 young Americans where 20% have delayed starting a family due to the continued economic slump would be considered unusual.

    d. A random sample of 150 young Americans where 20% have delayed starting a family due to the continued economic slump would be considered unusual.

    e. Tripling the sample size will reduce the standard error of the sample proportion by one-third.

1. **Gender equality.**
The General Social Survey asked a random sample of 1,390 Americans the following question: "On the whole, do you think it should or should not be the government's responsibility to promote equality between men and women?" 82% of the respondents said it "should be". At a 95% confidence level, this sample has 2% margin of error. Based on this information, determine if the following statements are true or false, and explain your reasoning. [@data:gssgendereq]

    a. We are 95% confident that between 80% and 84% of Americans in this sample think it's the government's responsibility to promote equality between men and women.

    b. We are 95% confident that between 80% and 84% of all Americans think it's the government's responsibility to promote equality between men and women.

    c. If we considered many random samples of 1,390 Americans, and we calculated 95% confidence intervals for each, 95% of these intervals would include the true population proportion of Americans who think it's the government's responsibility to promote equality between men and women.

    d. In order to decrease the margin of error to 1%, we would need to quadruple (multiply by 4) the sample size.

    e. Based on this confidence interval, there is sufficient evidence to conclude that a majority of Americans think it's the government's responsibility to promote equality between men and women.

1. **Elderly drivers.** 
The Marist Poll published a report stating that 66% of adults nationally think licensed drivers should be required to retake their road test once they reach 65 years of age. It was also reported that interviews were conducted on a random sample of 1,018 American adults, and that the margin of error was 3% using a 95% confidence level. [@data:elderlyDriving]

    a. Verify the margin of error reported by The Marist Poll.

    b. Based on a 95% confidence interval, does the poll provide convincing evidence that *more than* two thirds of the population think that licensed drivers should be required to retake their road test once they turn 65?

1. **Fireworks on July 4$^{\text{th}}$.** 
A local news outlet reported that 56% of 600 randomly sampled Kansas residents planned to set off fireworks on July $4^{th}$. 
Determine the margin of error for the 56% point estimate using a 95% confidence level. [@data:july4]

1. **Proof of COVID-19 vaccination.**
A Gallup poll surveyed 3,731 randomly sampled US in April 2021, asking how they felt about requiring proof of COVID-19 vaccination for travel by airplane. 
The poll found that 57% said they would favor it. [@data:gallupcovidvaccine]

    a. Describe the population parameter of interest. What is the value of the point estimate of this parameter?

    b. Check if the conditions required for constructing a confidence interval using mathematical models based on these data are met.

    c. Construct a 95% confidence interval for the proportion of US adults who favor requiring proof of COVID-19 vaccination for travel by airplane.

    d. Without doing any calculations, describe what would happen to the confidence interval if we decided to use a higher confidence level.

    e. Without doing any calculations, describe what would happen to the confidence interval if we used a larger sample.

1. **Study abroad** 
A survey on 1,509 high school seniors who took the SAT and who completed an optional web survey shows that 55% of high school seniors are fairly certain that they will participate in a study abroad program in college. [@data:studyAbroad]

    a.  Is this sample a representative sample from the population of all high school seniors in the US? Explain your reasoning.

    b.  Let's suppose the conditions for inference are met. Even if your answer to part (a) indicated that this approach would not be reliable, this analysis may still be interesting to carry out (though not report). Construct a 90% confidence interval for the proportion of high school seniors (of those who took the SAT) who are fairly certain they will participate in a study abroad program in college, and interpret this interval in context.

    c.  What does "90% confidence" mean?

    d.  Based on this interval, would it be appropriate to claim that the majority of high school seniors are fairly certain that they will participate in a study abroad program in college?

1. **Legalization of marijuana.**
The General Social Survey asked a random sample of 1,563 US adults: "Do you think the use of marijuana should be made legal, or not?" 60% of the respondents said it should be made legal. [@data:gssgrass]

    a. Is 60% a sample statistic or a population parameter? Explain.

    b. Construct a 95% confidence interval for the proportion of US adults who think marijuana should be made legal, and interpret it in the context of the data.

    c. A critic points out that this 95% confidence interval is only accurate if the statistic follows a normal distribution, or if the normal model is a good approximation. Is this true for these data? Explain.

    d. A news piece on this survey's findings states, "Majority of US adults think marijuana should be legalized." Based on your confidence interval, is this news piece's statement justified?

1. **National Health Plan.** 
A *Kaiser Family Foundation* poll for a random sample of US adults in 2019 found that 79% of Democrats, 55% of Independents, and 24% of Republicans supported a generic "National Health Plan". 
There were 347 Democrats, 298 Republicans, and 617 Independents surveyed. [@data:KFF2019nathealthplan]

    a. A political pundit on TV claims that a majority of Independents support a National Health Plan. Do these data provide strong evidence to support this type of statement?

    b. Would you expect a confidence interval for the proportion of Independents who oppose the public option plan to include 0.5? Explain.

1. **Is college worth it?**
Among a simple random sample of 331 American adults who do not have a four-year college degree and are not currently enrolled in school, 48% said they decided not to go to college because they could not afford school. [@data:collegeWorthIt]

    a. A newspaper article states that only a minority of the Americans who decide not to go to college do so because they cannot afford it and uses the point estimate from this survey as evidence. Conduct a hypothesis test to determine if these data provide strong evidence supporting this statement.

    b. Would you expect a confidence interval for the proportion of American adults who decide not to go to college because they cannot afford it to include 0.5? Explain.

1. **Taste test.**
Some people claim that they can tell the difference between a diet soda and a regular soda in the first sip.
A researcher wanting to test this claim randomly sampled 80 such people.
He then filled 80 plain white cups with soda, half diet and half regular through random assignment, and asked each person to take one sip from their cup and identify the soda as diet or regular.
53 participants correctly identified the soda.

    a. Do these data provide strong evidence that these people are able to detect the difference between diet and regular soda, in other words, are the results significantly better than just random guessing?

    b. Interpret the p-value in this context.

1. **Will the coronavirus bring the world closer together?**
An April 2021 YouGov poll asked 4,265 UK adults whether they think the coronavirus bring the world closer together or leave us further apart. 
12% of the respondents said it will bring the world closer together. 37% said it would leave us further apart, 39% said it won't make a difference and the remainder didn't have an opinion on the matter. [@data:yougovcovid]

    a. Calculate, using mathematical models, a 90% confidence interval for the proportion of UK adults who think the coronavirus will bring the world closer together, and interpret the interval in context.

    b. Suppose we wanted the margin of error for the 90% confidence level to be about 0.5%. How large of a sample size would you recommend for the poll?

1.  **Quality control** As part of a quality control process for
    computer chips, an engineer at a factory randomly samples 212 chips
    during a week of production to test the current rate of chips with
    severe defects. She finds that 27 of the chips are defective.

    a.  What population is under consideration in the data set?

    b.  What parameter is being estimated?

    c.  What is the point estimate for the parameter?

    d.  What is the name of the statistic can we use to measure the uncertainty of the point estimate?

    e.  Compute the value of the statistic from part (d) for this context.

    f.  The historical rate of defects is 10%. Should the engineer be surprised by the observed rate of defects during the current week?

    g.  Suppose the true population value was found to be 10%. If we use this proportion to recompute the value in part (d) using $p = 0.1$ instead of $\hat{p}$, how much does the resulting value of the statistic change?

1.  **Nearsighted** It is believed that nearsightedness affects about 8%
    of all children. In a random sample of 194 children, 21 are
    nearsighted. Conduct a hypothesis test for the following question:
    do these data provide evidence that the 8% value is inaccurate?

1.  **Coupons driving visits** A store randomly samples 603 shoppers
    over the course of a year and finds that 142 of them made their
    visit because of a coupon they'd received in the mail. Construct a
    95% confidence interval for the fraction of all shoppers during the
    year whose visit was because of a coupon they'd received in the
    mail.
