1. **Disaggregating Asian American tobacco use, hypothesis testing.**
Understanding cultural differences in tobacco use across different demographic groups can lead to improved health care education and treatment. A recent study disaggregated tobacco use across Asian American ethnic groups including Asian-Indian (n = 4,373), Chinese (n = 4,736), and Filipino (n = 4,912), in comparison to non-Hispanic Whites (n = 275,025). The number of current smokers in each group was reported as Asian-Indian (n = 223), Chinese (n = 279), Filipino (n = 609), and non-Hispanic Whites (n = 50,880). [@Rao:2021]

    To determine whether the proportion of Asian-Indian Americans who are current smokers is different from the proportion of Chinese Americans who are smokers, a randomization simulation was performed.

    ```{r}
    library(tidyverse)
    library(infer)
    library(openintro)

    asian_smoke <- tibble(
      ethnic = c(rep("Indian", 4373), rep("Chinese", 4736)),
      outcome = c(
        rep("smoke", 223), rep("healthy", 4150),
        rep("smoke", 279), rep("healthy", 4457)
      )
    )

    set.seed(47)
    asian_smoke %>%
      specify(outcome ~ ethnic, success = "smoke") %>%
      hypothesize(null = "independence") %>%
      generate(reps = 1000, type = "permute") %>%
      calculate(stat = "diff in props", order = c("Indian", "Chinese")) %>%
      # summarize(mean(stat < -0.008))
      ggplot(aes(x = stat)) +
      geom_histogram(binwidth = 0.0015, fill = IMSCOL["green", "full"]) +
      labs(
        title = "1,000 randomized differences",
        x = "Difference in randomized proportions\n(Indian - Chinese)",
        y = "Count"
        )
    ```

    a. In both words and symbols provide the parameter and statistic of interest for this study. Do you know the numerical value of either the parameter or statisic of interest?  If so, provide the numerical value.
   
    b. The histogram above provides the sampling distribution (under randomization) for $\hat{p}_{Asian-Indian} - \hat{p}_{Chinese}$ under repeated null randomizations ($\hat{p}$ is the proportion in the sample who are current smokers). Estimate the standard error of $\hat{p}_{Asian-Indian} - \hat{p}_{Chinese}$ based on the randomization histogram.
   
    c. Consider the hypothesis test to determine if there is a difference in proportion of Asian-Indian Americans as compared to Chinese Americans who are current smokers. Write out the null and alternative hypotheses, estimate a p-value using the randomization histogram, and conclude the test in the context of the problem.
    
    \clearpage

1. **Malaria vaccine effectiveness, hypothesis test.**
With no currently licensed vaccines to inhibit malaria, good news was welcomed with a recent study reporting long-awaited vaccine success for children in Burkina Faso. With 450 children randomized to either one of two different doses of the malaria vaccine or a control vaccine, 89 of 292 malaria vaccine and 106 out of 147 control vaccine children contracted malaria within 12 months after the treatment. [@Datoo:2021]

    ```{r}
    library(tidyverse)
    library(infer)

    malaria <- tibble(
      vaccine = c(rep("malaria", 292), rep("control", 147)),
      outcome = c(
        rep("sick", 89), rep("healthy", 203),
        rep("sick", 106), rep("healthy", 41)
      )
    )

    set.seed(47)
    malaria %>%
      specify(outcome ~ vaccine, success = "sick") %>%
      hypothesize(null = "independence") %>%
      generate(reps = 1000, type = "permute") %>%
      calculate(stat = "diff in props", order = c("malaria", "control")) %>%
      ggplot(aes(x = stat)) +
      geom_histogram(fill = IMSCOL["green", "full"], binwidth = 0.02) +
      labs(
        title = "1,000 randomized differences",
        x = "Difference in randomized proportions\n(malaria - control)",
        y = "Count"
      )
    ```

    a. In both words and symbols provide the parameter and statistic of interest for this study. Do you know the numerical value of either the parameter or statisic of interest? If so, provide the numerical value.
   
    b. The histogram above provides the sampling distribution (under randomization) for $\hat{p}_{malaria} - \hat{p}_{control}$ under repeated null randomizations ($\hat{p}$ is the proportion of children in the sample who contracted malaria). Estimate the standard error of $\hat{p}_{malaria} - \hat{p}_{control}$ based on the randomization histogram.
   
    c. Consider the hypothesis test constructed to show a lower proportion of children contracting malaria on the malaria vaccine as compared to the control vaccine. Write out the null and alternative hypotheses, estimate a p-value using the randomization histogram, and conclude the test in the context of the problem.
    
    \clearpage

1. **Disaggregating Asian American tobacco use, confidence interval.**
Based on a study on the degree to which smoking practices differ across ethnic groups, a confidence interval for the difference in current smoking status for Filipino versus Chinese Americans is desired. [@Rao:2021]

    ```{r}
    library(tidyverse)
    library(infer)
    library(openintro)

    asian_smoke <- tibble(
      ethnic = c(rep("Filipino", 4912), rep("Chinese", 4736)),
      outcome = c(
        rep("smoke", 609), rep("healthy", 4303),
        rep("smoke", 279), rep("healthy", 4457)
      )
    )
    
    set.seed(47)
    asian_smoke %>%
      specify(outcome ~ ethnic, success = "smoke") %>%
      generate(reps = 1000, type = "bootstrap") %>%
      calculate(stat = "diff in props", order = c("Filipino", "Chinese")) %>%
      ggplot(aes(x = stat)) +
      geom_histogram(binwidth = 0.0015, fill = IMSCOL["green", "full"]) +
      labs(
        title = "1,000 bootstrapped differences",
        x = "Difference in bootstrapped proportions\n(Filipino - Chinese)",
        y = "Count"
        )
    ```

    a. Consider the bootstrap distribution of difference in sample proportions of current smokers (Filipino Americans minus Chinese Americans) in 1,000 bootstrap repetitions as above. Estimate the standard error of the difference in sample proportions, as seen in the histogram.
    
    b. Using the standard error from the bootstrap distribution, find a 95% bootstrap SE confidence interval for the true difference in proportion of current smokers (Filipino Americans minus Chinese Americans) in the population. Interpret the interval in the context of the problem.
    
    c. Using the entire bootstrap distribution, find a 95% bootstrap percentile confidence interval for the true difference in proportion of current smokers (Filipino Americans minus Chinese Americans) in the population. Interpret the interval in the context of the problem.
    
    \clearpage

1. **Malaria vaccine effectiveness, confidence interval.**
With no currently licensed vaccines to inhibit malaria, good news was welcomed with a recent study reporting long-awaited vaccine success for children in Burkina Faso. With 450 children randomized to either one of two different doses of the malaria vaccine or a control vaccine, 89 of 292 malaria vaccine and 106 out of 147 control vaccine children contracted malaria within 12 months after the treatment. [@Datoo:2021]

    ```{r}
    library(tidyverse)
    library(infer)
    
    malaria <- tibble(
      vaccine = c(rep("malaria", 292), rep("control", 147)),
      outcome = c(
        rep("sick", 89), rep("healthy", 203),
        rep("sick", 106), rep("healthy", 41)
      )
    )
    
    set.seed(47)
    malaria %>%
      specify(outcome ~ vaccine, success = "sick") %>%
      generate(reps = 1000, type = "bootstrap") %>%
      calculate(stat = "diff in props", order = c("malaria", "control")) %>%
      ggplot(aes(x = stat)) +
      geom_histogram(fill = IMSCOL["green", "full"], binwidth = 0.02) +
      labs(
        title = "1,000 bootstrapped differences",
        x = "Difference in bootstrapped proportions\n(malaria - control)",
        y = "Count"
      )
    ```

    a. Consider the bootstrap distribution of difference in sample proportions of children who contracted malaria (malaria vaccine minus control vaccine) in 1000 bootstrap repetitions as above. Estimate the standard error of the difference in sample proportions, as seen in the histogram.
    
    b. Using the standard error from the bootstrap distribution, find a 95% bootstrap SE confidence interval for the true difference in proportion of children who contract malaria (malaria vaccine minus control vaccine) in the population. Interpret the interval in the context of the problem.
    
    c. Using the entire bootstrap distribution, find a 95% bootstrap percentile confidence interval for the true difference in proportion of children who contract malaria (malaria vaccine minus control vaccine) in the population. Interpret the interval in the context of the problem.
    
    \clearpage

1. **COVID-19 and degree completion.**
A 2021 Gallup poll surveyed 3,941 students pursuing a bachelor's degree and 2,064 students pursuing an associate degree (students were not randomly selected but were weighted so as to represent a random selection of currently enrolled US college students). The poll found that 51% of the bachelor's degree students and 44% of associate degree students said that the COVID-19 pandemic will negatively impact their ability to complete the degree. [@data:gallupcollegeimpact]

    Below are two histograms which represent different computational approaches (both use 1,000 repetitions) to research questions which could be asked from the Gallup data which was provided. One of the histograms can be used to do a randomization test on whether the proportions of bachelor's and associate students who think the COVID-19 pandemic will negatively impact their ability to complete the degree. The other histogram is a bootstrap distribution used to quantify the difference in the proportions of bachelor's and associate's students who feel this way.

    ```{r}
    library(tidyverse)
    library(infer)
    library(openintro)
    library(patchwork)

    covid_college <- data.frame(
      degree = c(rep("BA", 3941), rep("AA", 2064)),
      outcome = c(
        rep("impact", 2010), rep("no impact", 1931),
        rep("impact", 908), rep("no impact", 1156)
      )
    )

    set.seed(47)
    p_A <- covid_college %>%
      specify(outcome ~ degree, success = "impact") %>%
      generate(reps = 1000, type = "bootstrap") %>%
      calculate(stat = "diff in props", order = c("BA", "AA")) %>%
      ggplot(aes(x = stat)) +
      geom_histogram(fill = IMSCOL["green", "full"], bins = 20) +
      labs(
        title = "Computational method A",
        x = "Difference in simulated proportions\n(BA - AA)",
        y = "Count"
      )

    set.seed(47)
    p_B <- covid_college %>%
      specify(outcome ~ degree, success = "impact") %>%
      hypothesize(null = "independence") %>%
      generate(reps = 1000, type = "permute") %>%
      calculate(stat = "diff in props", order = c("BA", "AA")) %>%
      ggplot(aes(x = stat)) +
      geom_histogram(fill = IMSCOL["green", "full"], bins = 20) +
      labs(
        title = "Computational method B",
        x = "Difference in simulated proportions\n(BA - AA)",
        y = "Count"
      )
    
    p_A / p_B
    ```

    a. Are the center and standard error of the two graphs approximately the same?  Explain.
    
    b. Write a research question which could be addressed using this histogram with computational method A.
    
    c. Write a research question which could be addressed using this histogram with computational method B.
    
    \clearpage

1. **Renewable energy.**
A 2021 Gallup poll surveyed 5,447 randomly sampled US adults who are Republican (or Republican leaning) and 7,962 who are Democrats (or Democrat leaning). 31% of Republicans and 81% of Democrats said "government regulations are necessary to encourage businesses and consumers to rely more on renewable energy sources". [@data:gallupcollegeimpact]

    Below are two histograms which represent different computational approaches (both use 1,000 repetitions) to research questions which could be asked from the Gallup data which was provided. One of the histograms can be used to do a randomization test on whether the proportions of Republicans and Democrats who think government regulations are necessary to encourage businesses and consumers to rely more on renewable energy sources are different. The other histogram is a bootstrap distribution used to quantify the difference in the proportions of Republicans and Democrats who agree with this statement.

    ```{r}
    library(tidyverse)
    library(infer)
    library(openintro)
    library(patchwork)

    renewable_energy <- data.frame(
      party = c(rep("Republican", 5447), rep("Democrat", 7962)),
      opinion = c(
        rep("regs necessary", 1689), rep("regs not necessary", 3758),
        rep("regs necessary", 6449), rep("regs not necessary", 1513)
      )
    )

    set.seed(47)
    p_B <- renewable_energy %>%
      specify(opinion ~ party, success = "regs necessary") %>%
      generate(reps = 1000, type = "bootstrap") %>%
      calculate(stat = "diff in props", order = c("Republican", "Democrat")) %>%
      ggplot(aes(x = stat)) +
      geom_histogram(fill = IMSCOL["green", "full"], bins = 20) +
      labs(
        title = "Computational Method B",
        x = "Difference in simulated proportions\n(Republican - Democrat)",
        y = "Count"
      )

    set.seed(47)
    p_A <- renewable_energy %>%
      specify(opinion ~ party, success = "regs necessary") %>%
      hypothesize(null = "independence") %>%
      generate(reps = 1000, type = "permute") %>%
      calculate(stat = "diff in props", order = c("Republican", "Democrat")) %>%
      ggplot(aes(x = stat)) +
      geom_histogram(fill = IMSCOL["green", "full"], bins = 20) +
      labs(
        title = "Computational Method A",
        x = "Difference in simulated proportions\n(Republican - Democrat)",
        y = "Count"
      )
    
    p_A / p_B
    ```

    a. Are the center and standard error of the two graphs approximately the same?  Explain.
    
    b. Write a research question which could be addressed using this hitogram with computational method A.
    
    c. Write a research question which could be addressed using this hitogram with computational method B.

1. **HIV in sub-Saharan Africa.**
In July 2008 the US National Institutes of Health announced that it was stopping a clinical study early because of unexpected results. The study population consisted of HIV-infected women in sub-Saharan Africa who had been given single dose Nevaripine (a treatment for HIV) while giving birth, to prevent transmission of HIV to the infant. The study was a randomized comparison of continued treatment of a woman (after successful childbirth) with Nevaripine vs Lopinavir, a second drug used to treat HIV. 240 women participated in the study; 120 were randomized to each of the two treatments. Twenty-four weeks after starting the study treatment, each woman was tested to determine if the HIV infection was becoming worse (an outcome called *virologic failure*). Twenty-six of the 120 women treated with Nevaripine experienced virologic failure, while 10 of the 120 women treated with the other drug experienced virologic failure. [@Lockman:2007]

    a. Create a two-way table presenting the results of this study.

    b. State appropriate hypotheses to test for difference in virologic failure rates between treatment groups.

    c. Complete the hypothesis test and state an appropriate conclusion. (Reminder: Verify any necessary conditions for the test.)
    
    \clearpage

1. **Supercommuters.**
The fraction of workers who are considered "supercommuters", because they commute more than 90 minutes to get to work, varies by state. Suppose the 1% of Nebraska residents and 6% of New York residents are supercommuters. Now suppose that we plan a study to survey 1000 people from each state, and we will compute the sample proportions $\hat{p}_{NE}$ for Nebraska and $\hat{p}_{NY}$ for New York.

    a. What is the associated mean and standard deviation of $\hat{p}_{NE}$ in repeated samples of size 1000?

    b. What is the associated mean and standard deviation of $\hat{p}_{NY}$ in repeated samples of size 1000?

    c. Calculate and interpret the mean and standard deviation associated with the difference in sample proportions for the two groups, $\hat{p}_{NY} - \hat{p}_{NE}$ in repeated samples of 1000 in each group.

    d. How are the standard deviations from parts (a), (b), and (c) related?

1. **National Health Plan.**
A Kaiser Family Foundation poll for US adults in 2019 found that 79% of Democrats, 55% of Independents, and 24% of Republicans supported a generic “National Health Plan”. There were 347 Democrats, 298 Republicans, and 617 Independents surveyed. 79% of 347 Democrats and 55% of 617 Independents support a National Health Plan. [@data:KFF2019nathealthplan]

    a. Calculate a 95% confidence interval for the difference between the proportion of Democrats and Independents who support a National Health Plan $(p_{D} - p_{I})$, and interpret it in this context. We have already checked conditions for you.

    b. True or false: If we had picked a random Democrat and a random Independent at the time of this poll, it is more likely that the Democrat would support the National Health Plan than the Independent.

1. **Sleep deprivation, CA vs. OR, confidence interval.**
According to a report on sleep deprivation by the Centers for Disease Control and Prevention, the proportion of California residents who reported insufficient rest or sleep during each of the preceding 30 days is 8.0%, while this proportion is 8.8% for Oregon residents. These data are based on simple random samples of 11,545 California and 4,691 Oregon residents. Calculate a 95% confidence interval for the difference between the proportions of Californians and Oregonians who are sleep deprived and interpret it in context of the data. [@data:sleepCAandOR]

1.  **Gender pay gap in medicine.**
A study examined the average pay for men and women entering the workforce as doctors for 21 different positions. [@LoSassoMedicineGenderPayGap]

    a. If each gender was equally paid, then we would expect about half of those positions to have men paid more than women and women would be paid more than men in the other half of positions. Write appropriate hypotheses to test this scenario.

    b. Men were, on average, paid more in 19 of those 21 positions. Complete a hypothesis test using your hypotheses from part (a).

1. **Sleep deprivation, CA vs. OR, hypothesis test.**
A CDC report on sleep deprivation rates shows that the proportion of California residents who reported insufficient rest or sleep during each of the preceding 30 days is 8.0%, while this proportion is 8.8% for Oregon residents. These data are based on simple random samples of 11,545 California and 4,691 Oregon residents.

    a. Conduct a hypothesis test to determine if these data provide strong evidence that the rate of sleep deprivation is different for the two states. (Reminder: Check conditions)

    b. It is possible the conclusion of the test in part (a) is incorrect. If this is the case, what type of error was made?
    
    \clearpage

1. **Is yawning contagious?**
An experiment conducted by the MythBusters, a science entertainment TV program on the Discovery Channel, tested if a person can be subconsciously influenced into yawning if another person near them yawns. 50 people were randomly assigned to two groups: 34 to a group where a person near them yawned (treatment) and 16 to a group where there wasn't a person yawning near them (control). The visualization below displays how many participants yawned in each group.^[The [`yawn`](http://openintrostat.github.io/openintro/reference/yawn.html) data used in this exercise can be found in the [**openintro**](http://openintrostat.github.io/openintro) R package.]

    ```{r fig.asp = 0.45}
    library(openintro)
    library(tidyverse)
    
    yawn %>%
      mutate(group = if_else(group == "ctrl", "Control", "Treatment")) %>%
      ggplot(aes(y = group, fill = result)) +
      geom_bar() +
      labs(x = "Count", y = "Group", fill = "Outcome") +
      scale_fill_manual(values = c(IMSCOL["blue", "full"], IMSCOL["red", "full"])) +
      scale_x_continuous(breaks = seq(0, 35, 5)) +
      theme(
        legend.position = c(0.85, 0.3),
        legend.background = element_rect(fill = "white", color = "white")
      )
    ```
    
    Suppose we are interested in estimating the difference in yawning rates between the control and treatment groups using a confidence interval. Explain why we cannot construct such an interval using the normal approximation. What might go wrong if we constructed the confidence interval despite this problem?
    
    \vspace{5mm}

1. **Heart transplant success.**
The Stanford University Heart Transplant Study was conducted to determine whether an experimental heart transplant program increased lifespan. Each patient entering the program was officially designated a heart transplant candidate, meaning that he was gravely ill and might benefit from a new heart. Patients were randomly assigned into treatment and control groups. Patients in the treatment group received a transplant, and those in the control group did not. The visualization below displays how many patients survived and died in each group.^[The [`heart_transplant`](http://openintrostat.github.io/openintro/reference/heart_transplant.html) data used in this exercise can be found in the [**openintro**](http://openintrostat.github.io/openintro) R package.] [@Turnbull+Brown+Hu:1974]

    ```{r fig.asp = 0.45}
    library(openintro)
    library(tidyverse)
    
    heart_transplant %>%
      mutate(
        survived = as.character(survived),
        survived = if_else(survived == "dead", "deceased", survived)
        ) %>%
      ggplot(aes(y = transplant, fill = survived)) +
      geom_bar() +
      labs(x = "Count", y = "Group", fill = "Outcome") +
      scale_fill_manual(values = c(IMSCOL["blue", "full"], IMSCOL["red", "full"])) +
      theme(
        legend.position = c(0.85, 0.3),
        legend.background = element_rect(fill = "white", color = "white")
      )
    ```
    
    Suppose we are interested in estimating the difference in survival rate between the control and treatment groups using a confidence interval. Explain why we cannot construct such an interval using the normal approximation. What might go wrong if we constructed the confidence interval despite this problem?
    
    \clearpage

1.  **Government shutdown.**
The United States federal government shutdown of 2018--2019 occurred from December 22, 2018 until January 25, 2019, a span of 35 days. A Survey USA poll of 614 randomly sampled Americans during this time period reported that 48% of those who make less than \$40,000 per year and 55% of those who make \$40,000 or more per year said the government shutdown has not at all affected them personally. A 95% confidence interval for $(p_\text{$<$40K} - p_\text{$\ge$40K})$, where $p$ is the proportion of those who said the government shutdown has not at all affected them personally, is (-0.16, 0.02). Based on this information, determine if the following statements are true or false, and explain your reasoning if you identify the statement as false. [@data:govtshuthown]

    a. At the 5% significance level, the data provide convincing evidence of a real difference in the proportion who are not affected personally between Americans who make less than \$40,000 annually and Americans who make \$40,000 annually.

    b. We are 95% confident that 16% more to 2% fewer Americans who make less than \$40,000 per year are not at all personally affected by the government shutdown compared to those who make \$40,000 or more per year.

    c. A 90% confidence interval for $(p_\text{$<$40K} - p_\text{$\ge$40K})$ would be wider than the $(-0.16, 0.02)$ interval.

    d. A 95% confidence interval for $(p_\text{$\ge$40K} - p_\text{$<$40K})$ is (-0.02, 0.16).

1.  **Online harassment.**
A Pew Research poll asked US adults aged 18-29 and 30-49 whether they have personally experienced harassment online.
A 95% confidence interval for the difference between the proportions of 18-29 year olds and 30-49 year olds who have personally experienced harassment online $(p_{18-29} - p_{30-49})$ was calculated to be (0.115, 0.185).
Based on this information, determine if the following statements are true or false, and explain your reasoning for each statement you identify as false. [@onlineharassment2021]

    a. We are 95% confident that the true proportion of 18-29 year olds who have personally experienced harassment online is 11.5% to 18.5% lower than the true proportion of 30-49 year olds who have personally experienced harassment online.

    b. We are 95% confident that the true proportion of 18-29 year olds who have personally experienced harassment online is 11.5% to 18.5% higher than the true proportion of 30-49 year olds who have personally experienced harassment online.

    c. 95% of random samples will produce 95% confidence intervals that include the true difference between the population proportions of 18-29 year olds and 30-49 year olds who have personally experienced harassment online.

    d. We can conclude that there is a significant difference between the proportions of 18-29 year olds and 30-49 year olds who have personally experienced harassment online is too large to plausibly be due to chance, if in fact there is no difference between the two proportions.

    e. The 90% confidence interval for $(p_{18-29} - p_{30-49})$ cannot be calculated with only the information given in this exercise.


```{r include=FALSE}
# https://www.pewresearch.org/internet/2021/01/13/the-state-of-online-harassment/
# https://www.pewresearch.org/internet/2021/01/13/online-harassment-methodology/
n_young <- 983
n_old   <- 3239

p_young <- 0.64
p_old   <- 0.49

(p_young - p_old) + c(-1,1)*1.96*sqrt((p_young * (1-p_young)/n_young) + (p_old * (1-p_old)/n_old))
```

1. **Decision errors and comparing proportions, I.**
In the following research studies, conclusions were made based on the data provided. It is always possible that the analysis conclusion could be wrong, although we will almost never actually know if an error has been made or not.  For each study conclusion, specify which of a Type 1 or Type 2 error could have been made, and state the error in the context of the problem.

    a. The malaria vaccine was seen to be effective at lowering the rate of contracting malaria (when compared to the control vaccine).
    
    b. In the US population, Asian-Indian Americans and Chinese Americans are not observed to have different proportions of current smokers.
    
    c. There is no evidence to claim a difference in the proportion of Americans who are not affected personally by a government shutdown when comparing Americans who make less than \$40,000 annually and Americans who make \$40,000 annually.
    
    \clearpage

1. **Decision errors and comparing proportions, II.**
In the following research studies, conclusions were made based on the data provided.  It is always possible that the analysis conclusion could be wrong, although we will almost never actually know if an error has been made or not.  For each study conclusion, specify which of a Type 1 or Type 2 error could have been made, and state the error in the context of the problem.
    
    a. Of registered voters in California, the proportion who report not knowing enough to voice an opinion on whether they support off shore drilling is different across those who have a college degree and those who do not.
    
    b. In comparing Californians and Oregonians, there is no evidence to support a difference in the proportion of each who are sleep deprived.
    
    \vspace{5mm}

1. **Active learning.**
A teacher wanting to increase the active learning component of her course is concerned about student reactions to changes she is planning to make. She conducts a survey in her class, asking students whether they believe more active learning in the classroom (hands on exercises) instead of traditional lecture will helps improve their learning. She does this at the beginning and end of the semester and wants to evaluate whether students' opinions have changed over the semester. Can she used the methods we learned in this chapter for this analysis? Explain your reasoning.

    \vspace{5mm}

1. **An apple a day keeps the doctor away.**
A physical education teacher at a high school wanting to increase awareness on issues of nutrition and health asked her students at the beginning of the semester whether they believed the expression "an apple a day keeps the doctor away", and 40% of the students responded yes. Throughout the semester she started each class with a brief discussion of a study highlighting positive effects of eating more fruits and vegetables. She conducted the same apple-a-day survey at the end of the semester, and this time 60% of the students responded yes. Can she used a two-proportion method from this section for this analysis? Explain your reasoning.

    \vspace{5mm}

1. **Malaria vaccine effectiveness, effect size.**
A randomized controlled trial on malaria vaccine effectiveness randomly assigned 450 children intro either one of two different doses of the malaria vaccine or a control vaccine. 89 of 292 malaria vaccine and 106 out of 147 control vaccine children contracted malaria within 12 months after the treatment. [@Datoo:2021]

    Recall that in order to reject the null hypothesis that the two vaccines (malaria and control) are equivalent, we would need the sample proportion to be about 2 standard errors below the hypothesized value of zero.

    Say that the true difference (in the population) is given as $\delta,$ the sample sizes are the same in both groups $(n_{malaria} = n_{control}),$ and the true proportion who contract malaria on the control vaccine is $p_{control} = 0.7.$ If you ran your own study (in the future), how likely is it that you would get a difference in sample proportions that was sufficiently far from zero that you could reject under each of the conditions below. (*Hint:* Use the mathematical model.)

    a. $\delta = -0.1$ and $n_{malaria} = n_{control} = 20$
    
    b. $\delta = -0.4$ and $n_{malaria} = n_{control} = 20$
    
    c. $\delta = -0.1$ and $n_{malaria} = n_{control} = 100$
    
    d. $\delta = -0.4$ and $n_{malaria} = n_{control} = 100$
    
    e. What can you conclude about values of $\delta$ and the sample size?
    
    \clearpage

1.  **Diabetes and unemployment.**
A Gallup poll surveyed Americans about their employment status and whether they have diabetes. The survey results indicate that 1.5% of the 47,774 employed (full or part time) and 2.5% of the 5,855 unemployed 18-29 year olds have diabetes. [@data:employmentDiabetes]

    a. Create a two-way table presenting the results of this study.

    b. State appropriate hypotheses to test for difference in proportions of diabetes between employed and unemployed Americans.

    c. The sample difference is about 1%. If we completed the hypothesis test, we would find that the p-value is very small (about 0), meaning the difference is statistically significant. Use this result to explain the difference between statistically significant and practically significant findings.
