1. **Statistics vs. parameters: one mean.**
Each of the following scenarios were set up to assess an average value. For each one, identify, in words: the statistic and the parameter.

    a.  A sample of 25 New Yorkers were asked how much sleep they get per night.
    
    b.  Researchers at two different universities in California collected information on undergraduates' heights.

1. **Statistics vs. parameters: one mean.**
Each of the following scenarios were set up to assess an average value.  For each one, identify, in words: the statistic and the parameter.
    
    a.  Georgianna samples 20 children from a particular city and measures how many years they have each been playing piano.
    
    b.  Traffic police officers (who are regularly exposed to lead from automobile exhaust) had their lead levels measured in their blood.

1. **Heights of adults.** 
Researchers studying anthropometry collected body measurements, as well as age, weight, height and gender, for 507 physically active individuals. 
Summary statistics for the distribution of heights (measured in centimeters), along with a histogram, are provided below.^[The [`bdims`](http://openintrostat.github.io/openintro/reference/bdims.html) data used in this exercise can be found in the [**openintro**](http://openintrostat.github.io/openintro) R package.] [@Heinz:2003]

    ```{r out.width = "70%"}
    library(tidyverse)
    library(kableExtra)
    library(openintro)
    
    bdims %>%
      summarise(
        Min    = round(min(hgt), 1),
        Q1     = round(quantile(hgt, 0.25), 1),
        Median = round(median(hgt), 1),
        Mean   = round(mean(hgt), 1),
        Q3     = round(quantile(hgt, 0.75), 1),
        Max    = round(max(hgt), 1),
        SD     = round(sd(hgt), 1),
        IQR    = round(IQR(hgt), 1)
      ) %>%
      kbl(linesep = "", booktabs = TRUE, align = "cccccccc") %>%
      kable_styling(bootstrap_options = c("striped", "condensed"), 
                    latex_options = "HOLD_position",
                    full_width = FALSE)
    
    ggplot(bdims, aes(x = hgt)) +
      geom_histogram(binwidth = 2) +
      labs(x = "Height (centimeters)", y = "Count")
    ```

    a.  What is the point estimate for the average height of active individuals? What about the median?

    b.  What is the point estimate for the standard deviation of the heights of active individuals? What about the IQR?

    c.  Is a person who is 1m 80cm (180 cm) tall considered unusually tall? And is a person who is 1m 55cm (155cm) considered unusually short? Explain your reasoning.

    d.  The researchers take another random sample of physically active individuals. Would you expect the mean and the standard deviation of this new sample to be the ones given above? Explain your reasoning.

    e.  The sample means obtained are point estimates for the mean height of all active individuals, if the sample of individuals is equivalent to a simple random sample. What measure do we use to quantify the variability of such an estimate? Compute this quantity using the data from the original sample under the condition that the data are a simple random sample.

1. **Heights of adults, standard error.**
Heights of 507 physically active individuals have a mean of 171 centimeters and a standard deviation of 9.4 centimeters.
Provide an estimate for the standard error of the mean for samples of following sizes.^[The [`bdims`](http://openintrostat.github.io/openintro/reference/bdims.html) data used in this exercise can be found in the [**openintro**](http://openintrostat.github.io/openintro) R package.] [@Heinz:2003]

     a. n = 10
     
     b. n = 50
     
     c. n = 100
     
     d. n = 1000
     
     e. The standard error of the mean is a number which describes what?

1. **Heights of adults vs. kindergartners.**
Heights of 507 physically active individuals have a mean of 171 centimeters and a standard deviation of 9.4 centimeters.^[The [`bdims`](http://openintrostat.github.io/openintro/reference/bdims.html) data used in this exercise can be found in the [**openintro**](http://openintrostat.github.io/openintro) R package.] [@Heinz:2003]

    a. Would the standard deviation of the heights of a few hundred kindergartners be bigger or smaller than 9.4cm? Explain your reasoning.
    
    b. Suppose many samples of size 100 adults are taken and, separately, many samples of size 100 kindergarteners are taken. For each of the many samples, the average height is computed. Which set of sample averages would have a larger standard error of the mean, the adult sample averages or the kindergartner sample averages?

1. **Heights of adults, bootstrap interval.**
Researchers studying anthropometry collected body measurements, as well as age, weight, height and gender, for 507 physically active individuals. 
The histogram below shows the sample distribution of bootstrapped means from 1,000 different bootstrap samples.^[The [`bdims`](http://openintrostat.github.io/openintro/reference/bdims.html) data used in this exercise can be found in the [**openintro**](http://openintrostat.github.io/openintro) R package.] [@Heinz:2003]

    ```{r}
    library(tidyverse)
    library(infer)
    library(openintro)
    library(scales)

    set.seed(47)
    bdims %>%
      specify(response = hgt) %>%
      generate(reps = 1000, type = "bootstrap") %>%
      calculate(stat = "mean") %>%
      ggplot(aes(x = stat)) +
      geom_histogram(binwidth = 0.2, fill = IMSCOL["green", "full"]) +
      labs(
        x = "Bootstrapped mean of height (centimeters)",
        y = "Count",
        title = "1,000 bootstrap means"
      ) +
      scale_x_continuous(breaks = seq(170, 172.4, 0.4),
                         labels = label_comma(scale = 1, accuracy = 0.1))
    ```

    a.  Given the bootstrap sampling distribution for the sample mean, find an approximate value for the standard error of the mean. 
    
    b.  By looking at the bootstrap sampling distribution (1,000 bootstrap samples were taken), find an approximate 90% bootstrap percentile confidence interval for the true average adult height in the population from which the data were randomly sampled. Provide the interval as well as a one-sentence interpretation of the interval.

    c.  By looking at the bootstrap sampling distribution (1,000 bootstrap samples were taken), find an approximate 90% bootstrap SE confidence interval for the true average adult height in the population from which the data were randomly sampled. Provide the interval as well as a one-sentence interpretation of the interval.

1. **Identify the critical $t$.** 
A random sample is selected from an approximately normal population with unknown standard deviation.
Find the degrees of freedom and the critical $t$-value (t$^\star$) for the given sample size and confidence level.

    a.  $n = 6$, CL = 90%

    b.  $n = 21$, CL = 98%

    c.  $n = 29$, CL = 95%

    d.  $n = 12$, CL = 99%

1.  **$t$-distribution.** 
The figure below shows three unimodal and symmetric curves: the standard normal (z) distribution, the $t$-distribution with 5 degrees of freedom, and the $t$-distribution with 1 degree of freedom. 
Determine which is which, and explain your reasoning.

    ```{r fig.asp = 0.5}
    library(tidyverse)
    library(openintro)
    
    x <- seq(-5, 5, 0.01)
    z <- dnorm(x)
    t1 <- dt(x, 5)
    t5 <- dt(x, 1)
    
    df <- tibble(
        xx = c(x, x, x),
        yy = c(z, t1, t5),
        curve = factor(c(rep("solid", length(x)), rep("dashed", length(x)), rep("dotted", length(x))), 
                       levels = c("solid", "dashed", "dotted")),
    )
    
    ggplot(df, aes(x = xx, y = yy, color = curve, linetype = curve)) +
      geom_line(size = 0.8) +
      scale_color_openintro("three") +
      labs(x = NULL, y = NULL, color = NULL, linetype = NULL) +
      scale_linetype_manual(values = c("solid", "dashed", "dotted")) +
      scale_x_continuous(breaks = c(-4, -2, 0, 2, 4)) +
      theme(
        axis.text.y = element_blank(),
        legend.position = c(0.8, 0.8),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank()
      )
    ```

1. **Find the p-value, I.** 
A random sample is selected from an approximately normal population with an unknown standard deviation. 
Find the p-value for the given sample size and test statistic. 
Also determine if the null hypothesis would be rejected at $\alpha = 0.05$.

    a.  $n = 11$, $T = 1.91$

    b.  $n = 17$, $T = -3.45$

    c.  $n = 7$, $T = 0.83$

    d.  $n = 28$, $T = 2.13$

1. **Find the p-value, II.** 
A random sample is selected from an approximately normal population with an unknown standard deviation. 
Find the p-value for the given sample size and test statistic. 
Also determine if the null hypothesis would be rejected at $\alpha = 0.01$.

    a.  $n = 26$, $T = 2.485$

    b.  $n = 18$, $T = 0.5$
    
    \clearpage

1. **Length of gestation, confidence interval.**
Every year, the United States Department of Health and Human Services releases to the public a large dataset containing information on births recorded in the country. 
This dataset has been of interest to medical researchers who are studying the relation between habits and practices of expectant mothers and the birth of their children. 
In this exercise we work with a random sample of 1,000 cases from the dataset released in 2014.
The length of pregnancy, measured in weeks, is commonly referred to as gestation.
The histograms below show the distribution of lengths of gestation from the random sample of 1,000 births (on the left) and the distribution of bootstrapped means of gestation from 1,500 different bootstrap samples (on the right).^[The [`births14`](http://openintrostat.github.io/openintro/reference/births14.html) data used in this exercise can be found in the [**openintro**](http://openintrostat.github.io/openintro) R package.] 

    ```{r out.width = "100%", fig.width = 8, fig.asp = 0.5}
    library(openintro)
    library(tidyverse)
    library(infer)
    library(patchwork)
    
    p_sample <- ggplot(births14, aes(x = weeks)) +
      geom_histogram(binwidth = 1) +
      labs(
        x = "Gestation (weeks)",
        y = "Count",
        title = "Random sample of 1,000 births"
      )

    set.seed(47)
    births14_boot <- births14 %>%
      specify(response = weeks) %>%
      generate(reps = 1500, type = "bootstrap") %>%
      calculate(stat = "mean")
    
    p_boot <- ggplot(births14_boot, aes(x = stat)) +
      geom_histogram(binwidth = 0.05, fill = IMSCOL["green", "full"]) +
      scale_x_continuous(breaks = c(38.4, 38.5, 38.6, 38.7, 38.8, 38.9)) +
      labs(
        x = "Bootstrapped mean of gestation (weeks)",
        y = "Count",
        title = "1,500 bootstrap means"
        )

    p_sample + p_boot
    ```

    a. Given the bootstrap sampling distribution for the sample mean, find an approximate value for the standard error of the mean. 

    b. By looking at the bootstrap sampling distribution (1,500 bootstrap samples were taken), find an approximate 99% bootstrap percentile confidence interval for the true average gestation length in the population from which the data were randomly sampled. Provide the interval as well as a one-sentence interpretation of the interval.

    c. By looking at the bootstrap sampling distribution (1,500 bootstrap samples were taken), find an approximate 99% bootstrap SE confidence interval for the true average gestation length in the population from which the data were randomly sampled. Provide the interval as well as a one-sentence interpretation of the interval.
    
    \clearpage

1. **Length of gestation, hypothesis test.**
In this exercise we work with a random sample of 1,000 cases from the dataset released by the United States Department of Health and Human Services in 2014.
Provided below are sample statistics for gestation (length of pregnancy, measured in weeks) of births in this sample.^[The [`births14`](http://openintrostat.github.io/openintro/reference/births14.html) data used in this exercise can be found in the [**openintro**](http://openintrostat.github.io/openintro) R package.] 

    ```{r}
    library(openintro)
    library(tidyverse)
    library(kableExtra)

    births14 %>%
      summarise(
        Min    = round(min(weeks), 1),
        Q1     = round(quantile(weeks, 0.25), 1),
        Median = round(median(weeks), 1),
        Mean   = round(mean(weeks), 1),
        Q3     = round(quantile(weeks, 0.75), 1),
        Max    = round(max(weeks), 1),
        SD     = round(sd(weeks), 1),
        IQR    = round(IQR(weeks), 1)
      ) %>%
      kbl(linesep = "", booktabs = TRUE, align = "cccccccc") %>%
      kable_styling(bootstrap_options = c("striped", "condensed"), 
                    latex_options = "HOLD_position",
                    full_width = FALSE)
    ```

    a. What is the point estimate for the average length of pregnancy for all women? What about the median?

    b. You might have heard that human gestation is typically 40 weeks. Using the data, perform a complete hypothesis test, using mathematical models, to assess the 40 week claim. State the null and alternative hypotheses, find the T score, find the p-value, and provide a conclusion in context of the data.
    
    c. A quick internet search validates the claim of "40 weeks gestation" for humans. A friend of yours claims that there are different ways to measure gestation (starting at first day of last period, ovulation, or conception) which will result in estimates that are a week or two different. Another friend mentions that recent increases in cesarean births is likely to have decreased length of gestation. Do the data provide a mechanism to distinguish between your two friends' claims?

1. **Interpreting confidence intervals for population mean.**
For each of the following statements, indicate if they are a true or false interpretation of the confidence interval.
If false, provide a reason or correction to the misinterpretation.
You collect a large sample and calculate a 95% confidence interval for the average number of cans of sodas consumed annually per adult in the US to be (440 cans, 520 cans), i.e., on average, adults in the US consume just under two cans of soda per day.

    a. 95% of adults in the US consume between 440 and 520 cans of soda per year.

    b. There is a 95% probability that the true population average per adult yearly soda consumption is between 440 and 520 cans.

    c. The true population average per adult yearly soda consumption is between 440 and 520 cans, with 95% confidence.

    d. The average soda consumption of the people who were sampled is between 440 and 520 cans of soda per year, with 95% confidence.

1. **Interpreting p-values for population mean.**
For each of the following statements, indicate if they are a true or false interpretation of the p-value.
If false, provide a reason or correction to the misinterpretation.
You are wondering if the average amount of cereal in a 10oz cereal box is greater than 10oz. You collect 50 boxes of cereal, weigh them carefully, find a T score, and a p-value of 0.23.

    a. The probability that the average weight of all cereal boxes is 10 oz is 0.23.

    b. The probability that the average weight of all cereal boxes is greater than 10 oz is 0.23.

    c. Because the p-value is 0.23, the average weight of all cereal boxes is 10 oz.

    d. Because the p-value is small, the population average must be just barely above 10 oz (small effect).

    e. If $H_0$ is true, the probability of observing another sample with an average as or more extreme as the data is 0.23.
    
    \clearpage

1.  **Working backwards, I.** 
A 95% confidence interval for a population mean, $\mu$, is given as (18.985, 21.015). 
The population distribution is approximately normal and the population standard deviation is unknown. 
This confidence interval is based on a simple random sample of 36 observations. 
Calculate the sample mean, the margin of error, and the sample standard deviation.
Assume that all conditions necessary for inference are satisfied. 
Use the $t$-distribution in any calculations.

1. **Working backwards, II.** 
A 90% confidence interval for a population mean is (65, 77). 
The population distribution is approximately normal and the population standard deviation is unknown. 
This confidence interval is based on a simple random sample of 25 observations. 
Calculate the sample mean, the margin of error, and the sample standard deviation.
Assume that all conditions necessary for inference are satisfied. 
Use the $t$-distribution in any calculations.

1. **Sleep habits of New Yorkers.** 
New York is known as "the city that never sleeps". 
A random sample of 25 New Yorkers were asked how much sleep they get per night. 
Statistical summaries of these data are shown below. 
The point estimate suggests New Yorkers sleep less than 8 hours a night on average. 
Evaluate the claim that New York is the city that never sleeps keeping in mind that, despite this claim, the true average number of hours New Yorkers sleep could be less than 8 hours or more than 8 hours.

    ```{r}
    library(tidyverse)
    library(kableExtra)
    
    tribble(
      ~n, ~Mean, ~SD,  ~Min, ~Max,
      25,  7.73, 0.77, 6.17, 9.78
    ) %>%
      kbl(linesep = "", booktabs = TRUE, align = c("ccccc")) %>%
      kable_styling(bootstrap_options = c("striped", "condensed"), 
                    latex_options = "HOLD_position",
                    full_width = FALSE) %>%
      column_spec(1:5, width = "5em")
    ```

    a.  Write the hypotheses in symbols and in words.

    b.  Check conditions, then calculate the test statistic, $T$, and the associated degrees of freedom.

    c.  Find and interpret the p-value in this context. Drawing a picture may be helpful.

    d.  What is the conclusion of the hypothesis test?

    e.  If you were to construct a 90% confidence interval that corresponded to this hypothesis test, would you expect 8 hours to be in the interval?

1.  **Find the mean.** 
You are given the hypotheses shown below.
We know that the sample standard deviation is 8 and the sample size is 20. 
For what sample mean would the p-value be equal to 0.05? 
Assume that all conditions necessary for inference are satisfied.

    $$H_0: \mu = 60 \quad \quad H_A: \mu \neq 60$$

1. **$t^\star$ for the correct confidence level.** 
As you've seen, the tails of a $t-$distribution are longer than the standard normal which results in $t^{\star}_{df}$ being larger than $z^{\star}$ for any given confidence level. When finding a CI for a population mean, explain how mistakenly using $z^{\star}$ (instead of the correct $t^{*}_{df}$) would affect the confidence level.

1. **Possible bootstrap samples.**
Consider a simple random sample of the following observations: 47, 4, 92, 47, 12, 8.
Which of the following could be a possible bootstrap samples from the observed data above?
If the set of values could not be a bootstrap sample, indicate why not.

     a. 47, 47, 47, 47, 47, 47
     
     b. 92, 4, 13, 8, 47, 4
     
     c. 92, 47, 12
     
     d. 8, 47, 12, 12, 8, 4, 92
     
     e. 12, 4, 8, 8, 92, 12
     
     \clearpage

1. **Play the piano.** 
Georgianna claims that in a small city renowned for its music school, the average child takes less than 5 years of piano lessons. 
We have a random sample of 20 children from the city, with a mean of 4.6 years of piano lessons and a standard deviation of 2.2 years.

    a. Evaluate Georgianna's claim (or that the opposite might be true) using a hypothesis test.

    b. Construct a 95% confidence interval for the number of years students in this city take piano lessons, and interpret it in context of the data.

    c. Do your results from the hypothesis test and the confidence interval agree? Explain your reasoning.

1. **Auto exhaust and lead exposure.** 
Researchers interested in lead exposure due to car exhaust sampled the blood of 52 police officers subjected to constant inhalation of automobile exhaust fumes while working traffic enforcement in a primarily urban environment. 
The blood samples of these officers had an average lead concentration of 124.32 $\mu$g/l and a SD of 37.74 $\mu$g/l; a previous study of individuals from a nearby suburb, with no history of exposure, found an average blood level concentration of 35 $\mu$g/l. [@Mortada:2000]

    a. Write down the hypotheses that would be appropriate for testing if the police officers appear to have been exposed to a different concentration of lead.

    b. Explicitly state and check all conditions necessary for inference on these data.

    c. Test the hypothesis that the downtown police officers have a higher lead exposure than the group in the previous study. Interpret your results in context.
