1.  **Identify the critical $t$.** 
A random sample is selected from an approximately normal population with unknown standard deviation.
Find the degrees of freedom and the critical $t$-value (t$^\star$) for the given sample size and confidence level.

    a.  $n = 6$, CL = 90%

    b.  $n = 21$, CL = 98%

    c.  $n = 29$, CL = 95%

    d.  $n = 12$, CL = 99%

1.  **$t$-distribution.** 
The figure below shows three unimodal and symmetric curves: the standard normal (z) distribution, the $t$-distribution with 5 degrees of freedom, and the $t$-distribution with 1 degree of freedom. 
Determine which is which, and explain your reasoning.

    ```{r}
    library(tidyverse)
    library(openintro)
    
    x <- seq(-5, 5, 0.01)
    z <- dnorm(x)
    t1 <- dt(x, 5)
    t5 <- dt(x, 1)
    
    df <- tibble(
        xx = c(x, x, x),
        yy = c(z, t1, t5),
        curve = factor(c(rep("solid", length(x)), rep("dashed", length(x)), rep("dotted", length(x))), 
                       levels = c("solid", "dashed", "dotted")),
    )
    
    ggplot(df, aes(x = xx, y = yy, color = curve, linetype = curve)) +
      geom_line(size = 0.8) +
      scale_color_openintro("three") +
      labs(x = NULL, y = NULL, color = NULL, linetype = NULL) +
      scale_linetype_manual(values = c("solid", "dashed", "dotted")) +
      scale_x_continuous(breaks = c(-4, -2, 0, 2, 4)) +
      theme(
        axis.text.y = element_blank(),
        legend.position = c(0.8, 0.8),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank()
      )
    ```

1.  **Find the p-value, Part I.** 
A random sample is selected from an approximately normal population with an unknown standard deviation. 
Find the p-value for the given sample size and test statistic. 
Also determine if the null hypothesis would be rejected at $\alpha = 0.05$.

    a.  $n = 11$, $T = 1.91$

    b.  $n = 17$, $T = -3.45$

    c.  $n = 7$, $T = 0.83$

    d.  $n = 28$, $T = 2.13$

1.  **Find the p-value, Part II.** 
A random sample is selected from an approximately normal population with an unknown standard deviation. 
Find the p-value for the given sample size and test statistic. 
Also determine if the null hypothesis would be rejected at $\alpha = 0.01$.

    a.  $n = 26$, $T = 2.485$

    b.  $n = 18$, $T = 0.5$

1.  **Working backwards, Part I.** 
A 95% confidence interval for a population mean, $\mu$, is given as (18.985, 21.015). 
This confidence interval is based on a simple random sample of 36 observations. 
Calculate the sample mean and standard deviation. 
Assume that all conditions necessary for inference are satisfied. 
Use the $t$-distribution in any calculations.

1.  **Working backwards, Part II.** 
A 90% confidence interval for a population mean is (65, 77). 
The population distribution is approximately normal and the population standard deviation is unknown. 
This confidence interval is based on a simple random sample of 25 observations. 
Calculate the sample mean, the margin of error, and the sample standard deviation.

1.  **Sleep habits of New Yorkers.** 
New York is known as "the city that never sleeps". 
A random sample of 25 New Yorkers were asked how much sleep they get per night. 
Statistical summaries of these data are shown below. 
The point estimate suggests New Yorkers sleep less than 8 hours a night on average. 
Evaluate the claim that New York is the city that never sleeps keeping in mind that, despite this claim, the true average number of hours New Yorkers sleep could be less than 8 hours or more than 8 hours.

    ```{r}
    library(tidyverse)
    library(knitr)
    
    tribble(
      ~n, ~`$\\bar{x}$`, ~s,   ~min, ~max,
      25,  7.73,        0.77, 6.17, 9.78
    ) %>%
      kable(align = c("ccccc"))
    ```

    a.  Write the hypotheses in symbols and in words.

    b.  Check conditions, then calculate the test statistic, $T$, and the associated degrees of freedom.

    c.  Find and interpret the p-value in this context. Drawing a picture may be helpful.

    d.  What is the conclusion of the hypothesis test?

    e.  If you were to construct a 90% confidence interval that corresponded to this hypothesis test, would you expect 8 hours to be in the interval?

1.  **Heights of adults.** 
Researchers studying anthropometry collected body girth measurements and skeletal diameter measurements, as well as age, weight, height and gender, for 507 physically active individuals. 
The histogram below shows the sample distribution of heights in centimeters.^[The data used in this exercise can be found in the **openintro** package: [`bdims`](http://openintrostat.github.io/openintro/reference/bdims.html).] [@Heinz:2003]

    ```{r}
    library(tidyverse)
    library(knitr)
    
    bdims %>%
      summarise(
        Min    = round(min(hgt), 1),
        Q1     = round(quantile(hgt, 0.25), 1),
        Median = round(median(hgt), 1),
        Mean   = round(mean(hgt), 1),
        SD     = round(sd(hgt), 1),
        Q3     = round(quantile(hgt, 0.75), 1),
        Max    = round(max(hgt), 1)
      ) %>%
      kable()
    
    ggplot(bdims, aes(x = hgt)) +
      geom_histogram(binwidth = 5) +
      labs(x = "Height", y = "Count")
    ```

    a.  What is the point estimate for the average height of active individuals? What about the median?

    b.  What is the point estimate for the standard deviation of the heights of active individuals? What about the IQR?

    c.  Is a person who is 1m 80cm (180 cm) tall considered unusually tall? And is a person who is 1m 55cm (155cm) considered unusually short? Explain your reasoning.

    d.  The researchers take another random sample of physically active individuals. Would you expect the mean and the standard deviation of this new sample to be the ones given above? Explain your reasoning.

    e.  The sample means obtained are point estimates for the mean height of all active individuals, if the sample of individuals is equivalent to a simple random sample. What measure do we use to quantify the variability of such an estimate? Compute this quantity using the data from the original sample under the condition that the data are a simple random sample.

1.  **Find the mean.** 
You are given the hypotheses shown below.
We know that the sample standard deviation is 8 and the sample size is 20. 
For what sample mean would the p-value be equal to 0.05? 
Assume that all conditions necessary for inference are satisfied.
    
    $$
    \begin{align*}
    H_0&: \mu = 60 \\
    H_A&: \mu \neq 60
    \end{align*}
    $$

1.  **$t^\star$ vs. $z^\star$.** 
For a given confidence level, $t^{\star}_{df}$ is larger than $z^{\star}$. Explain how $t^{*}_{df}$ being slightly larger than $z^{*}$ affects the width of the confidence interval.

1.  **Play the piano.** 
Georgianna claims that in a small city renowned for its music school, the average child takes less than 5 years of piano lessons. 
We have a random sample of 20 children from the city, with a mean of 4.6 years of piano lessons and a standard deviation of 2.2 years.

    a.  Evaluate Georgianna's claim (or that the opposite might be true) using a hypothesis test.

    b.  Construct a 95% confidence interval for the number of years students in this city take piano lessons, and interpret it in context of the data.

    c.  Do your results from the hypothesis test and the confidence interval agree? Explain your reasoning.

1.  **Auto exhaust and lead exposure.** 
Researchers interested in lead exposure due to car exhaust sampled the blood of 52 police officers subjected to constant inhalation of automobile exhaust fumes while working traffic enforcement in a primarily urban environment. 
The blood samples of these officers had an average lead concentration of 124.32 $\mu$g/l and a SD of 37.74 $\mu$g/l; a previous study of individuals from a nearby suburb, with no history of exposure, found an average blood level concentration of 35 $\mu$g/l. [@Mortada:2000]

    a.  Write down the hypotheses that would be appropriate for testing if the police officers appear to have been exposed to a different concentration of lead.

    b.  Explicitly state and check all conditions necessary for inference on these data.

    c.  Test the hypothesis that the downtown police officers have a higher lead exposure than the group in the previous study. Interpret your results in context.
