1. **Air quality.** 
Air quality measurements were collected in a random sample of 25 country capitals in 2013, and then again in the same cities in 2014. 
We would like to use these data to compare average air quality between the two years. 
Should we use a paired or non-paired test? 
Explain your reasoning.

1. **True / False: paired.** 
Determine if the following statements are true or false. If false, explain.

    a.  In a paired analysis we first take the difference of each pair of observations, and then we do inference on these differences.

    b.  Two data sets of different sizes cannot be analyzed as paired data.

    c.  Consider two sets of data that are paired with each other. Each observation in one data set has a natural correspondence with exactly one observation from the other data set.

    d.  Consider two sets of data that are paired with each other. Each observation in one data set is subtracted from the average of the other data set's observations.

1. **Paired or not? Part I.** 
In each of the following scenarios, determine if the data are paired.

    a. Compare pre- (beginning of semester) and post-test (end of semester) scores of students.

    b.  Assess gender-related salary gap by comparing salaries of randomly sampled men and women.

    c.  Compare artery thicknesses at the beginning of a study and after 2 years of taking Vitamin E for the same group of patients.

    d.  Assess effectiveness of a diet regimen by comparing the before and after weights of subjects.

1. **Paired or not? Part II.** 
In each of the following scenarios, determine if the data are paired.

    a. We would like to know if Intel's stock and Southwest Airlines' stock have similar rates of return. To find out, we take a random sample of 50 days, and record Intel's and Southwest's stock on those same days.

    b.  We randomly sample 50 items from Target stores and note the price for each. Then we visit Walmart and collect the price for each of those same 50 items.

    c.  A school board would like to determine whether there is a difference in average SAT scores for students at one high school versus another high school in the district. To check, they take a simple random sample of 100 students from each high school.

1. **Global warming, Part I.** 
Let's consider a limited set of climate data, examining temperature differences in 1948 vs 2018. 
We sampled 197 locations from the National Oceanic and Atmospheric Administration's (NOAA) historical data, where the data was available for both years of interest. 
We want to know: were there more days with temperatures exceeding 90F in 2018 or in 1948? [@webpage:noaa_1948_2018] 
The difference in number of days exceeding 90F (number of days in 2018 - number of days in 1948) was calculated for each of the 197 locations. 
The average of these differences was 2.9 days with a standard deviation of 17.2 days. 
We are interested in determining whether these data provide strong evidence that there were more days in 2018 that exceeded 90F from NOAA's weather stations.

    ```{r}
    library(tidyverse)
    library(openintro)
    
    climate70 %>%
      mutate(diff = dx90_2018 - dx90_1948) %>%
      ggplot(aes(x = diff)) +
      geom_histogram(binwidth = 10) +
      labs(x = "Differences in numbers of days", y = "Count")
    ```

    a. Is there a relationship between the observations collected in 1948 and 2018? Or are the observations in the two groups independent? Explain.

    b.  Write hypotheses for this research in symbols and in words.

    c.  Check the conditions required to complete this test. A histogram of the differences is given to the right.

    d.  Calculate the test statistic and find the p-value.

    e.  Use $\alpha = 0.05$ to evaluate the test, and interpret your conclusion in context.

    f.  What type of error might we have made? Explain in context what the error means.

    g.  Based on the results of this hypothesis test, would you expect a confidence interval for the average difference between the number of days exceeding 90F from 1948 and 2018 to include 0? Explain your reasoning.

1. **High School and Beyond, Part I.** 
The National Center of Education Statistics conducted a survey of high school seniors, collecting test data on reading, writing, and several other subjects. 
Here we examine a simple random sample of 200 students from this survey. 
Side-by-side box plots of reading and writing scores as well as a histogram of the differences in scores are shown below.

    ```{r}
    library(tidyverse)
    library(openintro)
    library(patchwork)
    
    set.seed(1234)
    p_box <- hsb2 %>%
      pivot_longer(cols = c(read, write), names_to = "test", values_to = "score") %>%
      ggplot(aes(x = test, y = score, color = test)) +
      geom_boxplot(show.legend = FALSE) +
      geom_jitter(alpha = 0.6, width = 0.35, show.legend = FALSE) +
      scale_color_manual(values = c(IMSCOL["green", "full"], IMSCOL["red", "full"])) +
      labs(x = NULL, y = "Score")
    
    p_hist <- ggplot(hsb2, aes(x = read - write)) +
      geom_histogram(binwidth = 5) +
      labs(x = "Differences in scores (read - write)", y = "Count")
    
    p_box + p_hist
    ```

    a.  Is there a clear difference in the average reading and writing scores?

    b.  Are the reading and writing scores of each student independent of each other?

    c.  Create hypotheses appropriate for the following research question: is there an evident difference in the average scores of students in the reading and writing exam?

    d.  Check the conditions required to complete this test.

    e.  The average observed difference in scores is $\bar{x}_{read-write} = -0.545$, and the standard deviation of the differences is $s_{read-write} = 8.887$ points. Do these data provide convincing evidence of a difference between the average scores on the two exams?

    f.  What type of error might we have made? Explain what the error means in the context of the application.

    g.  Based on the results of this hypothesis test, would you expect a confidence interval for the average difference between the reading and writing scores to include 0? Explain your reasoning.

1. **Global warming, Part II.** 
We considered the change in the number of days exceeding 90F from 1948 and 2018 at 197 randomly sampled locations from the NOAA database. 
The mean and standard deviation of the reported differences are 2.9 days and 17.2 days.

    a.  Calculate a 90% confidence interval for the average difference between number of days exceeding 90F between 1948 and 2018. We've already checked the conditions for you.

    b.  Interpret the interval in context.

    c.  Does the confidence interval provide convincing evidence that there were more days exceeding 90F in 2018 than in 1948 at NOAA stations? Explain your reasoning.

1. **High school and beyond, Part II.** 
We considered the differences between the reading and writing scores of a random sample of 200 students who took the High School and Beyond Survey. 
The mean and standard deviation of the differences are $\bar{x}_{read-write} = -0.545$ and $s_{read-write}$ = 8.887 points.

    a.  Calculate a 95% confidence interval for the average difference between the reading and writing scores of all students.

    b.  Interpret this interval in context.

    c.  Does the confidence interval provide convincing evidence that there is a real difference in the average scores? Explain.
