1.  **Marvel Cinematic Universe films.** The data frame below contains information on Marvel Cinematic Universe films through the Infinity saga (a movie storyline spanning from Ironman in 2008 to Endgame in 2019).
    Box office totals are given in millions of US Dollars.
    How many observations and how many variables does this data frame have?[^_01-ex-data-hello-1]

    ```{r}
    mcu_films_scaled <- mcu_films |>
      mutate(
        opening_weekend_us = round(opening_weekend_us / 1000000, 2),
        gross_us = round(gross_us / 1000000, 2),
        gross_world = round(gross_world / 1000000, 2),
      ) |>
      rownames_to_column() |>
      rename(
        ` ` = rowname,
        Title = movie,
        Hrs = length_hrs,
        Mins = length_min,
        `Release Date` = release_date,
        `Opening Wknd US` = opening_weekend_us,
        US = gross_us,
        World = gross_world
      )

    mcu_films_1_to_5 <- mcu_films_scaled |>
      slice(1:5) |>
      mutate(across(everything(), as.character))

    mcu_films_n <- mcu_films_scaled |>
      slice(nrow(mcu_films)) |>
      mutate(across(everything(), as.character))

    mcu_films_filler <- mcu_films_scaled |>
      slice(1) |>
      mutate(across(everything(), str_replace, ".*", "..."))

    mcu_films_1_to_5 |>
      bind_rows(mcu_films_filler) |>
      bind_rows(mcu_films_n) |>
      kbl(linesep = "", booktabs = TRUE, align = "clcccccc") |>
      kable_styling(
        bootstrap_options = c("striped", "condensed"),
        latex_options = "HOLD_position"
      ) |>
      add_header_above(c(" " = 2, "Length" = 2, " " = 2, "Gross" = 2)) |>
      column_spec(2, width = "10em")
    ```

\vfill

2.  **Cherry Blossom Run.** The data frame below contains information on runners in the 2017 Cherry Blossom Run, which is an annual road race that takes place in Washington, DC. Most runners participate in a 10-mile run while a smaller fraction take part in a 5k run or walk.
    How many observations and how many variables does this data frame have?[^_01-ex-data-hello-2]

    ```{r}
    run17_to_print <- run17 |>
      rownames_to_column() |>
      rename_with(str_to_title) |>
      rename(
        ` ` = Rowname,
        Net = Net_sec,
        Clock = Clock_sec,
        Pace = Pace_sec,
        `City / Country` = City
      )

    run17_1_to_5 <- run17_to_print |>
      slice(1:5) |>
      mutate(across(everything(), as.character))

    run17_n <- run17_to_print |>
      slice(nrow(run17)) |>
      mutate(across(everything(), as.character))

    run17_filler <- run17_to_print |>
      slice(1) |>
      mutate(across(everything(), str_replace, ".*", "..."))

    run17_1_to_5 |>
      bind_rows(run17_filler) |>
      bind_rows(run17_n) |>
      kbl(linesep = "", booktabs = TRUE, align = "llllclcccl") |>
      kable_styling(
        bootstrap_options = c("striped", "condensed"),
        latex_options = "HOLD_position"
      ) |>
      add_header_above(c(" " = 6, "Time" = 2, " " = 2))
    ```

\vfill

3.  **Air pollution and birth outcomes, study components.** Researchers collected data to examine the relationship between air pollutants and preterm births in Southern California.
    During the study air pollution levels were measured by air quality monitoring stations.
    Specifically, levels of carbon monoxide were recorded in parts per million, nitrogen dioxide and ozone in parts per hundred million, and coarse particulate matter (PM$_{10}$) in $\mu g/m^3$.
    Length of gestation data were collected on 143,196 births between the years 1989 and 1993, and air pollution exposure during gestation was calculated for each birth.
    The analysis suggested that increased ambient PM$_{10}$ and, to a lesser degree, CO concentrations may be associated with the occurrence of preterm births.
    [@Ritz+Yu+Chapa+Fruin:2000]

    a.  Identify the main research question of the study.

    b.  Who are the subjects in this study, and how many are included?

    c.  What are the variables in the study?
        Identify each variable as numerical or categorical.
        If numerical, state whether the variable is discrete or continuous.
        If categorical, state whether the variable is ordinal.

\vfill

\clearpage

4.  **Cheaters, study components.** Researchers studying the relationship between honesty, age and self-control conducted an experiment on 160 children between the ages of 5 and 15.
    Participants reported their age, sex, and whether they were an only child or not.
    The researchers asked each child to toss a fair coin in private and to record the outcome (white or black) on a paper sheet, and said they would only reward children who report white.
    [@Bucciol:2011]

    a.  Identify the main research question of the study.

    b.  Who are the subjects in this study, and how many are included?

    c.  The study's findings can be summarized as follows: *"Half the students were explicitly told not to cheat and the others were not given any explicit instructions. In the no instruction group probability of cheating was found to be uniform across groups based on child's characteristics. In the group that was explicitly told to not cheat, girls were less likely to cheat, and while rate of cheating didn't vary by age for boys, it decreased with age for girls."* How many variables were recorded for each subject in the study in order to conclude these findings?
        State the variables and their types.

\vfill

5.  **Gamification and statistics, study components.** Gamification is the application of game-design elements and game principles in non-game contexts.
    In educational settings, gamification is often implemented as educational activities to solve problems by using characteristics of game elements.
    Researchers investigating the effects of gamification on learning statistics conducted a study where they split college students in a statistics class into four groups: (1) no reading exercises and no gamification, (2) reading exercises but no gamification, (3) gamification but no reading exercises, and (4) gamification and reading exercises.
    Students in all groups also attended lectures.
    Students in the class were from two majors: Electrical and Computer Engineering (n = 279) and Business Administration (n = 86).
    After their assigned learning experience, each student took a final evaluation comprised of 30 multiple choice question and their score was measured as the number of questions they answered correctly.
    The researchers considered students' gender, level of studies (first through fourth year) and academic major.
    Other variables considered were expertise in the English language and use of personal computers and games, both of which were measured on a scale of 1 (beginner) to 5 (proficient).
    The study found that gamification had a positive effect on student learning compared to traditional teaching methods involving lectures and reading exercises.
    They also found that the effect was larger for females and Engineering students.
    [@Legaki:2020]

    a.  Identify the main research question of the study.

    b.  Who were the subjects in this study, and how many were included?

    c.  What are the variables in the study?
        Identify each variable as numerical or categorical.
        If numerical, state whether the variable is discrete or continuous.
        If categorical, state whether the variable is ordinal.

\vfill

6.  **Stealers, study components.** In a study of the relationship between socio-economic class and unethical behavior, 129 University of California undergraduates at Berkeley were asked to identify themselves as having low or high social-class by comparing themselves to others with the most (least) money, most (least) education, and most (least) respected jobs.
    They were also presented with a jar of individually wrapped candies and informed that the candies were for children in a nearby laboratory, but that they could take some if they wanted.
    After completing some unrelated tasks, participants reported the number of candies they had taken.
    [@Piff:2012]

    a.  Identify the main research question of the study.

    b.  Who were the subjects in this study, and how many were included?

    c.  The study found that students who were identified as upper-class took more candy than others.
        How many variables were recorded for each subject in the study in order to conclude these findings?
        State the variables and their types.

\vfill

\clearpage

7.  **Migraine and acupuncture.** A migraine is a particularly painful type of headache, which patients sometimes wish to treat with acupuncture.
    To determine whether acupuncture relieves migraine pain, researchers conducted a randomized controlled study where 89 individuals who identified as female diagnosed with migraine headaches were randomly assigned to one of two groups: treatment or control.
    Forty-three (43) patients in the treatment group received acupuncture that is specifically designed to treat migraines.
    Forty-six (46) patients in the control group received placebo acupuncture (needle insertion at non-acupoint locations).
    Twenty-four (24) hours after patients received acupuncture, they were asked if they were pain free.
    Results are summarized in the contingency table below.
    Also provided is a figure from the original paper displaying the appropriate area (M) versus the inappropriate area (S) used in the treatment of migraine attacks.
    [^_01-ex-data-hello-3] [@Allais:2011]

    ```{r}
    migraine |>
      mutate(
        group = toTitleCase(as.character(group)),
        pain_free = toTitleCase(as.character(pain_free))
      ) |>
      count(group, pain_free) |>
      pivot_wider(names_from = pain_free, values_from = n) |>
      rename(Group = group) |>
      kbl(linesep = "", booktabs = TRUE) |>
      kable_styling(
        bootstrap_options = c("striped", "condensed"),
        latex_options = "HOLD_position",
        full_width = FALSE
      ) |>
      add_header_above(c("", "Pain free?" = 2)) |>
      column_spec(1:3, width = "5em")
    ```

    a.  What percent of patients in the treatment group were pain free 24 hours after receiving acupuncture?

    b.  What percent were pain free in the control group?

    c.  In which group did a higher percent of patients become pain free 24 hours after receiving acupuncture?

    d.  Your findings so far might suggest that acupuncture is an effective treatment for migraines for all people who suffer from migraines.
        However this is not the only possible conclusion.
        What is one other possible explanation for the observed difference between the percentages of patients that are pain free 24 hours after receiving acupuncture in the two groups?

    e.  What are the explanatory and response variables in this study?

\vfill

8.  **Sinusitis and antibiotics.** Researchers studying the effect of antibiotic treatment for acute sinusitis compared to symptomatic treatments randomly assigned 166 adults diagnosed with acute sinusitis to one of two groups: treatment or control.
    Study participants received either a 10-day course of amoxicillin (an antibiotic) or a placebo similar in appearance and taste.
    The placebo consisted of symptomatic treatments such as acetaminophen, nasal decongestants, etc.
    At the end of the 10-day period, patients were asked if they experienced improvement in symptoms.
    The distribution of responses is summarized below.[^_01-ex-data-hello-4]
    [@Garbutt:2012]

    ```{r}
    #| message: false
    sinusitis |>
      mutate(
        group = toTitleCase(as.character(group)),
        self_reported_improvement = toTitleCase(as.character(self_reported_improvement))
      ) |>
      count(group, self_reported_improvement) |>
      pivot_wider(names_from = self_reported_improvement, values_from = n) |>
      rename(Group = group) |>
      kbl(linesep = "", booktabs = TRUE) |>
      kable_styling(
        bootstrap_options = c("striped", "condensed"),
        latex_options = "HOLD_position",
        full_width = FALSE
      ) |>
      add_header_above(c("", "Improvement" = 2)) |>
      column_spec(1:3, width = "5em")
    ```

    a.  What percent of patients in the treatment group experienced improvement in symptoms?

    b.  What percent experienced improvement in symptoms in the control group?

    c.  In which group did a higher percentage of patients experience improvement in symptoms?

    d.  Your findings so far might suggest a real difference in the effectiveness of antibiotic and placebo treatments for improving symptoms of sinusitis.
        However this is not the only possible conclusion.
        What is one other possible explanation for the observed difference between the percentages patients who experienced improvement in symptoms?

    e.  What are the explanatory and response variables in this study?

\vfill

\clearpage

9.  **Daycare fines, study components.** Researchers tested the deterrence hypothesis which predicts that the introduction of a penalty will reduce the occurrence of the behavior subject to the fine, with the condition that the fine leaves everything else unchanged, by instituting a fine for late pickup at daycare centers.
    For this study, they worked with 10 volunteer daycare centers that did not originally impose a fine to parents for picking up their kids late.
    They randomly selected 6 of these daycare centers and instituted a monetary fine (of a considerable amount) for picking up children late and then removed it.
    In the remaining 4 daycare centers no fine was introduced.
    The study period was divided into four: before the fine (weeks 1–4), the first 4 weeks with the fine (weeks 5-8), the last 8 weeks with fine (weeks 9–16), and the after fine period (weeks 17-20).
    Throughout the study, the number of kids who were picked up late was recorded each week for each daycare.
    The study found that the number of late-coming parents increased discernibly when the fine was introduced, and no reduction occurred after the fine was removed.[^_01-ex-data-hello-5]
    [@Gneezy:2000]

    ```{r}
    daycare_fines_1_to_5 <- daycare_fines |>
      rename(study_period = study_period_4) |>
      select(center, week, group, late_pickups, study_period) |>
      slice(1:5) |>
      mutate(across(everything(), as.character))

    daycare_fines_200 <- daycare_fines |>
      rename(study_period = study_period_4) |>
      select(center, week, group, late_pickups, study_period) |>
      slice(200) |>
      mutate(across(everything(), as.character))

    daycare_fines_filler <- daycare_fines |>
      rename(study_period = study_period_4) |>
      select(center, week, group, late_pickups, study_period) |>
      slice(1) |>
      mutate(across(everything(), str_replace, ".*", "..."))

    daycare_fines_1_to_5 |>
      bind_rows(daycare_fines_filler) |>
      bind_rows(daycare_fines_200) |>
      kbl(linesep = "", booktabs = TRUE, align = "cclcl") |>
      kable_styling(
        bootstrap_options = c("striped", "condensed"),
        latex_options = "HOLD_position",
        full_width = FALSE
      )
    ```

    a.  Is this an observational study or an experiment?
        Explain your reasoning.

    b.  What are the cases in this study and how many are included?

    c.  What is the response variable in the study and what type of variable is it?

    d.  What are the explanatory variables in the study and what types of variables are they?

\vfill

10. **Efficacy of COVID-19 vaccine on adolescents, study components.** Results of a Phase 3 trial announced in March 2021 show that the Pfizer-BioNTech COVID-19 vaccine demonstrated 100% efficacy and robust antibody responses on 12 to 15 years old adolescents with or without prior evidence of SARS-CoV-2 infection.
    In this trial 2,260 adolescents were randomly assigned to two groups: one group got the vaccine (n = 1,131) and the other got a placebo (n = 1,129).
    While 18 cases of COVID-19 were observed in the placebo group, none were observed in the vaccine group.[^_01-ex-data-hello-6]
    [@Pfizer:2021]

    a.  Is this an observational study or an experiment?
        Explain your reasoning.

    b.  What are the cases in this study and how many are included?

    c.  What is the response variable in the study and what type of variable is it?

    d.  What are the explanatory variables in the study and what types of variables are they?

\vfill

11. **Palmer penguins.** Data were collected on 344 penguins living on three islands (Torgersen, Biscoe, and Dream) in the Palmer Archipelago, Antarctica.
    In addition to which island each penguin lives on, the data contains information on the species of the penguin (*Adelie*, *Chinstrap*, or *Gentoo*), its bill length, bill depth, and flipper length (measured in millimeters), its body mass (measured in grams), and the sex of the penguin (female or male).
    [^_01-ex-data-hello-7] [@palmerpenguins]

    a.  How many cases were included in the data?
    b.  How many numerical variables are included in the data? Indicate what they are, and if they are continuous or discrete.
    c.  How many categorical variables are included in the data, and what are they? List the corresponding levels (categories) for each.

\vfill

\clearpage

12. **Smoking habits of UK residents.** A survey was conducted to study the smoking habits of 1,691 UK residents.
    Below is a data frame displaying a portion of the data collected in this survey.
    A blank cell indicates that data for that variable was not available for a given respondent.[^_01-ex-data-hello-8]

    ```{r}
    set.seed(456)
    smoking_to_print <- smoking |>
      rename(sex = gender) |>
      select(sex, age, marital_status, gross_income, smoke, amt_weekends, amt_weekdays) |>
      slice_sample(n = nrow(smoking)) |>
      rownames_to_column() |>
      rename(
        weekend = amt_weekends,
        weekday = amt_weekdays,
        ` ` = rowname
      )

    smoking_1_to_5 <- smoking_to_print |>
      slice(1:5) |>
      mutate(across(everything(), as.character))

    smoking_n <- smoking_to_print |>
      slice(nrow(smoking)) |>
      mutate(across(everything(), as.character))

    smoking_filler <- smoking_to_print |>
      slice(1) |>
      mutate(across(everything(), str_replace, ".*", "..."))

    smoking_1_to_5 |>
      bind_rows(smoking_filler) |>
      bind_rows(smoking_n) |>
      kbl(linesep = "", booktabs = TRUE, align = "llllclcccl") |>
      kable_styling(
        bootstrap_options = c("striped", "condensed"),
        latex_options = "HOLD_position",
        full_width = FALSE
      ) |>
      add_header_above(c(" " = 6, "amount" = 2))
    ```

    a.  What does each row of the data frame represent?

    b.  How many participants were included in the survey?

    c.  Indicate whether each variable is numerical or categorical.
        If numerical, identify as continuous or discrete.
        If categorical, indicate if the variable is ordinal.

\vfill

13. **US Airports.** The visualization below shows the geographical distribution of airports in the contiguous United States and Washington, DC. This visualization was constructed based on a dataset where each observation is an airport.[^_01-ex-data-hello-9]

    ```{r}
    #| fig-asp: 0.5
    #| out-width: 100%

    # clean airport data -------------------------------------------------------
    usairports <- usairports |>
      filter(
        !str_detect(arp_latitude, "S"),
        !str_detect(state, "AK|HI|PR|MQ|GU|CQ|VI")
      ) |>
      mutate(
        lat_dms = str_replace(arp_latitude, "N", "") |>
          str_replace_all("-", " "),
        lon_dms = str_replace(arp_longitude, "W", "") |>
          str_replace_all("-", " "),
        lat_dd = conv_unit(lat_dms, from = "deg_min_sec", to = "dec_deg") |> as.numeric(),
        lon_dd = -1 * (conv_unit(lon_dms, from = "deg_min_sec", to = "dec_deg") |> as.numeric())
      ) |>
      filter(ownership %in% c("PR", "PU")) |> # only want public and private owned
      mutate(
        ownership = case_when(
          ownership == "PR" ~ "Privately owned",
          ownership == "PU" ~ "Publicly owned"
        ),
        use = case_when(
          use == "PR" ~ "Private use",
          use == "PU" ~ "Public use"
        ),
        region = case_when(
          region == "AAL" ~ "Alaska",
          region == "ACE" ~ "Central",
          region == "AEA" ~ "Eastern",
          region == "AGL" ~ "Great Lakes",
          region == "ANE" ~ "New England",
          region == "ANM" ~ "Northwest Mountain",
          region == "ASO" ~ "Southern",
          region == "ASW" ~ "Southwest",
          region == "AWP" ~ "West-Pacific"
        )
      )

    # plot ---------------------------------------------------------------------
    states_map <- map_data("state")

    ggplot(states_map, aes(long, lat, group = group)) +
      geom_polygon(fill = "white", color = "gray") +
      geom_point(
        data = usairports,
        aes(x = lon_dd, y = lat_dd, group = state, color = str_wrap(region, 10)),
        alpha = 0.3, size = 0.5
      ) +
      facet_grid(ownership ~ use) +
      labs(x = "", y = "", color = "Region") +
      scale_color_openintro() +
      scale_y_continuous(breaks = c(30, 40, 50)) +
      #theme(
      #  legend.position = "bottom",
      #  legend.direction = "horizontal",
      #) +
      guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
    ```

    a.  List the variables you believe were necessary to create this visualization.

    b.  Indicate whether each variable is numerical or categorical.
        If numerical, identify as continuous or discrete.
        If categorical, indicate if the variable is ordinal.

\vfill

\clearpage

14. **UN Votes.** The visualization below shows voting patterns in the United States, Canada, and Mexico in the United Nations General Assembly on a variety of issues.
    Specifically, for a given year between 1946 and 2019, it displays the percentage of roll calls in which the country voted yes for each issue.
    This visualization was constructed based on a dataset where each observation is a country/year pair.[^_01-ex-data-hello-10]

    ```{r}
    #| fig-asp: 0.5
    #| fig-width: 7
    #| out-width: 100%

    # plot unvotes by issues ---------------------------------------------------
    un_votes |>
      filter(country %in% c("United States", "Mexico", "Canada")) |>
      inner_join(un_roll_calls, by = "rcid") |>
      inner_join(un_roll_call_issues, by = "rcid") |>
      mutate(vote = fct_relevel(vote, "yes", "no", "abstain")) |>
      group_by(country, year = year(date), issue) |>
      summarize(
        votes = n(),
        percent_yes = mean(vote == "yes"), 
        .groups = "drop_last"
      ) |>
      filter(votes > 5) |>  # only use records with more than 5 votes
      ggplot(mapping = aes(x = year, y = percent_yes, color = str_wrap(country, 5))) +
        geom_point(alpha = 0.5) +
        geom_smooth(method = "loess", se = FALSE) +
        facet_wrap(~ issue, labeller = label_wrap_gen(width = 20)) +
        labs(
          y = "% Yes",
          x = "Year",
          color = "Country"
        ) +
        scale_color_openintro("three") +
        scale_x_continuous(breaks = c(1960, 1990, 2020)) +
        scale_y_continuous(breaks = c(0, 0.5, 1), labels = label_percent(scale = 1)) #+
      #theme(
      #  legend.position = "top",
      #  legend.direction = "horizontal"
      #)
    ```

    a.  List the variables used in creating this visualization.

    b.  Indicate whether each variable is numerical or categorical.
        If numerical, identify as continuous or discrete.
        If categorical, indicate if the variable is ordinal.

\vfill

15. **UK baby names.** The visualization below shows the number of baby girls born in the United Kingdom (comprised of England & Wales, Northern Ireland, and Scotland) who were given the name "Fiona" over the years.[^_01-ex-data-hello-11]

    ```{r}
    #| fig-asp: 0.5
    ukbabynames |>
      filter(name == "Fiona", sex == "F", year >= 2000) |>
      ggplot(aes(x = year, y = n, group = nation, color = nation, linetype = nation)) +
      geom_line() +
      geom_point() +
      scale_color_openintro("three") +
      labs(
        title = "Number of baby girls named Fiona",
        x = "Year",
        y = "Number of babies",
        color = "Nation", linetype = "Nation"
      ) +
      theme(
        legend.position = c(0.8, 0.7),
        legend.background = element_rect(color = "white")
      )
    ```

    a.  List the variables you believe were necessary to create this visualization.

    b.  Indicate whether each variable is numerical or categorical.
        If numerical, identify as continuous or discrete.
        If categorical, indicate if the variable is ordinal.

\vfill

\clearpage

16. **Shows on Netflix.** The visualization below shows the distribution of ratings of TV shows on Netflix (a streaming entertainment service) based on the decade they were released in and the country they were produced in.
    In the dataset, each observation is a TV show.[^_01-ex-data-hello-12]

    ```{r}
    #| fig-width: 7
    #| out-width: 90%
    # data from https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-04-20/netflix_titles.csv
    netflix_titles <- read_csv("data/netflix_titles.csv")

    netflix_to_plot <- netflix_titles |>
      filter(
        release_year >= 1980,
        type == "TV Show",
        !is.na(rating),
        !(rating %in% c("R", "NR")),
        !is.na(country)
      ) |>
      mutate(
        production_country = if_else(str_detect(country, "United States"), "United States", "Other"),
        production_country = fct_relevel(production_country, "United States", "Other"),
        release_decade = factor(floor(release_year / 10) * 10),
        rating = if_else(rating == "TV-Y7-FV", "TV-Y7", rating),
        rating = fct_relevel(rating, "TV-Y", "TV-Y7", "TV-G", "TV-PG", "TV-14", "TV-MA"),
        rating = fct_recode(rating,
          `All children (TV-Y)` = "TV-Y",
          `Older children (TV-Y7)` = "TV-Y7",
          `General audience (TV-G)` = "TV-G",
          `Parental guidance (TV-PG)` = "TV-PG",
          `Parents cautioned (TV-14)` = "TV-14",
          `Mature audience (TV-MA)` = "TV-MA"
        )
      ) |>
      rename(`Production country` = production_country)

    ggplot(netflix_to_plot, aes(x = release_decade, fill = rating)) +
      geom_bar(position = "fill", linewidth = 1) +
      scale_fill_manual(
        values = c(
          IMSCOL["green", "full"], IMSCOL["blue", "full"], IMSCOL["yellow", "f3"],
          IMSCOL["yellow", "full"], IMSCOL["pink", "full"], IMSCOL["red", "full"]
        )
      ) +
      labs(
        x = "Release decade", y = "Proportion", fill = "Rating"
      ) +
      facet_wrap(
        ~`Production country`,
        ncol = 1,
        labeller = labeller(`Production country` = label_both)
      ) +
      theme(legend.position = "bottom")
    ```

    a.  List the variables you believe were necessary to create this visualization.

    b.  Indicate whether each variable is numerical or categorical.
        If numerical, identify as continuous or discrete.
        If categorical, indicate if the variable is ordinal.

\vfill

17. **Stanford Open Policing.** The Stanford Open Policing project gathers, analyzes, and releases records from traffic stops by law enforcement agencies across the United States.
    Their goal is to help researchers, journalists, and policy makers investigate and improve interactions between police and the public.
    The following is an excerpt from a summary table created based off of the data collected as part of this project.
    [@pierson2020large]

    ```{r}
    # data from https://raw.githubusercontent.com/5harad/openpolicing/master/results/data_for_figures/combined_data.csv
    sop <- read_csv("data/stanford-open-policing-combined-data.csv")

    sop_to_print <- sop |>
      select(location, state, driver_race, stops_per_year, search_rate, arrest_rate) |>
      filter(
        location %in% c("APACHE COUNTY", "COCHISE COUNTY", "WOOD COUNTY"),
        state %in% c("AZ", "WI")
      ) |>
      mutate(location = str_to_title(location)) |>
      relocate(arrest_rate, .after = driver_race) |>
      rename_with(str_to_title) |>
      rename(
        `County` = Location,
        `Race / Ethnicity` = Driver_race,
        `Arrest rate` = Arrest_rate,
        `Stops / year` = Stops_per_year,
        `Search rate` = Search_rate,
      ) |>
      mutate(across(everything(), as.character))

    sop_top <- sop_to_print |> filter(State == "AZ") |> slice_head(n = 4)
    sop_filler <- sop_to_print |>
      slice(1) |>
      mutate(across(everything(), str_replace, ".*", "..."))
    sop_bottom <- sop_to_print |> filter(State == "WI") |> slice_tail(n = 2)

    sop_top |>
      bind_rows(sop_filler) |>
      bind_rows(sop_bottom) |>
      kbl(linesep = "", booktabs = TRUE, align = "lllccc") |>
      kable_styling(
        bootstrap_options = c("striped", "condensed"),
        latex_options = "HOLD_position",
        full_width = FALSE
      ) |>
      add_header_above(c(" " = 2, "Driver" = 2, "Car" = 2))
    ```

    a.  What variables were collected on each individual traffic stop in order to create the summary table above?

    b.  State whether each variable is numerical or categorical.
        If numerical, state whether it is continuous or discrete.
        If categorical, state whether it is ordinal or not.

    c.  Suppose we wanted to evaluate whether vehicle search rates are different for drivers of different races.
        In this analysis, which variable would be the response variable and which variable would be the explanatory variable?

\vfill

18. **Space launches.** The following summary table shows the number of space launches in the US by the type of launching agency and the outcome of the launch (success or failure).[^_01-ex-data-hello-13]

    ```{r}
    launches_1957_1999 <- tribble(
      ~agency,   ~Failure, ~Success,
      "Private", 13,       295,
      "State",   281,      3751,
      "Startup", 0,        0
    ) |>
      pivot_longer(cols = -agency, names_to = "outcome", values_to = "n") |>
      mutate(year = "1957-1999") |>
      uncount(n)

    launches_2000_2018 <- tribble(
      ~agency,   ~Failure, ~Success,
      "Private", 10,       562,
      "State",   33,       711,
      "Startup", 5,        65
    ) |>
      pivot_longer(cols = -agency, names_to = "outcome", values_to = "n") |>
      mutate(year = "2000-2018") |>
      uncount(n)

    launches_1957_1999 |>
      bind_rows(launches_2000_2018) |>
      count(year, agency, outcome) |>
      pivot_wider(names_from = outcome, values_from = n) |>
      pivot_wider(names_from = year, values_from = c("Failure", "Success"), values_fill = 0) |>
      select(agency, `Failure_1957-1999`, `Success_1957-1999`, `Failure_2000-2018`, `Success_2000-2018`) |>
      kbl(linesep = "", booktabs = TRUE, col.names = c("", "Failure", "Success", "Failure", "Success")) |>
      kable_styling(
        bootstrap_options = c("striped", "condensed"),
        latex_options = "HOLD_position",
        full_width = FALSE
      ) |>
      add_header_above(c(" " = 1, "1957 - 1999" = 2, "2000-2018" = 2))
    ```

    a.  What variables were collected on each launch in order to create to the summary table above?

    b.  State whether each variable is numerical or categorical.
        If numerical, state whether it is continuous or discrete.
        If categorical, state whether it is ordinal or not.

    c.  Suppose we wanted to study how the success rate of launches vary between launching agencies and over time.
        In this analysis, which variable would be the response variable and which variable would be the explanatory variable?

\vfill

19. **Pet names.** The city of Seattle, WA has an open data portal that includes pets registered in the city.
    For each registered pet, we have information on the pet's name and species.
    The following visualization plots the proportion of dogs with a given name versus the proportion of cats with the same name.
    The 20 most common cat and dog names are displayed.
    The diagonal line on the plot is the $x = y$ line; if a name appeared on this line, the name's popularity would be exactly the same for dogs and cats.[^_01-ex-data-hello-14]

    ```{r}
    #| fig-asp: 0.5
    # data prep ----------------------------------------------------------------
    name_props <- seattlepets |>
      filter(
        !is.na(animal_name),
        species %in% c("Dog", "Cat")
      ) |>
      group_by(species) |>
      count(animal_name, sort = TRUE) |>
      mutate(prop = n / sum(n))

    cat_name_props <- name_props |>
      filter(species == "Cat") |>
      rename(cat_prop = prop) |>
      slice(1:30)

    dog_name_props <- name_props |>
      filter(species == "Dog") |>
      rename(dog_prop = prop) |>
      slice(1:30)

    comb_name_props <- inner_join(cat_name_props, dog_name_props,
      by = "animal_name"
    ) |>
      ungroup() |>
      select(animal_name, cat_prop, dog_prop)

    # create viz ---------------------------------------------------------------
    ggplot(comb_name_props, aes(x = cat_prop, y = dog_prop)) +
      geom_abline(
        intercept = 0,
        alpha = 0.8,
        linewidth = 1.5, color = IMSCOL["gray", "full"]
      ) +
      geom_text_repel(aes(label = animal_name),
        seed = 291252, max.iter = 10000
      ) +
      geom_point(alpha = 0.8) +
      labs(x = "Proportion of cats", y = "Proportion of dogs") +
      xlim(0.002, 0.01) +
      ylim(0.002, 0.01)
    ```

    a.  Are these data collected as part of an experiment or an observational study?

    b.  What is the most common dog name?
        What is the most common cat name?

    c.  What names are more common for cats than dogs?

    d.  Is the relationship between the two variables positive or negative?
        What does this mean in context of the data?

\vfill

20. **Stressed out in an elevator.** In a study evaluating the relationship between stress and muscle cramps, half the subjects are randomly assigned to be exposed to increased stress by being placed into an elevator that falls rapidly and stops abruptly and the other half are left at no or baseline stress.

    a.  What type of study is this?

    b.  Can this study be used to conclude a causal relationship between increased stress and muscle cramps?

\vfill

[^_01-ex-data-hello-1]: The [`mcu_films`](http://openintrostat.github.io/openintro/reference/mcu_films.html) data used in this exercise can be found in the [**openintro**](http://openintrostat.github.io/openintro) R package.

[^_01-ex-data-hello-2]: The [`run17`](http://openintrostat.github.io/openintro/reference/run17.html) data used in this exercise can be found in the [**cherryblossom**](http://openintrostat.github.io/cherryblossom) R package.

[^_01-ex-data-hello-3]: The [`migraine`](http://openintrostat.github.io/openintro/reference/migraine.html) data used in this exercise can be found in the [**openintro**](http://openintrostat.github.io/openintro) R package.

[^_01-ex-data-hello-4]: The [`sinusitis`](http://openintrostat.github.io/openintro/reference/sinusitis.html) data used in this exercise can be found in the [**openintro**](http://openintrostat.github.io/openintro) R package.

[^_01-ex-data-hello-5]: The [`daycare_fines`](http://openintrostat.github.io/openintro/reference/daycare_fines.html) data used in this exercise can be found in the [**openintro**](http://openintrostat.github.io/openintro) R package.

[^_01-ex-data-hello-6]: The [`biontech_adolescents`](http://openintrostat.github.io/openintro/reference/biontech_adolescents.html) data used in this exercise can be found in the [**openintro**](http://openintrostat.github.io/openintro) R package.

[^_01-ex-data-hello-7]: Artwork by [Allison Horst](https://twitter.com/allison_horst).

[^_01-ex-data-hello-8]: The [`smoking`](http://openintrostat.github.io/openintro/reference/smoking.html) data used in this exercise can be found in the [**openintro**](http://openintrostat.github.io/openintro) R package.

[^_01-ex-data-hello-9]: The [`usairports`](http://openintrostat.github.io/airports/reference/usairports.html) data used in this exercise can be found in the [**airports**](http://openintrostat.github.io/airports/) R package.

[^_01-ex-data-hello-10]: The data used in this exercise can be found in the [**unvotes**](https://cran.r-project.org/web/packages/unvotes/index.html) R package.

[^_01-ex-data-hello-11]: The [`ukbabynames`](https://mine-cetinkaya-rundel.github.io/ukbabynames/reference/ukbabynames.html) data used in this exercise can be found in the [**ukbabynames**](https://mine-cetinkaya-rundel.github.io/ukbabynames/) R package.

[^_01-ex-data-hello-12]: The [`netflix_titles`](https://github.com/rfordatascience/tidytuesday/blob/master/data/2021/2021-04-20/readme.md) data used in this exercise can be found in the [**tidytuesdayR**](https://cran.r-project.org/web/packages/tidytuesdayR/index.html) R package.

[^_01-ex-data-hello-13]: The data used in this exercise comes from the [JSR Launch Vehicle Database, 2019 Feb 10 Edition](https://www.openintro.org/go?id=textbook-space-launches-data&referrer=ims0_html).

[^_01-ex-data-hello-14]: The [`seattlepets`](http://openintrostat.github.io/openintro/reference/seattlepets.html) data used in this exercise can be found in the [**openintro**](http://openintrostat.github.io/openintro) R package.
