\documentclass[11pt]{article}
\usepackage{geometry, graphicx, amssymb, amsmath, epstopdf, multicol, multirow}
\geometry{letterpaper}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
\newcounter{eoceSolCh}
\setcounter{eoceSolCh}{0}
\newcommand{\eoceSolCh}[1]{
\refstepcounter{eoceSolCh}\noindent\textbf{\arabic{eoceSolCh}\hspace{2mm}#1}

\addvspace{2mm}

}
\newcounter{eoceSol}[eoceSolCh]
\newcommand{\eoceSol}[1]{\refstepcounter{eoceSol}\noindent\small\textbf{\arabic{eoceSolCh}.\arabic{eoceSol}}\hspace{2mm}#1\addtocounter{eoceSol}{1}

\addvspace{1mm}

}
\begin{document}


%%%%%%%%%%%%%%%%%%%%%%%

\setcounter{eoceSolCh}{5}
\eoceSolCh{Small sample inference}

%\begin{multicols}{2}

\eoceSol{(a) $t^*_{41} = 1.68$ (b) $t^*_{20} = 1.33$ (c) $t^*_{28} = 2.05$ (d) $t^*_{11} = 3.11$}

\eoceSol{With a larger critical value the confidence interval ends up being wider.}

\eoceSol{(a) $H_0: \mu = 8$ (New Yorkers on average sleep 8 hrs per night.), $H_A: \mu < 8$ (New Yorkers on average sleep less than 8 hrs per night.) (b) Independence is satisfied since the sample is random ad less than 10\% of the population. The distribution doesn't appear to be strongly skewed. (c) T = -1.75, df = 24. (d) If in fact the true population mean of the amount New Yorkers sleep per night was 8 hours, the probability of getting a random sample of 25 New Yorkers where the average amount of sleep is 7.73 hrs per night or less is between 0.025 and 0.05. (f) Reject $H_0$, the data provide convincing evidence to suggest that New Yorkers on average sleep less than 8 hours per night. (g) No.}

\eoceSol{(a) We are 90\% confident that New Yorkers on average sleep 7.46 to 7.99 hours per night. (b) Yes.}

\eoceSol{No, a t-test is not appropriate since the distributions of total personal income are extremely skewed.}

\eoceSol{(a) $p-value < 0.005$, Reject $H_0$. (b) $0.01 < p-value < 0.02$, Reject $H_0$. (c) $0.05 < p-value < 0.10$, Fail to reject $H_0$. (d) $p-value < 0.20$, Fail to reject $H_0$.}

\eoceSol{(a) (-0.91, 4.91) (b) We are 95\% confident that those in the group that got the weight loss pill lost 0.91\textit{lbs} less to 4.91\textit{lbs} more than those in the placebo group. (c) No. (d) No.}

\eoceSol{(a) Chicken that were fed linseed on average weigh 218.75 grams while those that were given horsebean weigh on average 160.20 grams. Both distributions are relatively symmetric with no apparent outliers. Weights of chicken that were fed horsebean range from about 100 grams to 225 grams, and weights of chicken that were fed horsebean range from about 150 grams to 300 grams. There is a lot more variability in the weights of chicken that were given linseed. (b) $H_0: \mu_L = \mu_H, H_A: \mu_L \ne \mu_H$. (c) Independence is satisfied since both samples are random and less than 10\% of their prospective populations. The distributions do not appear to be extremely skewed and the samples are independent of each other. (d) $T = 3.02, df = 10, 0.01 < p-value < 0.02$. (e) Reject $H_0$, the data provide convincing evidence to suggest that there is a significant difference between the average weights of chicken that were fed linseed and horsebean. (f) Type I. (g) Yes.}

\eoceSol{(a) $H_0: \mu_A = \mu_M, H_A: \mu_A \ne \mu_M$. $T = 5.46, df = 25,  p-value < 0.01$. Reject $H_0$, the data provide convincing evidence to suggest that there a significant difference between the average fuel efficiency of cars with manual and automatic transmission in terms of their city mileage.}

\eoceSol{We are 95\% confident that cars with manual transmission on average get 5.13 to 10.73 miles per gallon more than cars with automatic transmission.}

\eoceSol{(a) $H_0: p = 0.69, H_A: p < 0.69$. (b) $\hat{p} = 0.57$. (c) The success-failure condition is not satisfied. (d) Each student can be represented with a card. Take 100 cards, 69 black cards representing those who follow the news about Egypt and 31 red cards representing those who do not. Shuffle the cards and draw 30 cards representing the 30 high school students. Calculate the proportion of black cards in this sample, $\hat{p}_{sim}$, i.e. the proportion of those who follow the news. Repeat 10,000 times and plot the resulting sample proportions. The p-value will be the proportion of simulations where $\hat{p}_{sim} \le 0.57$. (e) $p-value = 0.1348$ (Note: Answers may vary.) Fail to reject $H_0$, the data do not provide convincing evidence that the proportion of high school students who followed the news about Egypt is lower than the proportion of American adults who did.}

\eoceSol{(a) $H_0: p_{P} = p_{C}, H_A: p_{P} \ne p_{C}$. (b) -0.35. (c) $p-value = 0.0286$ (Note: Answers may vary.) Reject $H_0$, data provide convincing evidence that people react differently under the two scenarios.}


%\end{multicols}

\end{document}  