\section{Simulating randomness}
\label{simulateRandomness}

\subsection{Why simulate?}

Simulations are useful in examining the results of a sample. In Section~1.6%\ref{chanceModel} ZZQ
, we simulated an experiment 100 times and used those results to conclude that the drug sulphinpyrazone was effective at reducing deaths in heart attack victims.

Collecting data to investigate the properties of randomness can be very expensive, and simulation provides a cheap alternative. Simulations become even cheaper and faster when statisticians write a program to substitute for simulations by hand. %Each of the examples and problems in this section can be coded by an experienced statistician in just a few minutes. Once simulation is coded into a computer, the statistician can run millions of these simulations in just a few seconds.
Figure~\ref{dieProp} on page~\pageref{dieProp} represents 100,000 die rolls. We tried to pay an undergraduate student to do it but she quit\footnote{Kidding. We wouldn't want anyone to have to go through that.}, so we turned to a computer to help us. That program was written in under 10 minutes and it took a desktop computer about 1 second to simulate all 100,000 die rolls (!).

In this section, we describe how to successfully design and run simulations. This foundational understanding is a prerequisite for using the power of a computer to run hundreds, thousands, or even millions of simulations.

\subsection{Simulation examples}

\begin{example}{Below is a series of random numbers between 0 and 9 generated from a computer:
\begin{center}
7 1 5 9 1 7 4 4 3 1 \\
1 3 5 3 9 8 2 0 7 6 \\
%\resp{7} \resp{1} \resp{5} \resp{9} \resp{1} \resp{7} \resp{4} \resp{4} \resp{3} \resp{1} \\
%\resp{1} \resp{3} \resp{5} \resp{3} \resp{9} \resp{8} \resp{2} \resp{0} \resp{7} \resp{6} \\
\end{center}
This is much like rolling a ten sided die where each outcome is equally probable and recording the outcomes. Is there a way we could use these 20 random numbers to represent 20 coin flips?}\label{coinSimExam}
Yes, we can use these random numbers to \emph{simulate} coin tosses. The outcomes \resp{H} and \resp{T} are equally likely, and a random number has an equal chance of being in $\{0, 1, 2, 3, 4\}$ %$\{$\resp{0}, \resp{1}, \resp{2}, \resp{3}, \resp{4}$\}$ 
and $\{5, 6, 7, 8, 9\}$.% $\{$\resp{5}, \resp{6}, \resp{7}, \resp{8}, \resp{9}$\}$.
 We represent a result of \resp{H} by the random numbers {0}-{4} and let the remaining numbers, {5}-{9}, correspond to tossing a \resp{T}. Under this setup, independent random numbers provides a way to represent independent coin tosses. Furthermore, we confirm the probability of \resp{H} is 0.5:
\begin{eqnarray*}
P(\text{\resp{T}}) = P(\text{0 or 1 or 2 or 3 or 4})  %P(\text{\resp{0} or \resp{1} or \resp{2} or \resp{3} or \resp{4}})
= \frac{1}{10} + \frac{1}{10} + \frac{1}{10} + \frac{1}{10} + \frac{1}{10} = 1/2
\end{eqnarray*}
%Let $X_1, ..., X_{20}$ represent 20 random numbers (not the ones above since those are already known), which are assumed independent. Then
%\begin{eqnarray*}
%P(X_1 = 0) = P(X_1 = 1) = \cdots = P(X_1 =8) = P(X_1=9) = 0.1
%\end{eqnarray*}
%And this is also true for $X_2$, ..., $X_{20}$. Then, how can a random number $X_1$ be transformed into a coin flip where $P($\var{H}$)=P($\var{T}$)=0.5$? (Here, \var{H} means ``heads'' and \var{T} means ``tails''.)
%One method is to split the numbers 0-9 up and assign some to mean \var{H} and the others to mean \var{T}. For instance, if $X_1$ is 0, 1, 2, 3, or 4, then call the coin flip \var{H}. If the random number is anything else -- 5, 6, 7, 8, or 9 -- then the simulated coin flip is \var{T}. Under this setup,
%\begin{eqnarray*}
%P(\text{\var{H}}) = P(0)+P(1)+P(2)+P(3)+P(4)=0.5
%\end{eqnarray*}
%and the same is true of \var{T}. 
Using our assignment from random numbers to outcomes of a coin flip, we can simulate 20 tosses, as shown in Table~\ref{coinSim}. The simulation results in 11 \resp{H}s and 9 \resp{T}s.
\end{example}
\begin{table}[hht]
\begin{center}
\begin{tabular}{l ccccc ccccc}
Random \# & 7 & 1 & 5 & 9 & 1 & 7 & 4 & 4 & 3 & 1 \\
\hline
Coin flip & \var{T} &\var{H} & \var{T} & \var{T} & \var{H} & \var{T} & \var{H} & \var{H} & \var{H} & \var{H}  \\
\hline
\hline
Random \# & 1 & 3 & 5 & 3 & 9 & 8 & 2 & 0 & 7 & 6 \\
\hline
Coin flip & \var{H} & \var{H} & \var{T} & H & \var{T} & \var{T} & \var{H} & \var{H} & \var{T} & \var{T} \\
\end{tabular}
\end{center}
\caption{Simulation of coin flips with 20 random numbers.}
\label{coinSim}
\end{table}

In Example~\exam{coinSimExam}, it would have been equally reasonable to run the simulation with 0-4 corresponding to Tails and 5-9 to Heads. The point of a simulation is to represent a random process using random numbers, and there are many ways to do this. \\

\begin{example}{We now use random numbers between {00} and {99} to simulate the revenue of 10 statistics students of the bookstore from Section~\ref{expectation}:
\begin{center}
16 59 65 81 58 33 29 94 62 85
%\resp{16} \resp{59} \resp{65} \resp{81} \resp{58} \resp{33} \resp{29} \resp{94} \resp{62} \resp{85}
\end{center}
What is the mean revenue based on these 10 simulated students.}\label{simBookstoreRev}
The bookstore classifies students into three types: those who buy no books (20\%), those who buy only the textbook for \$137 (55\%), and those who buy both the textbook and the study guide (25\%). Unlike Example~\exam{coinSimExam}, the group probabilities are multiples of $1/100$, not $1/10$. Thus, we cannot represent all these probabilities by grouping the numbers {0}-{9}. Instead we use the numbers {00}, {01}, ..., {99}.

We assign ranges of the numbers {00}-{99} to each outcome. Each random pair of digits has probability $1/100$. Because $P($\var{purchase} = \resp{none}$) = 20/100$, twenty random numbers must be assigned to \resp{none}: {00}-{19}. Likewise, the outcomes \resp{textbook} and \resp{both} are assigned appropriate portions of the random numbers: {20}-{74} and {75}-{99}, respectively.

The revenue from 10 student can be simulated using the random numbers above, shown in Table~\ref{bookSim}. The mean revenue per student is \$133.20. Earlier we computed the true mean for this probability distribution as \$117.85.
\end{example}
\begin{table}[hht]
\begin{center}
\begin{tabular}{l ccccc}
\hline
\hline
Random \# & 16 & 59 & 65 & 81 & 58 \\%\resp{16} & \resp{59} & \resp{65}       & \resp{81} & \resp{58}    \\
\hline
\var{purchase} & \resp{none} & \small\resp{textbook} & \small\resp{textbook} & \resp{both} & \small\resp{textbook}   \\
\var{revenue} & \resp{0} & \resp{137} & \resp{137} & \resp{170} & \resp{137}  \\
\hline
\hline
Random \# &  33 & 29 & 94 & 62 & 85 \\ % \resp{33} &   \resp{29} & \resp{94} &    \resp{62} &    \resp{85} \\
\hline
\var{purchase} & \small\resp{textbook} & \small\resp{textbook} & \resp{both} & \small\resp{textbook} & \resp{both}  \\
\var{revenue} & \resp{137} & \resp{137} & \resp{170} & \resp{137} & \resp{170} \\
\hline
\hline
\end{tabular}
\end{center}
\caption{Simulation of 10 students and their corresponding revenues.}
\label{bookSim}
% R <- c(0, 137, 137, 170, 137, 137, 137, 170, 137, 170)
\end{table}

\begin{exercise}\label{simulatingDice}
Let $X$ represent a random number between {0} and {9}. Verify
\begin{eqnarray*}
P(X=1 \text{ } | \text{ } X\text{ is 1, 2, 3, 4, 5, or 6}) = 1/6 %\resp{1}, \resp{2}, \resp{3}, \resp{4}, \resp{5}, or \resp{6}}) = 1/6
\end{eqnarray*}
Answer in the footnote\footnote{This conditional probability is found by first finding the following two probabilities:
\begin{eqnarray*}
P(X=\text{ 1 and }X\text{ is in }\{1,2,3,4,5,6\})
%P(X=\text{ \resp{1} and }X\text{ is in }\{\text{\resp{1}, \resp{2}, \resp{3}, \resp{4}, \resp{5}, \resp{6}}\})
\quad \text{ and } \quad
P(X\text{ is in }\{1,2,3,4,5,6\})
%P(X\text{ is in }\{\text{\resp{1}, \resp{2}, \resp{3}, \resp{4}, \resp{5}, \resp{6}}\})
\end{eqnarray*}
The first probability is the same as $P(X=1)$, and this probability equals 1/10. The second probability is just 6/10. Therefore, the conditional probability is the ratio of these two probabilities: $\frac{1/10}{6/10} = 1/6$.}.
\end{exercise}

\begin{example}{Can you use the information in Exercise~\exer{simulatingDice} to simulate 10 die rolls using the random numbers below?
\begin{center}
8 5 4 7 5 4 3 5 0 9 \\
4 5 7 5 3 3 6 1 5 3
%\resp{8} \resp{5} \resp{4} \resp{7} \resp{5} \resp{4} \resp{3} \resp{5} \resp{0} \resp{9} \\
%\resp{4} \resp{5} \resp{7} \resp{5} \resp{3} \resp{3} \resp{6} \resp{1} \resp{5} \resp{3}
\end{center}
}\label{simulatingDiceExam}
%We can use Exercise~\exer{simulatingDice} to simulate die rolls based on the random numbers.
Instead of obtaining a die roll from every random number, we only consider those numbers {1}, {2}, ..., {6} and let each represent a die roll. If a random number is 0, 7, 8, or 9, we ignore it. Exercise~\exer{simulatingDice} guarantees the probability each outcome in a simulated roll is $1/6$. This technique is applied in Table~\ref{dieSim} to simulate the outcomes of ten die rolls.
\end{example}
\begin{table}[hht]
\begin{center}
\begin{tabular}{l ccccc ccccc}
Random \# &{8} &{5} &{4} &{7} &{5} &{4} &{3} &{5} &{9} &{0} \\
\hline
Outcome & - & \resp{5} & \resp{4} & - & \resp{5} & \resp{4} & \resp{3} & \resp{5} & - & -  \\
\hline
\hline
Random \# &{4} &{5} &{7} &{5} &{3} &{3} &{6} &{1} &{5} &{3} \\
\hline
Outcome & \resp{4} & \resp{5} & - & \resp{5} & stop & -- & -- & -- & --  & -- 
\end{tabular}
\end{center}
\caption{Simulation of 10 die rolls. We stop after we simulate the tenth roll.}
\label{dieSim}
\end{table}

\subsection{Simulations using random numbers}

In Examples~\exam{coinSimExam}-\exam{simulatingDiceExam}, random variables and processes were simulated using careful assignment of random numbers to outcomes\footnote{The random numbers are generated by a computer using a complex algorithm that produces seemingly random and independent numbers. Because these numbers are not truly random but are computer generated, they are often called \term{pseudo random numbers}.}. In a simulation setup, each of the following must be guaranteed:
\begin{itemize}
\item[(1)] The probability of each outcome in a simulation is equal to the actual probability of the outcome in reality.
\item[(2)] Each random number is only assigned to a single outcome.
\end{itemize}
It is also acceptable to consider only some subset of the random numbers. For instance, in Example~\exam{simulatingDiceExam}, we only used random numbers {1} through {6}. \\

\begin{exercise}
About one in five people smoke in the US. Using the random numbers below, simulate the smoking behavior of 20 people.
\begin{center}
4 5 7 0 6\hspace{3mm} 2 1 1 7 5 \\
9 7 7 6 8\hspace{3mm} 7 4 3 2 2 \\
2 4 4 1 7\hspace{3mm} 8 8 7 5 2 \\
\end{center}
Note: you will not need all thirty numbers. One assignment arrangement is provided in the footnote\footnote{Assign 0-1 to \resp{smoker} and 2-9 to \resp{nonSmoker}.}.
\end{exercise}

Example~\exam{simBookstoreRev} required two random numbers per simulation because one revenue probability was not a multiples of $1/10$ but instead of $1/100$. If the probabilities had been in multiples of $1/1000$, then three digits would have been necessary per simulated outcome. \\

\begin{exercise}\label{retailReturnAnIem}
A large retail company knows that 23.7\% of its customers return an item.

(a) Using some of the random numbers below, simulate 15 customers, classifying each as either \resp{returnItem} or \resp{not}. Hint in the footnote\footnote{Use 738 for the first customer, 675 for the second, 443 for the third, etc.}.
\begin{center}
73867\hspace{1mm} 54439\hspace{1mm} 92138 \\
01549\hspace{1mm} 38302\hspace{1mm} 08879 \\
80786\hspace{1mm} 81483\hspace{1mm} 75366 \\
\end{center}

(b) If it takes 10 minutes for an employee to accept a return and restock the item(s), how much time must employees devote to returns in this simulation?
\end{exercise}

For simulating the die rolls, the desired probabilities were $1/6 = 0.16\bar{6}$, which is not a multiple of $1/10$, $1/100$, $1/1000$, etc. Instead of using more digits, a better method was needed. Using only a subset of the random numbers ensured we could simulate the probabilities exactly, whicih you proved in Exercise~\exer{simulatingDice}. \\

\begin{exercise}
The probability a randomly selected person was born on a given day of the week (\resp{Sunday}, \resp{Monday}, etc.) is $1/7$.
\begin{center}
05685\hspace{1mm} 93746\hspace{1mm} 82365 \\
43132\hspace{1mm} 73550\hspace{1mm} 79515 \\
\end{center}
Using these random numbers, simulate the days of the week that 15 random babies were born.
\end{exercise}

\subsection{Estimating distribution properties}

We computed the expected revenue per student in Example~\exam{revFromStudent} on page~\pageref{revFromStudent}. We can also estimate this average revenue through simulation. If we take the average revenues from the simulated sample in Example~\exam{simBookstoreRev}, it will provide an estimate of the true expected value.
\begin{eqnarray*}
\frac{\$0 + \$137 + \$137 + \cdots + \$170}{10} = \$133.20
\end{eqnarray*}
The true value is \$117.85. If we simulate a larger set of revenues, shown in Table~\ref{largeRevSim}, we can usually obtain a better estimate of the mean: \$114.78.
\begin{table}
\begin{center}
\begin{tabular}{rrrrr rrrrr}
\$137 & \$137 & \$137 & \$0 & \$137 & \hspace{0.8cm} \$137 & \$0 & \$0 & \$0 & \$0  \\
\$170 & \$0 & \$137 & \$137 & \$137 &  \$137 & \$170 & \$137 & \$0 & \$137 \\
\$137 & \$137 & \$170 & \$170 & \$137 & \$0 & \$0 & \$170 & \$137 & \$137 \\
\$170 & \$137 & \$0 & \$170 & \$137 & \$137 & \$0 & \$170 & \$137 & \$137 \\
\$137 & \$137 & \$170 & \$170 & \$137 & \$170 & \$170 & \$137 & \$137 & \$137
\end{tabular}
\end{center}
\caption{Fifty simulated bookstore revenues.}
\label{largeRevSim}
% x <- sample(c(0,137,170),50,T,c(0.2,.55,.25)); for(i in 1:50) cat(x[i], '& \\$')
\end{table}

%We can also simulate this  provided  method to approximate the mean of the process, i.e. estimate the expectation of the distribution. Using a small simulation sample, the expectation was estimated by simply providing the average of the simulation sample. This simulation-to-estimate procedure is not limited to the expectation. 

A simulation can also be run to estimate the variance, standard deviation, or any other property of the distibution. The sample standard deviation from the results of Example~\exam{simBookstoreRev} and Table~\ref{largeRevSim} are \$49.32 and \$63.06, respectively. The true standard deviation can be computed as \$60.49. In general, the larger the sample (or simulation), the more accurate estimates tend to be. \\

\begin{exercise}
Using the simulation results of Exercise~\exer{retailReturnAnIem}(b), estimate how much time staff would expect to devote to accepting returns and restocking items per customer transaction. Your estimate should take the form
\begin{eqnarray*}
\frac{\text{total time devoted to returns in the simulation sample}}
	{\text{total number of simulated customers}}
\end{eqnarray*}
\end{exercise}

\vspace{5cm}

We will include a random number table for this chapter when we release an end-of-chapter problem set (scheduled for Jan 4th, 2010).

%\LARGE\color{red} MUST MAKE A NOTE SOMEWHERE OF A TABLE OF RANDOM NUMBERS.

